### 2023-11-29 paper/main.tex
switch to natbib because of acmart and commit a tex file

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  sigconf,
  screen,
  review]{acmart}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{\textbf{#1}}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\widowpenalty10000
\clubpenalty10000
\author{Bohdan Liesnikov}
\orcid{0009-0000-2216-8830}
\affiliation{
  \institution{Delft University of Technology}
  \city{Delft}
  \country{The Netherlands}}
\email{B.Liesnikov@tudelft.nl}
\author{Jesper Cockx}
\orcid{0000-0003-3862-4073}
\affiliation{
  \institution{Delft University of Technology}
  \city{Delft}
  \country{The Netherlands}}
\email{J.G.H.Cockx@tudelft.nl}
\begin{abstract}
Dependently typed languages such as Agda, Coq, Lean, and Idris are used for statically enforcing properties of programs and for enabling type-driven development.
While there has been a lot of work to find the right theoretical foundations for the core type theory underlying these languages, their implementations have been studied to a smaller extent.
In particular, theoretical works rarely consider the plethora of features that exist in big languages like Agda.
As a consequence the developers have to make a lot of adaptations in the implementations, which leads to codebases that are hard to maintain.
In this work we focus on a part of the type-checker called the elaborator that transforms user-friendly surface syntax to a small well-behaved core theory.
Our core idea is to use an open datatype for constraints and plugin system for solvers.
This allows for a more compact base elaborator implementation while enabling extensions to the type system.
\end{abstract}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={ExEl: Building an Elaborator Using Extensible Constraints},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{ExEl: Building an Elaborator Using Extensible Constraints}
\author{}
\date{\today}

\begin{document}
\maketitle

\hypertarget{sec:introduction}{%
\section{Introduction}\label{sec:introduction}}

Strongly-typed languages allow us to catch certain classes of bugs at
compile-time by checking the implementation against the type-signature.
When the types are provided by the user they can be viewed as a form of
specification, constraining the behaviour of the programs. This comes
with the benefit of more static guarantees but with an increased toll on
the user to supply more precise information about the program. Many
languages choose to infer types, but another option is to use the
information in the types to infer parts of the program. This idea was
aptly worded by Conor McBride as ``Write more types and fewer
programs.''
\citep[chap.~2.1]{ptoopTypeInferenceThought2022, mcbrideEpigramPracticalProgramming2005}
Some examples of this include overloaded functions in Java, implicits in
Scala, and type classes in Haskell.

In dependently typed languages like Agda
\citep{norellPracticalProgrammingLanguage2007, theagdateamAgdaUserManual2023a},
Coq \citep{thecoqdevelopmentteamCoqProofAssistant2022} or Idris
\citep{bradyIdrisGeneralpurposeDependently2013} the types can be much
more precise. This gives us an opportunity to infer even larger parts of
the program from the type in the process of elaboration. Examples
include implicit arguments in Agda, implicit coercions in Coq, and
tactic arguments in Idris. The inference or ``solving'' can be not only
automatic but also interactive or partially automatic. Holes in Agda and
proof obligations in Coq are examples of interactive, while Canonical
Structures \citep{mahboubiCanonicalStructuresWorking2013} in Coq and
program-synthesis for holes in Haskell
\citep{koppelSearchingEntangledProgram2022} are partially automatic. All
these mechanisms use different algorithms and have various degrees of
extensibility. These are typically not very well isolated from each
other and can therefore interact in unexpected ways, for example in case
between implicits and instance search
\citep{agdausersPerformanceRegressionIssue2018}. This puts a toll on the
language developer to modulate the interactions and for the user to
understand these features.

In all of these examples, solvers evolved organically over time together
with the language. Coq historically struggled with similar issues in the
elaborator: for example, Canonical Structures was not properly
documented for 15 years \citep{mahboubiCanonicalStructuresWorking2013}.
Agda experimented with features baked into the core of the type system
-- like sized types -- which brought their own solver infrastructure
\citep{abelExtensionMartinLofType2016}. Lean 4 is aiming to allow the
users to develop new surface-level features
\citep{leonardodemouraLeanMetaprogramming2021} using elaboration monads
\citep{mouraLeanTheoremProver2021}, somewhat akin to elaborator
reflection in Idris
\citep{christiansenElaboratorReflectionExtending2016}, but Lean 3 was
built in a more conventional way \citep{demouraLeanTheoremProver2015a}.

One common piece of infrastructure needed by these inference algorithms
are metavariables, also known as ``existential variables''
\citep[chap.~2.2.1]{thecoqdevelopmentteamCoqProofAssistant2022}, which
represent as-of-yet unknown parts of the program. Another one is the
need is to constrain two terms to be equal, known as unification.
Metavariables and unification are heavily used throughout the compiler,
for inference of implicit arguments and for general type-checking,
making the compiler sensitive towards changes in unification algorithms.
Because of the complexity unification breaking changes are often
discovered only when ran against a large existing project on CI, like
\texttt{cubical} or \texttt{stdlib} for Agda or \texttt{unimath} for
Coq.

In this paper we propose a new architecture for an extensible elaborator
for dependently typed languages. The idea is to provide an API for
developers to tap into the elaboration procedure with custom solvers
that can manipulate metavariables and constraints placed on them. This
design separates the `what' the solvers are doing from the `when',
making the interaction points between different parts of the
type-checker explicit. As a result, this allows the developer to reason
more easily about exceptions and asynchronicity in the type-checker and
add new features in a more contained fashion. Practically, this means
that each feature is contained within one module, as opposed to being
spread around the codebase.

Contributions:

\begin{itemize}
\tightlist
\item
  We propose a new design blueprint for implementing a language that is
  extensible with new constraints and new solvers. It supports type
  classes (Section \ref{sec:case-typeclasses}), implicit arguments
  (Section \ref{sec:case-implicits}), implicit coercions and tactic
  arguments (Section \ref{sec:coercion-tactics}).
\item
  We propose a view on metavariables as communication channels for the
  solvers, drawing parallels with asynchronous programming primitives
  (Section \ref{sec:solvers-implementation}).
\item
  We decompose the usual components of a type-checker, like the unifier
  in Agda, into a suite of solvers which can be extended and interleaved
  by user-provided plugins (Section
  \ref{sec:constraints_and_unification}).
\item
  Following the blueprint, we present a prototype implementation of a
  dependently typed language available at
  \href{https://github.com/liesnikov/extensible-elaborator}{github.com/liesnikov/extensible-elaborator}.
\end{itemize}

\hypertarget{sec:unification_constraint_based_elaboration_and_design_challanges}{%
\section{Unification, constraint-based elaboration and design
challenges}\label{sec:unification_constraint_based_elaboration_and_design_challanges}}

Constraints have been an integral part of compilers for strongly-typed
languages for a while \citep{oderskyTypeInferenceConstrained1999}. For
example, both Haskell \citep{vytiniotisOutsideInModularType2011} and
Agda \citep[ chap.~3]{norellPracticalProgrammingLanguage2007} use
constraints extensively. In the former case, they are even reflected and
can be manipulated by the user
\citep[chap.~6.10.3]{orchardHaskellTypeConstraints2010a, ghcdevelopmentteamGHCUserGuide2022}.
This has proved to be a good design decision for GHC, as is reflected,
for example in a talk by \citet{peytonjonesTypeInferenceConstraint2019},
as well as in a few published sources
\citep{vytiniotisOutsideInModularType2011, peytonjonesPracticalTypeInference2007}.

In the land of dependently typed languages constraints are much less
principled. Agda has a family of constraints\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Monad/Base.hs\#L1157-L1191}{./src/full/Agda/TypeChecking/Monad/Base.hs\#L1157-L1191}.
  Here and hencefoth we shorten the links in footnotes to paths in the
  repository, the source code can be found at
  \href{https://github.com/agda/agda/blob/v2.6.4/}{github.com/agda/agda/blob/v2.6.4/}.}
that grew organically, currently counting 19 constructors. Idris
technically has constraints\footnote{\href{https://github.com/idris-lang/Idris2/blob/e673d05a67b82591131e35ccd50fc234fb9aed85/src/Core/UnifyState.idr}{./src/Core/UnifyState.idr}
  at
  \href{https://github.com/idris-lang/Idris2/blob/e673d05a67b82591131e35ccd50fc234fb9aed85}{github.com/idris-lang/Idris2/blob/e673d0}},
with the only two constructors being equality constraints for two terms
and for two sequences of terms. The same\footnote{\href{https://github.com/leanprover/lean4/blob/0a031fc9bbb43c274bb400f121b13711e803f56c/src/Lean/Meta/Match/Basic.lean\#L161}{./src/Lean/Meta/Match/Basic.lean\#L161}
  at
  \href{https://github.com/leanprover/lean4/blob/0a031fc9bbb43c274bb400f121b13711e803f56c/}{github.com/leanprover/lean4/blob/0a031f}}
holds for Lean. These languages either use constraints in a restricted,
single use-case manner -- namely, for unification -- or in an ad-hoc
manner.

In this section we will demonstrate why a more methodical approach to
constraints will result in more robust elaborators across the board. We
go over three typical challenges that come up when building a compiler
for a dependently typed language and the way they are usually solved. We
cover unification of the base language and the complexity of managing
the state of the unifier in Section \ref{sec:conversion_checking}. Then
we take a look at different kinds of implicit arguments and their
implementation in Section \ref{sec:implicit-arguments}. We briefly touch
on the problem of extending the unifier in Section
\ref{sec:extending-unification}. Finally, we summarize the issues in
Section \ref{sec:summary-of-the-issues}.

\hypertarget{sec:conversion_checking}{%
\subsection{Unification in presence of
meta-variables}\label{sec:conversion_checking}}

As mentioned in the introduction, in the process of type-checking we use
unification to compare terms, which is notoriously hard to implement.
The complexity stems from the desire of compiler writers to implement
the most powerful unifier, while being limited by the fact that
higher-order unification is undecidable in general. Some of this
complexity is unavoidable, but we can manage it better by splitting up
the unifier into smaller modular parts. In practice, this means that one
does not have to fit together an always-growing unifier but can instead
write different cases separately.

An example from Agda's elaborator is type-driven unification, which is
spread between about a hundred functions and 2200 lines of
code\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs}{./src/full/Agda/TypeChecking/Conversion.hs}}.
Each of the functions implements part of the ``business logic'' of the
unifier. But all of them contain a lot of code that deals with
bookkeeping related to metavariables and constraints:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  They have to throw and catch exceptions, driving the control flow of
  the unification.
\item
  They have to compute blockers that determine when a postponed
  constraint is retried.
\item
  They have to deal with cases where either or both of the sides
  equation or its type are either metavariables or terms whose
  evaluation is blocked on some metavariables.
\end{enumerate}

Concretely, this code uses functions like \texttt{noConstraints} and
\texttt{dontAssignMetas} which rely on specific behaviour of the
constraint solver system. Other functions like \texttt{abortIfBlocked},
\texttt{reduce} and \texttt{catchConstraint}/\texttt{patternViolation}
force the programmer to make a choice between letting the constraint
system handle blockers or doing it manually. These things are known to
be brittle and pose an increased mental overhead when making changes.

The unifier is heavily used throughout the type-checker: either as
function calls to \texttt{leqType} when type-checking terms and
\texttt{compareType} when type-checking applications, or as raised
constraints \texttt{ValueCmp} and \texttt{SortCmp} from
\texttt{equalTerm} while checking applications or definitions,
\texttt{ValueCmpOnFace} from \texttt{equalTermOnFace} again while
checking applications. At the same it is unintuitive and full of
intricacies as indicated by multiple comments\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs\#L484-L485}{./src/full/Agda/TypeChecking/Conversion.hs\#\#L484-L485},
  \href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs\#L541-L549}{./src/full/Agda/TypeChecking/Conversion.hs\#L541-L549}}.

We would like the compiler-writer to separate the concerns of managing
constraints and blockers from the actual logic of the comparison
function. In fact, if we zoom in on the \texttt{compareAtom} function,
the core can be expressed in about 20 lines\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs\#L550-L599}{./src/full/Agda/TypeChecking/Conversion.hs\#L550-L599}}
of simplified code, stripping out size checks, cumulativity, polarity,
and forcing. This is precisely what we would like the developer to
write.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{case}\NormalTok{ (m, n) }\KeywordTok{of}
\NormalTok{  (}\DataTypeTok{Lit}\NormalTok{ l1, }\DataTypeTok{Lit}\NormalTok{ l2) }\OperatorTok{|}\NormalTok{ l1 }\OperatorTok{==}\NormalTok{ l2 }\OtherTok{{-}\textgreater{}} \FunctionTok{return}\NormalTok{ ()}
\NormalTok{  (}\DataTypeTok{Var}\NormalTok{ i es, }\DataTypeTok{Var}\NormalTok{ i\textquotesingle{} es\textquotesingle{}) }\OperatorTok{|}\NormalTok{ i }\OperatorTok{==}\NormalTok{ i\textquotesingle{} }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
\NormalTok{      a }\OtherTok{\textless{}{-}}\NormalTok{ typeOfBV i}
\NormalTok{      compareElims [] [] a (var i) es es\textquotesingle{}}
\NormalTok{  (}\DataTypeTok{Con}\NormalTok{ x ci xArgs, }\DataTypeTok{Con}\NormalTok{ y \_ yArgs) }\OperatorTok{|}\NormalTok{ x }\OperatorTok{==}\NormalTok{ y }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
\NormalTok{      t\textquotesingle{} }\OtherTok{\textless{}{-}}\NormalTok{ conType x t}
\NormalTok{      compareElims t\textquotesingle{} (}\DataTypeTok{Con}\NormalTok{ x ci []) xArgs yArgs}
  \OperatorTok{...}
\end{Highlighting}
\end{Shaded}

The functions described above are specific to Agda but in other major
languages we can find similar problems with unifiers being large modules
that are hard to understand. The sizes of modules with unifiers are as
follows: Idris (1.5kloc\footnote{\href{https://github.com/idris-lang/Idris2/blob/102d7ebc18a9e881021ed4b05186cccda5274cbe/src/Core/Unify.idr}{./src/Core/Unify.idr}}),
Lean (1.8kloc\footnote{\href{https://github.com/leanprover/lean4/blob/75252d2b85df8cb9231020a556a70f6d736e7ee5/src/Lean/Meta/ExprDefEq.lean}{./src/Lean/Meta/ExprDefEq.lean}}),
Coq (1.8kloc\footnote{\href{https://github.com/coq/coq/blob/b35c06c3ab3ed4911311b4a9428a749658d3eff1/pretyping/evarconv.mli}{./pretyping/evarconv.mli}
  at
  \href{github.com/coq/coq/blob/b35c06c3ab3ed4911311b4a9428a749658d3eff1/}{github.com/coq/coq/blob/b35c06}}).
For Haskell, which is not a dependently typed language yet, but does
have a constraints system
\citep{peytonjonesTypeInferenceConstraint2019}, this number is at
2kloc\footnote{\href{https://gitlab.haskell.org/ghc/ghc/-/blob/b81cd709df8054b8b98ac05d3b9affcee9a8b840/compiler/GHC/Core/Unify.hs}{./compiler/GHC/Core/Unify.hs}
  at
  \href{https://gitlab.haskell.org/ghc/ghc/-/blob/b81cd709df8054b8b98ac05d3b9affcee9a8b840}{gitlab.haskell.org/ghc/ghc/-/blob/b81cd709d}}.

\hypertarget{sec:implicit-arguments}{%
\subsection{Type-checking function application in presence of implicit
arguments}\label{sec:implicit-arguments}}

During the type-checking of function application there may be different
kinds of implicit arguments to infer, for example, instance arguments,
or tactic arguments. If we start from a simple case of type-checking an
application of a function symbol to regular arguments, every next
extension needs to be handled as a special case.

Take Agda -- when checking an application of implicit
arguments\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Implicit.hs\#L96-L128}{./src/full/Agda/TypeChecking/Implicit.hs\#L96-L128}}
we already have to carry the information on how the argument will be
resolved and then create a specific kind of metavariable\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Implicit.hs\#L130-L149}{./src/full/Agda/TypeChecking/Implicit.hs\#L130-L149}}
\citep[chap.~3]{norellPracticalProgrammingLanguage2007} for each of
those cases.

For handling \texttt{auto} variables, Idris 2
\citep[chap.~13.1]{theidristeamIdrisTutorial2021} has to essentially
inline the search procedure through a chain of elaboration function
calls (\texttt{checkApp} to \texttt{checkAppWith} to
\texttt{checkAppWith\textquotesingle{}}) to
\texttt{makeAutoImplicit}\footnote{\href{https://github.com/idris-lang/Idris2/blob/870bc824371d504a03af937f326216302210a875/src/TTImp/Elab/App.idr\#L224-L241}{./src/TTImp/Elab/App.idr\#L224-L241}}.
This can accommodate interfaces (or type classes), but one can imagine
that if a different kind of implicit was added, like tactic arguments or
Canonical Structures, we would have to inline the search again,
requiring a non-trivial modification to the elaboration mechanism.

While the codebases above show that it is certainly possible to extend
languages with new features if the language was not written with
extensibility in mind, this can lead to rather ad hoc solutions.

\hypertarget{sec:extending-unification}{%
\subsection{Extending unification}\label{sec:extending-unification}}

While writing a unifier is hard enough as it is, at times the developers
might want to give their users ability to extend the unification
procedure.

Canonical Structures
\citep{saibiOutilsGeneriquesModelisation1999, mahboubiCanonicalStructuresWorking2013}
was already mentioned as it is in the overlap between type classes and
unification hints. Adding it to a language that does not support them
requires an extension of the unification algorithm with a rule that says
that projection from a canonical structure is an injective function
\citep[eq. 1]{mahboubiCanonicalStructuresWorking2013}.

One could also provide means to do so manually in a more general case,
by allowing users to declare certain symbols as injective. This is one
of the features requested by the Agda users
\citep{agdausersInjectiveUnificationPragma2023}.

Another example of this can be adding rules of associativity and
commutativity to the unifier, as described in the thesis by
\citet{holtenDependentTypeCheckingModulo2023}. It required 2000 lines of
code\footnote{\href{https://github.com/LHolten/agda-commassoc/tree/defenitional-commutativity}{github.com/LHolten/agda-commassoc/tree/defenitional-commutativity}}.
We would like to make changes such as this more feasible.

\hypertarget{sec:summary-of-the-issues}{%
\subsection{Summary of the issues}\label{sec:summary-of-the-issues}}

The examples above show that when building a dependently typed language
the core might be perfectly elegant and simple, but the features that
appear on top of it complicate the design.

One can also observe that while the code in existing implementations
might rely on constraints, the design at large does not put them at the
centre of the picture and instead views them primarily as a gadget.
Agda's constraint solver\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Constraints.hs\#L247-L298}{src/full/Agda/TypeChecking/Constraints.hs\#L247-L298}}
relies on the type-checker to call it at the point where it is needed
and has to be carefully engineered to work with the rest of the
codebase.

\hypertarget{sec:what-is-our-design-bringing-into-the-picture}{%
\section{A blueprint for extensible
elaborators}\label{sec:what-is-our-design-bringing-into-the-picture}}

Our idea for a new design is to shift focus more towards the constraints
themselves:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We give a stable API for raising constraints that can be called by the
  type-checker, essentially creating an ``ask'' to be fulfilled by the
  solvers. This is not dissimilar to the idea of mapping object-language
  unification variables to host-language ones as done by
  \citet{guidiImplementingTypeTheory2017}, view of the ``asks'' as a
  general effect \citep[chap.~4.4]{bauerEqualityCheckingGeneral2020}, or
  communication between actors \citep{allaisTypOSOperatingSystem2022a}.
\item
  To make the language more modular, we make constraints an extensible
  data type in the style of \citet{swierstraDataTypesCarte2008} and give
  an API to define new solvers with the ability to specify what kinds of
  constraints they can solve.
\end{enumerate}

In the examples in this paper, we follow the bidirectional style of
type-checking, but in practice, the design decisions are agnostic of the
underlying system, as long as it adheres to the principle of stating the
requirements on terms in terms of raising a constraint and not by, say,
pattern-matching on a concrete term representation.

From a birds-eye view the architecture looks as depicted in Figure
\ref{architecture-figure}. The type-checking begins by initializing
state and doing the syntax traversal. The traversal raises the
constraints, and for the moment, the constraints are simply stored. As
soon as we finish the traversal of some block (one declaration in our
case), the solver dispatcher is called. It goes over the set of
constraints, and for each active constraint calls different plugins to
make an attempt at solving it. Each plugin, whether user-supplied
(\texttt{Plugin\ A}) or provided by us (\texttt{unification}) consist of
a handler and a solver. The handler determines if the plugin can
potentially solve a constraint, if so the dispatcher runs the
corresponding solver. All components have some read access to the state,
including handlers which might for example verify that there are no
extra constraints on the metavariable. For the write access: syntax
traversal writes new metavariables to the state and elaborated
definitions; the solver dispatcher writes updated meta-information;
solvers write solutions to the metavariables and can raise new
constraints.

\begin{figure*}
  \center

  \includegraphics[width=\textwidth]{architecture-diagram.png}

  \caption{Architecture diagram}
  \label{architecture-figure}
\end{figure*}

For the moment we need to recompile the project in order to include new
plugins, but this is not necessary and a system that dynamically loads
plugins is possible to implement in a way that is similar to GHC
Plugins\footnote{\href{https://mpickering.github.io/plugins.html}{mpickering.github.io/plugins.html}}
or Accelerate \footnote{\href{https://github.com/tmcdonell/accelerate-llvm/blob/master/accelerate-llvm-native/src/Data/Array/Accelerate/LLVM/Native/Link/Runtime.hs\#L40}{github.com/tmcdonell/accelerate-llvm/blob/master/accelerate-llvm-native/src/Data/Array/Accelerate/LLVM/Native/Link/Runtime.hs}}
\citep{mcdonellTypesafeRuntimeCode2015}.

\hypertarget{sec:bidirectional}{%
\section{Dependently typed calculus and bidirectional
typing}\label{sec:bidirectional}}

In this section, we describe the core of the type system we implement.
We take pi-forall \citep{weirichImplementingDependentTypes2022} as a
basis for the system and extend it with metavariables in core syntax and
implicit arguments in the surface syntax. However, for all other
purposes, we leave the core rules intact and therefore, the core
calculus too.

\hypertarget{basic-language-and-rules}{%
\subsection{Basic language and rules}\label{basic-language-and-rules}}

This is a dependently typed calculus that includes Pi, Sigma and indexed
inductive types.

Here is the internal syntax term data type, apart from service
constructor omissions, like a placeholder \texttt{TRUSTME}.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{Term} \OtherTok{=}
  \CommentTok{{-}{-} type of types Type}
    \DataTypeTok{Type}
  \CommentTok{{-}{-} variables x}
  \OperatorTok{|} \DataTypeTok{Var} \DataTypeTok{TName}
  \CommentTok{{-}{-} abstraction \textbackslash{}x. a}
  \OperatorTok{|} \DataTypeTok{Lam}\NormalTok{ (}\DataTypeTok{Bind} \DataTypeTok{TName} \DataTypeTok{Term}\NormalTok{)}
  \CommentTok{{-}{-} application a b}
  \OperatorTok{|} \DataTypeTok{App} \DataTypeTok{Term} \DataTypeTok{Arg}
  \CommentTok{{-}{-} function type (x : A) {-}\textgreater{} B}
  \OperatorTok{|} \DataTypeTok{Pi} \DataTypeTok{Epsilon} \DataTypeTok{Type}\NormalTok{ (}\DataTypeTok{Bind} \DataTypeTok{TName} \DataTypeTok{Type}\NormalTok{)}
  \CommentTok{{-}{-} Sigma{-}type \{ x : A | B \}}
  \OperatorTok{|} \DataTypeTok{Sigma} \DataTypeTok{Term}\NormalTok{ (}\DataTypeTok{Bind} \DataTypeTok{TName} \DataTypeTok{Term}\NormalTok{)}
  \OperatorTok{|} \DataTypeTok{Prod} \DataTypeTok{Term} \DataTypeTok{Term}
  \OperatorTok{|} \DataTypeTok{LetPair} \DataTypeTok{Term}\NormalTok{ (}\DataTypeTok{Bind}\NormalTok{ (}\DataTypeTok{TName}\NormalTok{, }\DataTypeTok{TName}\NormalTok{) }\DataTypeTok{Term}\NormalTok{)}
  \CommentTok{{-}{-} Equality type  a = b}
  \OperatorTok{|} \DataTypeTok{TyEq} \DataTypeTok{Term} \DataTypeTok{Term}
  \OperatorTok{|} \DataTypeTok{Refl}
  \OperatorTok{|} \DataTypeTok{Subst} \DataTypeTok{Term} \DataTypeTok{Term}
  \OperatorTok{|} \DataTypeTok{Contra} \DataTypeTok{Term}
  \CommentTok{{-}{-} inductive datatypes}
  \OperatorTok{|} \DataTypeTok{TCon} \DataTypeTok{TCName}\NormalTok{ [}\DataTypeTok{Arg}\NormalTok{] }\CommentTok{{-}{-} types (fully applied)}
  \OperatorTok{|} \DataTypeTok{DCon} \DataTypeTok{DCName}\NormalTok{ [}\DataTypeTok{Arg}\NormalTok{] }\CommentTok{{-}{-} terms (fully applied)}
  \OperatorTok{|} \DataTypeTok{Case} \DataTypeTok{Term}\NormalTok{ [}\DataTypeTok{Match}\NormalTok{]}
    \CommentTok{{-}{-} metavariables}
  \OperatorTok{|} \DataTypeTok{MetaVar} \DataTypeTok{MetaClosure}
\end{Highlighting}
\end{Shaded}

Equality is not defined as a regular inductive type, but is instead
built-in. The user has access to the type and term constructor, but not
the ability to pattern-matching on it. Instead the language provides a
\texttt{subst} primitive of type
\texttt{(A\ x)\ -\textgreater{}\ (x=y)\ -\textgreater{}\ A\ y} and
\texttt{contra} that takes an equality of two different inductive type
constructors and produces an element of any type.

On top of the above the language includes indexed inductive datatypes
and case-constructs for their elimination. Indexed inductive datatypes
are encoded as parameterised datatypes with an equality argument
constraining the index.

As for metavariables \texttt{MetaVar}: as mentioned in the introduction,
they are placeholders in the syntax tree (AST) that are produced in the
process of elaboration. Metavariables do not appear in the surface
syntax as they are not created by the user. In this paper we implement
metavariables in the contextual style, as described by
\citet{abelHigherOrderDynamicPattern2011}, therefore they have to carry
around a closure of type \texttt{MetaClosure}.

\hypertarget{syntax-traversal}{%
\subsection{Syntax traversal}\label{syntax-traversal}}

We implement the core of the elaborator as a bidirectional syntax
traversal, raising a constraint every time we need to assert something
about the type.

This includes the expected use of unification constraints, like in case
we enter checking mode while the term should be inferred:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checkType tm ty }\OtherTok{=} \KeywordTok{do}
\NormalTok{  (etm, ty\textquotesingle{}) }\OtherTok{\textless{}{-}}\NormalTok{ inferType tm}
\NormalTok{  constrainEquality ty\textquotesingle{} ty }\DataTypeTok{I.Type}
  \FunctionTok{return}\NormalTok{ etm}
\end{Highlighting}
\end{Shaded}

Any time we want to decompose the type provided in checking mode we pose
a constraint that guarantees the type being in a specific shape:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checkType (}\DataTypeTok{S.Lam}\NormalTok{ lam) ty }\OtherTok{=} \KeywordTok{do}
\NormalTok{  mtyA }\OtherTok{\textless{}{-}}\NormalTok{ createMetaTerm}
\NormalTok{  mtx }\OtherTok{\textless{}{-}}\NormalTok{ createUnknownVar}
\NormalTok{  mtyB }\OtherTok{\textless{}{-}}\NormalTok{ extendCtx (}\DataTypeTok{I.TypeSig}\NormalTok{ (}\DataTypeTok{I.Sig}\NormalTok{ mtx mtyA))}
\NormalTok{                    (createMetaTerm)}
  \KeywordTok{let}\NormalTok{ mbnd }\OtherTok{=}\NormalTok{ bind mtx mtyB}
  \KeywordTok{let}\NormalTok{ metaPi }\OtherTok{=} \DataTypeTok{I.Pi}\NormalTok{ mtyA mbnd}

\NormalTok{  constrainEquality ty metaPi }\DataTypeTok{I.Type}
  \CommentTok{{-}{-} rest of the traversal can now use mtyA and mbnd}
  \OperatorTok{...}
\end{Highlighting}
\end{Shaded}

At certain points we have to raise a constraint which has an associated
continuation. Like for checking the type of a data constructor -- the
part of the program that comes as an argument to
\texttt{constrainTConAndFreeze} will be suspended (or ``blocked'') until
the meta is solved with something of shape \texttt{TCon\ \_\ \_}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checkType t}\OperatorTok{@}\NormalTok{(}\DataTypeTok{S.DCon}\NormalTok{ c args) ty }\OtherTok{=} \KeywordTok{do}
\NormalTok{  elabpromise }\OtherTok{\textless{}{-}}\NormalTok{ createMetaTerm}
\NormalTok{  constrainTConAndFreeze ty }\OperatorTok{$} \KeywordTok{do}
\NormalTok{    mty }\OtherTok{\textless{}{-}}\NormalTok{ substMetas ty}
    \KeywordTok{case}\NormalTok{ mty }\KeywordTok{of}
\NormalTok{      (}\DataTypeTok{I.TCon}\NormalTok{ tname params) }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
        \OperatorTok{...}
\NormalTok{      \_ }\OtherTok{{-}\textgreater{}} \FunctionTok{error} \StringTok{"impossible"}
\end{Highlighting}
\end{Shaded}

We add one more case to the elaborator for implicit arguments.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checkType (}\DataTypeTok{Implicit}\NormalTok{) ty }\OtherTok{=} \KeywordTok{do}
\NormalTok{  m }\OtherTok{\textless{}{-}}\NormalTok{ createMetaTerm}
\NormalTok{  raiseConstraint }\OperatorTok{$} \DataTypeTok{FillInTheMeta}\NormalTok{ m ty}
  \FunctionTok{return}\NormalTok{ m}
\end{Highlighting}
\end{Shaded}

This is used for implicit arguments as will be described in more detail
in Section \ref{sec:case-implicits}. The syntax traversal does not need
to know anything at all about the implicits. The only thing we require
is that the elaboration of the argument is called with the type
information available. This corresponds to how in bidirectional typing
function application is done in the inference mode but the arguments are
processed in checking mode.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{inferType (}\DataTypeTok{App}\NormalTok{ t1 t2) }\OtherTok{=} \KeywordTok{do}
\NormalTok{  (et1, }\DataTypeTok{Pi}\NormalTok{ tyA tyB) }\OtherTok{\textless{}{-}}\NormalTok{ inferType t1}
\NormalTok{  et2 }\OtherTok{\textless{}{-}}\NormalTok{ checkType t2 tyA}
  \FunctionTok{return}\NormalTok{ (}\DataTypeTok{App}\NormalTok{ et1 et2, subst tyB et2)}
\end{Highlighting}
\end{Shaded}

\hypertarget{sec:constraints_and_unification}{%
\section{Constraints and
unification}\label{sec:constraints_and_unification}}

In Section \ref{sec:bidirectional} we described the syntax traversal
part of the elaborator, which generates the constraints. In this section
we will go over the constraint datatype needed for the base language,
how unification is implemented and how we can extend the unification
procedure.

\hypertarget{base-constraints}{%
\subsection{Base constraints}\label{base-constraints}}

The datatype of constraints is open which means the user can write a
plugin to extend it. However, we offer a few constraints out of the box
to be able to type-check the base language.

For the purposes of the base language it suffices to have the following.

\begin{itemize}
\item
  A constraint that enforces equality of two terms of a given type

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} two terms given should be equal}
\KeywordTok{data} \DataTypeTok{EqualityConstraint}\NormalTok{ e }\OtherTok{=}
     \DataTypeTok{EqualityConstraint} \DataTypeTok{Syntax.Term} \DataTypeTok{Syntax.Term}
                        \DataTypeTok{Syntax.Type}
                        \DataTypeTok{Sytax.MetaVarId}
\end{Highlighting}
\end{Shaded}
\item
  A constraint that ensures that a metavariable is resolved eventually:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} this terms has to be filled in}
\KeywordTok{data} \DataTypeTok{FillInTheTerm}\NormalTok{ e }\OtherTok{=}
     \DataTypeTok{FillInTheTerm} \DataTypeTok{Syntax.Term}
\NormalTok{                   (}\DataTypeTok{Maybe} \DataTypeTok{Syntax.Type}\NormalTok{)}
\end{Highlighting}
\end{Shaded}
\item
  Lastly, a constraint which ensures that a term is a type constructor:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} the term passed to the constraint}
\CommentTok{{-}{-} should be a type constructor}
\KeywordTok{data} \DataTypeTok{TypeConstructorConstraint}\NormalTok{ e }\OtherTok{=}
     \DataTypeTok{TypeConstructorConstraint} \DataTypeTok{Syntax.Type}
\end{Highlighting}
\end{Shaded}
\end{itemize}

The type-checker raises them supplying the information necessary, but
agnostic of how they will be solved.

\hypertarget{sec:solvers-interface}{%
\subsection{Interface of the solvers}\label{sec:solvers-interface}}

On the solver side we provide a suite\footnote{In the prototype we
  implement a subset of all unification rules, here are they listed:
  identityPlugin, propagateMetasEqPlugin, reduceLeftPlugin,
  reduceRightPlugin, leftMetaPlugin, rightMetaPlugin,
  typeConstructorPlugin, typeConstructorWithMetasPlugin,
  piEqInjectivityPlugin, tyEqInjectivityPlugin, consInjectivityPlugin,
  typeInjectivityPlugin, unificationStartMarker, unificationEndMarker}
of solvers for unification that handle different cases of the problem.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} solves syntactically equal terms}
\OtherTok{syntacticHandler ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ c)}
                 \OtherTok{=\textgreater{}} \DataTypeTok{HandlerType}\NormalTok{ c}
\OtherTok{syntacticSolver ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ c)}
                \OtherTok{=\textgreater{}} \DataTypeTok{SolverType} \DataTypeTok{Bool}
\OtherTok{syntactic ::} \DataTypeTok{Plugin}
\NormalTok{syntactic  }\OtherTok{=} \DataTypeTok{Plugin}\NormalTok{ \{ solver  }\OtherTok{=}\NormalTok{ syntacticSolver}
\NormalTok{                    , handler }\OtherTok{=}\NormalTok{ syntacticHandler}
                    \OperatorTok{...}
\NormalTok{                    \}}
\end{Highlighting}
\end{Shaded}

We first define the class of constraints that will be handled by the
solver via providing a ``handler'' -- function that decides whether a
given solver has to fire.\footnote{The code shown above and in the rest
  of the paper is close to the actual implementation, but has been
  simplified for presentation purposes.} In this case, this amounts to
checking that the constraint given is indeed an
\texttt{EqualityConstraint} and that the two terms given to it are
syntactically equal. Then we define the solver itself, which in this
case does not have to do anything except \texttt{return\ True} to
indicate that the constraint is solved, as we it only fires once it has
been cleared to do so by the handler and the equality has already been
checked. Finally, we register the solver by declaring it using a plugin
interface.

The reason for this separation between a decision procedure and the
execution of solver is to ensure separation between pontentially slow,
effectful solving and fast read-only decision-making in the handler. We
opt for this division since handlers will be run on many constraints
that do not fit, therefore any write effects would have to be rolled
back, while solvers should be fired only in cases when we can reasonably
hope that the constraint will be solved and the effects will not have to
be rolled back.

Similarly, we can define \texttt{leftMetaSolver} and
\texttt{rightMetaSolver} that only work on problems where only one of
the sides is a metavariable. Here the job of the solver is not as
trivial -- it has to check that the type of the other side indeed
matches the needed one and then register the instantiation of the
metavariable in the state.

Since the constraints they match on overlap we can provide priority
preferences, using the \texttt{pre} and \texttt{suc} fields of the
plugin interface. They are used to indicate whether the currently
defined plugin should run before or after, respectively, which other
plugins. At the time of running the compiler, these preferences are
loaded into a big pre-order relation for all the plugins, which is then
linearised and used to guide the solving procedure.

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{rightMetaPlugin ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
                \OtherTok{=\textgreater{}} \DataTypeTok{Plugin}\NormalTok{ cs}
\NormalTok{rightMetaPlugin }\OtherTok{=}
  \DataTypeTok{Plugin}\NormalTok{ \{ handler }\OtherTok{=}\NormalTok{ rightMetaHandler}
\NormalTok{         , solver  }\OtherTok{=}\NormalTok{ rightMetaSolver}
\NormalTok{         , symbol  }\OtherTok{=}\NormalTok{ rightMetaSymbol}
\NormalTok{         , pre }\OtherTok{=}\NormalTok{ []}
\NormalTok{         , suc }\OtherTok{=}\NormalTok{ [leftMetaSymbol]}
\NormalTok{         \}}
\end{Highlighting}
\end{Shaded}

\hypertarget{sec:solvers-implementation}{%
\subsection{Implementation of the solvers and unification
details}\label{sec:solvers-implementation}}

We implement a system close to the one described by
\citet{abelHigherOrderDynamicPattern2011}. We modularise the
implementation by mapping every function call in the simplification
procedure to a raised constraint and every simplification rule to a
separate solver. For example, the ``decomposition of functions''
\citep[fig.~2]{abelHigherOrderDynamicPattern2011} rule is translated to
the following implementation.

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{piEqInjectivityHandler ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
                       \OtherTok{=\textgreater{}} \DataTypeTok{HandlerType}\NormalTok{ cs}
\NormalTok{piEqInjectivityHandler constr }\OtherTok{=} \KeywordTok{do}
  \KeywordTok{let}\NormalTok{ eqcm }\OtherTok{=}\NormalTok{ match }\OperatorTok{@}\DataTypeTok{EqualityConstraint}\NormalTok{ constr}
  \KeywordTok{case}\NormalTok{ eqcm }\KeywordTok{of}
    \DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{EqualityConstraint}\NormalTok{ pi1 pi2 \_ \_) }\OtherTok{{-}\textgreater{}}
      \KeywordTok{case}\NormalTok{ (pi1, pi2) }\KeywordTok{of}
\NormalTok{        (}\DataTypeTok{I.Pi}\NormalTok{ \_ \_ \_, }\DataTypeTok{I.Pi}\NormalTok{ \_ \_ \_) }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{True}
\NormalTok{        \_ }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
\NormalTok{    \_ }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
\end{Highlighting}
\end{Shaded}

The handler is simply checking that both sides of the equality are
indeed Pi-types, and in case either of the matches fails, failure will
be reported and the solver will not be fired. The \texttt{match}
function above comes from the open datatype of constraints, checking if
\texttt{constr} can be projected from \texttt{cs} to
\texttt{EqualityConstraint}.

Now let us take a look at the solver.

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{piEqInjectivitySolver ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
                      \OtherTok{=\textgreater{}} \DataTypeTok{SolverType}\NormalTok{ cs}
\NormalTok{piEqInjectivitySolver constr }\OtherTok{=} \KeywordTok{do}
  \KeywordTok{let}\NormalTok{ (}\DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{EqualityConstraint}\NormalTok{ (}\DataTypeTok{I.Pi}\NormalTok{ a1 b1)}
\NormalTok{                                (}\DataTypeTok{I.Pi}\NormalTok{ a2 b2) \_ m)) }\OtherTok{=}
\NormalTok{        match }\OperatorTok{@}\DataTypeTok{EqualityConstraint}\NormalTok{ constr}
\NormalTok{  ma }\OtherTok{\textless{}{-}}\NormalTok{ constrainEquality a1 a2 }\DataTypeTok{I.Type}
\NormalTok{  (x, tyB1) }\OtherTok{\textless{}{-}}\NormalTok{ Unbound.unbind b1}
\NormalTok{  (y, tyB2\textquotesingle{}) }\OtherTok{\textless{}{-}}\NormalTok{ Unbound.unbind b2}
  \KeywordTok{let}\NormalTok{ tyB2 }\OtherTok{=}\NormalTok{ Unbound.subst y (}\DataTypeTok{I.Var}\NormalTok{ x) tyB2\textquotesingle{}}
\NormalTok{      mat }\OtherTok{=}\NormalTok{ I.identityClosure ma}
\NormalTok{  mb }\OtherTok{\textless{}{-}}\NormalTok{ extendCtx (}\DataTypeTok{I.TypeSig}\NormalTok{ (}\DataTypeTok{I.Sig}\NormalTok{ x e1 mat)) }\OperatorTok{$}
\NormalTok{                  constrainEquality tyB1 tyB2 }\DataTypeTok{I.Type}
  \KeywordTok{let}\NormalTok{ mbt }\OtherTok{=}\NormalTok{ bind x }\OperatorTok{$}\NormalTok{ I.identityClosure mb}
\NormalTok{  solveMeta m (}\DataTypeTok{I.Pi}\NormalTok{ mat mbt)}
  \FunctionTok{return} \DataTypeTok{True}
\end{Highlighting}
\end{Shaded}

Following pi-forall, we use the unbound-generics\footnote{\href{https://hackage.haskell.org/package/unbound-generics-0.4.3}{hackage.haskell.org/package/unbound-generics-0.4.3}}
library to deal with the names and binders. As it will fire after a
handler returns \texttt{True}, we can assume the pattern-matches will
not fail.

First we constrain the equality of the domain of the Pi-type:
\texttt{a1} and \texttt{a2}. The seemingly spurious metavariable
\texttt{ma} returned from this call serves as an anti-unification
\citep{pfenningUnificationAntiunificationCalculus1991} communication
channel. Every time an equality constraint is created we return a
metavariable that stands for the unified term. This metavariable is used
for unification problems that are created in the extended contexts -- in
this case second argument of the Pi-type, but also when solving
equalities concerning two data constructors. We do this to tackle the
``spine problem''
\citep[sec.~1.4]{victorlopezjuanPracticalHeterogeneousUnification2021}
-- as we operate according to the ``well-typed modulo constraints''
principle, essentially providing a placeholder that is guaranteed to
preserve well-typedness in the extended context. \footnote{\(\text{Tog}^{+}\)
  \citep{victorlopezjuanTog2020, victorlopezjuanPracticalHeterogeneousUnification2021}
  focuses on extending unification algorithm for the case where two
  sides of equality might not be of the same type, which is also a
  problem relevant for us. Their main argument against the usage of
  anti-unification in Agda provided there is that it is bug-prone. We
  think that in Agda the problems were stemming from the fact that
  anti-unification were implemented separately from unification, in
  which case it is indeed hard to keep the two in sync. There is no such
  duplication in our case since unification and anti-unification are
  one.} Finally \texttt{ma} has to be applied to a closure which will
keep track of the delayed substitution.

Then we can constrain the co-domains of Pi-types in an extended context.
In case one of solvers the constraints created in the extended context
might need to know the exact shape of \texttt{ma}, we can block on the
metavariable later, freezing the rest of the problem until it is
instantiated.

As for the actual unification steps, we implement them in a similar
fashion to simplification procedure. Take a look at the following
example, when only the left hand side of the constaint is an unsolved
meta:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{leftMetaSolver ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
               \OtherTok{=\textgreater{}} \DataTypeTok{SolverType}\NormalTok{ cs}
\NormalTok{leftMetaSolver constr }\OtherTok{=} \KeywordTok{do}
  \KeywordTok{let}\NormalTok{ (}\DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{EqualityConstraint}\NormalTok{ t1 t2 \_ m)) }\OtherTok{=}
\NormalTok{        match }\OperatorTok{@}\DataTypeTok{EqualityConstraint}\NormalTok{ constr}
\NormalTok{      (}\DataTypeTok{MetaVar}\NormalTok{ (}\DataTypeTok{MetaVarClosure}\NormalTok{ m1 c1)) }\OtherTok{=}\NormalTok{ t1}
\NormalTok{  mt2 }\OtherTok{\textless{}{-}}\NormalTok{ occursCheck m1 t2}
  \KeywordTok{case}\NormalTok{ mt2 }\KeywordTok{of}
    \CommentTok{{-}{-} indicates a failure in occurs{-}check}
    \DataTypeTok{Left}\NormalTok{ e }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
    \CommentTok{{-}{-} indicates a passed occurs{-}check}
    \DataTypeTok{Right}\NormalTok{ t2 }\OtherTok{{-}\textgreater{}}
      \KeywordTok{case}\NormalTok{ (invertClosureOn c1 (freeVarList t2)) }\KeywordTok{of}
        \DataTypeTok{Just}\NormalTok{ s }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
          \KeywordTok{let}\NormalTok{ st2 }\OtherTok{=}\NormalTok{ Unbound.substs s t2}
          \CommentTok{{-}{-} apply the subsitution}
\NormalTok{          solveMeta m1 st2}
          \CommentTok{{-}{-} instantiate the anti{-}unification variable}
\NormalTok{          solveMeta m st2}
          \FunctionTok{return} \DataTypeTok{True}
        \DataTypeTok{Nothing} \OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
\end{Highlighting}
\end{Shaded}

Once the occurs-check returns and if it was successful we apply the
inverted closure to the right-hand side of the equality.

By splitting up the rules into individual, simple solvers we can
compartmentalise the complexity of the unifier, making sure that each
rule is as decoupled from the others as possible. This does not
deteriorate the properties of the system, but does not help to guarantee
them either. We talk more about the challenge of proving correctness in
Section \ref{sec:limitations}.

\hypertarget{extending-unification}{%
\subsection{Extending unification}\label{extending-unification}}

Now that unification is implemented let us create a simple plugin that
makes certain symbols declared by the user injective
\citep{agdausersInjectiveUnificationPragma2023}. The actual
implementation is relatively simple and is not dissimilar to the
Pi-injectivity solver we showed above.

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{userInjectivitySolver ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
                      \OtherTok{=\textgreater{}} \DataTypeTok{SolverType}\NormalTok{ cs}
\NormalTok{userInjectivitySolver constr }\OtherTok{=} \KeywordTok{do}
  \KeywordTok{let}\NormalTok{ (}\DataTypeTok{Just} \DataTypeTok{EqualityConstraint}
\NormalTok{              (}\DataTypeTok{I.App}\NormalTok{ (}\DataTypeTok{I.Var}\NormalTok{ f) a)}
\NormalTok{              (}\DataTypeTok{I.App}\NormalTok{ (}\DataTypeTok{I.Var}\NormalTok{ g) b) \_ m)) }\OtherTok{=}
\NormalTok{        match }\OperatorTok{@}\DataTypeTok{EqualityConstraint}\NormalTok{ constr}
  \KeywordTok{if}\NormalTok{ f }\OperatorTok{==}\NormalTok{ g}
  \KeywordTok{then} \KeywordTok{do}
\NormalTok{    ifM (queryInjectiveDeclarations f)}
\NormalTok{        (}\KeywordTok{do}
\NormalTok{           ms }\OtherTok{\textless{}{-}}\NormalTok{ constrainEquality a b }\DataTypeTok{I.Type}
\NormalTok{           solveMeta m ms}
           \FunctionTok{return} \DataTypeTok{True}\NormalTok{)}
\NormalTok{        (}\FunctionTok{return} \DataTypeTok{False}\NormalTok{)}
  \KeywordTok{else} \FunctionTok{return} \DataTypeTok{False}
\end{Highlighting}
\end{Shaded}

where \texttt{queryInjectiveDeclarations} simply scans the available
declarations for a marker that \texttt{f} has been declared injective.

The only big thing left is to make sure that this solver fires at the
right time. This can only conflict with the ``decomposition of
neutrals'' rule, so we indicate to the solver dispatcher that our plugin
should run before the it:

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{userInjectivityPlugin ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
                      \OtherTok{=\textgreater{}} \DataTypeTok{Plugin}\NormalTok{ cs}
\NormalTok{userInjectivityPlugin }\OtherTok{=}
  \DataTypeTok{Plugin}\NormalTok{ \{ }\OperatorTok{...}
\NormalTok{         , solver }\OtherTok{=}\NormalTok{ userInjectivitySolver}
\NormalTok{         , symbol }\OtherTok{=} \DataTypeTok{PluginId} \OperatorTok{$} \StringTok{"userInjectivity"}
\NormalTok{         , pre }\OtherTok{=}\NormalTok{ [ unifyNeutralsDecomposition}
\NormalTok{                 , unificationEndMarkerSymbol]}
\NormalTok{         , suc }\OtherTok{=}\NormalTok{ [unificationStartMarkerSymbol]}
\NormalTok{         \}}
\end{Highlighting}
\end{Shaded}

This modification does not alter the core of the language.

\hypertarget{sec:casestudies}{%
\section{Case studies}\label{sec:casestudies}}

Once we implement basic elaboration and unification we can extend the
language. This is where we make use of the fact that the constraints
datatype is open.

We saw before in Section \ref{sec:implicit-arguments} that conventional
designs require separate handling of different kinds of implicit
variables. To simplify the design we would like to uniformly dispatch a
search for the solution, which would be handled by the a fitting solver.
We can achieve this by communicating the kind of the meta through its
type in the second argument of \texttt{FillInTheMeta\ m\ ty}. The
solvers then match on the shape of the type of the metavariable and
handle it in a case-specific manner: instance-search for type classes,
tactic execution for a tactic argument, waiting for regular unification
to solve the metavariable for regular implicit arguments.

In this section we will describe the implementation details of regular
implicit arguments (Section \ref{sec:case-implicits}), the
implementation of type classes added on top of the implicit arguments
(Section \ref{sec:case-typeclasses}). And finally (Section
\ref{sec:coercion-tactics}) we sketch the implementation of coercive
subtyping and tactic arguments.

\hypertarget{sec:case-implicits}{%
\subsection{Implicit arguments}\label{sec:case-implicits}}

As we mentioned, the rule for \texttt{Implicit} had to be added to the
syntax traversal part of the elaborator. In fact, we require not one but
two modifications that lie outside of the solvers-constraints part of
the system. The first one is, indeed, the addition of a separate case in
the syntax traversal, however contained. The second one lies in the
purely syntactical part of the compiler. We need the pre-processor to
insert the placeholder terms in the surface syntax. Particularly, we
need to desugar declarations of functions in the following way:

For any declaration of a function \texttt{f} with some implicit argument
\texttt{a}:

\begin{verbatim}
def f : {A : Type} -> {a : A} -> B a -> C
\end{verbatim}

Which should desugar to the following:

\begin{verbatim}
def f : (A : Implicit Type)
     -> (a : Implicit (deImp A))
     -> (b : B (deImp a)) -> C
\end{verbatim}

Then, for each function call \texttt{f\ b1} we need to insert the
corresponding number of implicit arguments, transforming it to
\texttt{f\ \_\ \_\ b1}.

For simple implicits this suffices -- as soon as we have the
placeholders in the surface syntax we create the constraints in the
\texttt{Implicit} case of the syntax traversal. The only solver that is
needed in this case is a trivial one that checks that the metavariable
has been instantiated indeed. This is because a regular implicit should
be instantiated by the unifier at some point later. This constraint
simply serves as a guarantee that all implicits have been instantiated.

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{fillInImplicitSymbol ::} \DataTypeTok{PluginId}

\OtherTok{fillInImplicitHandler ::}\NormalTok{ (}\DataTypeTok{FillInImplicit} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
                      \OtherTok{=\textgreater{}} \DataTypeTok{HandlerType}\NormalTok{ cs}
\NormalTok{fillInImplicitHandler constr }\OtherTok{=} \KeywordTok{do}
  \KeywordTok{let}\NormalTok{ ficm }\OtherTok{=}\NormalTok{ match }\OperatorTok{@}\DataTypeTok{FillInImplicit}\NormalTok{ constr}
  \KeywordTok{case}\NormalTok{ ficm }\KeywordTok{of}
    \DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{FillInImplicit}\NormalTok{ term ty) }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
      \KeywordTok{case}\NormalTok{ term }\KeywordTok{of}
        \DataTypeTok{MetaVar}\NormalTok{ (}\DataTypeTok{MetaVarClosure}\NormalTok{ mid \_) }\OtherTok{{-}\textgreater{}}
\NormalTok{          isMetaSolved mid}
\NormalTok{        \_ }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
    \DataTypeTok{Nothing} \OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}

\OtherTok{fillInImplicitPlugin ::}\NormalTok{ (}\DataTypeTok{FillInImplicit} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
                     \OtherTok{=\textgreater{}} \DataTypeTok{Plugin}\NormalTok{ cs}
\NormalTok{fillInImplicitPlugin }\OtherTok{=} \DataTypeTok{Plugin}\NormalTok{ \{}
\NormalTok{  solver }\OtherTok{=}\NormalTok{ fillInImplicitSolver,}
\NormalTok{  handler }\OtherTok{=}\NormalTok{ fillInImplicitHandler,}
\NormalTok{  symbol }\OtherTok{=}\NormalTok{ fillInImplicitSymbol,}
\NormalTok{  suc }\OtherTok{=}\NormalTok{ [],}
\NormalTok{  pre }\OtherTok{=}\NormalTok{ []}
\NormalTok{  \}}
\end{Highlighting}
\end{Shaded}

Additionally, there is a design choice to be made the implementation of
\texttt{Implicit\ A} and \texttt{deImp}. One option is to turn them into
a constructor and a projection of a record type, the other is to make
them computationally equivalent to \texttt{id}. In the former case we
need to manually unwrap them both in the pre-processor and during the
constraint-solving, but since the head symbol is distinct we can
guarantee that other solvers will not match on it, unless explicitly
instructed to. In the latter case, one has to be cautious of the order
in which the solvers are activated, particularly in the case of
different search procedures, should they be implemented. However, in the
example above it does not make a difference.

\hypertarget{sec:case-typeclasses}{%
\subsection{Type classes}\label{sec:case-typeclasses}}

Next, let us implement a plugin for type classes. Same as for the
implicit arguments in general, we rely here on a pre-processor to
transform user-friendly syntax to simple declarations. We do not focus
on this part, since such a pre-processor does a simple local
transformation.

Let us now go through an example of the elaboration process for a simple
term. For typeclasses we need a few declarations, listed below in
Agda-like syntax:

\begin{verbatim}
plus  :  {A : Type} -> {{PlusOperation A}}
     -> (a : A) -> (b : A) -> A

-- a semigroup on booleans, addition is OR
instance BoolPlus : PlusOperation Bool where
  plus = orb
\end{verbatim}

And the exemplary term itself is:

\begin{verbatim}
m = plus True False
\end{verbatim}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  First, the pre-processor eliminates the implicits and type class
  arguments. We end with the following declarations:

\begin{verbatim}
plus : (impA : Implicit Type)
    -> Instance PlusOperation (deImp impA)
    -> (a : deImp impA) -> (b :  deImp impA)
    -> deImp impA

instanceBoolPlus : InstanceT PlusOperation Bool
instanceBoolPlus =
  InstanceC (TypeClassC (Plus orb))

m = plus _ _ True False
\end{verbatim}

  Declaration \texttt{instance\ BoolPlus} turning into a constructor of
  \texttt{InstanceC} is precisely the part we need the pre-processor to
  do.
\item
  We go into the elaboration of \texttt{m} now. The elaborator applies
  \texttt{inferType\ (App\ t1\ t2)} rule four times and
  \texttt{checkType\ (Implicit)\ ty} twice on the two placeholders. The
  output of the elaborator is

\begin{verbatim}
m = plus ?_1 ?_2 True False
\end{verbatim}

  And the state of the elaborator contains four more constraints:

\begin{verbatim}
C1: FillInTheTerm ?_1 (Implicit Type)
C2: FillInTheTerm ?_2 (InstanceT PlusOperation
                                 (deImp ?_1))
C3: EqualityConstraint (deImp ?_1) Bool Type
C4: EqualityConstraint (deImp ?_1) Bool Type
\end{verbatim}

  The first two correspond to implicit arguments. The latter two are
  unification problems rendered into constraints.
\item
  Now we step into the constraint-solving world. First, the unifier
  solves the latter two constraints, instantiating \texttt{?\_1} to
  \texttt{Implicit\ Bool}. \texttt{C1} is then discarded as solved since
  \texttt{?\_1} is already instantiated to \texttt{Implicit\ Bool}.
  Next, the typeclass resolution launches a search for the instance of
  type \texttt{Instance\ PlusOperation\ Bool}.
\item
  This is where the type class plugin can take over. It transforms
  \texttt{C2:\ FillInTheTerm\ ?\_2\ (InstanceT\ PlusOperation\ Bool)} to
  \texttt{C5:\ InstanceSearch\ PlusOperation\ Bool\ ?\_2}. \texttt{C5}
  then gets matched by the search plugin for concrete instances, simply
  weeding through available declarations, looking for something of the
  shape \texttt{InstanceT\ PlusOperation\ Bool}. Such a declaration
  exists indeed and we can instantiate \texttt{?\_2} to
  \texttt{instanceBoolPlus}.
\end{enumerate}

Now let us take a look at the plugin for the constraint system. It is
contained in a single file
\href{https://github.com/liesnikov/extensible-elaborator/blob/elaborator-experiments/exel/src/Plugins/Typeclasses.hs}{\texttt{./exel/src/Plugins/Typeclasses.hs}}.
In it we define a new constraint type \texttt{InstanceSearch}, a solver
that transforms constraints of the shape
\texttt{FillInTheType\ ?\ (InstanceT\ \_)} to the new constraint, and
finally, a solver for instance search problems. The reason to transform
the original constraint to the new type is to ensure that no other
solver will make an attempt at this problem, therefore modulating
interactions with other plugins.

Finally, the implementation of search for concrete instances is quite
simple:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{instanceConcreteSolver constr }\OtherTok{=} \KeywordTok{do}
  \KeywordTok{let}\NormalTok{ (}\DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{InstanceSearch}\NormalTok{ tcn ty m)) }\OtherTok{=}
\NormalTok{        match }\OperatorTok{@}\DataTypeTok{InstanceSearch}\NormalTok{ constr}
\NormalTok{  alldecls }\OtherTok{\textless{}{-}}\NormalTok{ collectInstancesOf tcn }\OperatorTok{\textless{}$\textgreater{}}\NormalTok{ SA.getDecls}
\NormalTok{  sty }\OtherTok{\textless{}{-}}\NormalTok{ SA.substAllMetas ty}
  \KeywordTok{case}\NormalTok{ Map.lookup sty alldecls }\KeywordTok{of}
    \DataTypeTok{Just}\NormalTok{ i }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
\NormalTok{      SA.solveMeta m (}\DataTypeTok{I.Var}\NormalTok{ i)}
      \FunctionTok{return} \DataTypeTok{True}
    \DataTypeTok{Nothing} \OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
\end{Highlighting}
\end{Shaded}

We conjecture that implementation of Canonical Structures
\citep{mahboubiCanonicalStructuresWorking2013} would be relatively
simple in such a system due to openness of both unification procedure
and instance search.

\hypertarget{sec:coercion-tactics}{%
\subsection{Tactic arguments and coercions}\label{sec:coercion-tactics}}

In a similar fashion to the transformation of \texttt{FillInImplicit}
constraints to \texttt{InstanceSearch}, we could implement tactic
arguments and coercive subtyping.

The former would be quite similar to what we saw in the previous section
\ref{sec:case-typeclasses}, except we would have to resolve
\texttt{FillInImplicit} to \texttt{RunTactic} constraint or fill in
directly. The question of actually running the tactics is independent of
constraints, since it is essentially another type-checking action that
we accomodate already for blockers.

Coercive subtyping is of a slightly different nature. First, it would
rely on a pre-processor the heaviest out of all of the examples
described above, since naively one would insert a (potentially identity)
coercion in each argument of the application and potentially around
heads of applications and types of abstractions
\citep{tassiBiDirectionalRefinementAlgorithm2012}. This incurs not only
syntactic noise, but also potential performance penalty, which we
describe further in Section \ref{sec:limitations}. The second challenge
comes from the fact that unlike the above this feature would be
anti-modular in a sense that we need to access several constraints at
once, at least read-only. Consider the following example:

\begin{verbatim}
f : (A : Type) -> A -> A
arg : ArgTy

t : ExpTy
t = f _ arg
\end{verbatim}

This declaration would desugar to something like

\begin{verbatim}
t = coerce _ (f _ (coerce _ arg))
\end{verbatim}

which gives raise to two constraints, where \texttt{Coercion\ A\ B}
stands for the evidence of \texttt{A} being a subtype of \texttt{B}:

\begin{verbatim}
C1 : Coercion ArgTy ?1
C3 : Coercion ?1 ExpTy
\end{verbatim}

Clearly, we can not solve the first constraint in isolation and need to
consider all coercion constraints related to a particular type at the
same time. While we can accommodate such solvers, due to solvers having
write-access to the state, it becomes much harder to modulate
interactions between different plugins.

\hypertarget{sec:limitations}{%
\section{Limitations}\label{sec:limitations}}

While our design offers a lot of flexibility, it does not solve every
problem. In this section we describe a few examples of extensions that
do not quite fit in this framework and general limitations.

\hypertarget{handling-of-meta-variables-outside-of-definition-sites}{%
\subsection{Handling of meta-variables outside of definition
sites}\label{handling-of-meta-variables-outside-of-definition-sites}}

After we elaborate a definition there can still be unsolved
metavariables in it. This presents us with a design choice. The first
option is to freeze the metavariables and instantiate them per-use site,
essentially allowing for an expression to have multiple types. The
second option is to leave them up to be solved later, which might make
elaboration less predictable since now the use sites can influence
whether a particular definition type-checks. The third option is to
report them as an error and exit immediately.

In particular, the second option allows us to incorporate more involved
inference algorithms into the system. For example, if we were to
implement an erasure inference algorithm as described by
\citet{tejiscakDependentlyTypedCalculus2020}, we would have to create
metavariable annotations (described as ``evars'' in the paper) that can
be instantiated beyond the definition site. The downside is that the
defintions can change depending on their use and type-checking has to be
done effectively for the whole program and not per-definition.

\hypertarget{the-language-is-only-as-extensible-as-the-syntax-traversal-is}{%
\subsection{The language is only as extensible, as the syntax traversal
is}\label{the-language-is-only-as-extensible-as-the-syntax-traversal-is}}

Extensibility via constraints allows for a flexible user-specified
control flow as soon as we step into the constraints world. But control
flow of the syntax traversal for the basic language elaborator is fixed
by the basic language. For example, consider the following simplified
lambda-function type-checking function\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Rules/Term.hs\#L430-L518}{./src/full/Agda/TypeChecking/Rules/Term.hs\#L430-L518}}
from Agda:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checkLambda\textquotesingle{} cmp b xps typ body target }\OtherTok{=} \KeywordTok{do}
  \DataTypeTok{TelV}\NormalTok{ tel btyp }\OtherTok{\textless{}{-}}\NormalTok{ telViewUpTo numbinds target}
  \KeywordTok{if}\NormalTok{ size tel }\OperatorTok{\textless{}}\NormalTok{ numbinds }\OperatorTok{||}\NormalTok{ numbinds }\OperatorTok{/=} \DecValTok{1}
    \KeywordTok{then}\NormalTok{ dontUseTargetType}
    \KeywordTok{else}\NormalTok{ useTargetType tel btyp}
\end{Highlighting}
\end{Shaded}

Here Agda steps away from the bidirectional discipline and infers a
(lambda) function if the target type is not fully known. If in our
design the developer chooses to go only with a pure bidirectional style
of type-checking inferred lambda functions would be impossible to
emulate.

The solution in this case is effectively to replicate what Agda is doing
by implementing each type-checking rule in inference mode, essentially
factoring out \texttt{dontUseTargetType} in Agda's code snippet above.

Similarly, implementing something like commutativity and associativity
unifier plugins can require modifications in the core, since we two
terms that are equal during elaboration have to equal during core
type-checking too.

\hypertarget{lack-of-backtracking}{%
\subsection{Lack of backtracking}\label{lack-of-backtracking}}

We do not implement any backtracking in the solver dispatcher as it is
now. This means that every step taken is committing, which can be a
limitation in cases where one would like to have backtracking -- for
example, Agda's instance arguments search (with
\texttt{-\/-overlapping-instances} flag)
\citep[chap.~3.18]{theagdateamAgdaUserManual2023a}, as well as type
classes in Lean \citep{selsamTabledTypeclassResolution2020}.

Backtracking in principle could be achieved by tracking changes to the
state of the elaborator and the production graph for constraints, but
such a system would be rather awkward. Alternatively, such a solver can
be implemented within one solver, removing the extension points, but
allowing arbitrary control-flow within the algorithm.

\hypertarget{reliance-on-a-pre-processor}{%
\subsection{Reliance on a
pre-processor}\label{reliance-on-a-pre-processor}}

This work crucially relies on a pre-processor of some kind, be it macro
expansion or some other way to extend the parser with custom desugaring
rules. In particular, in order to implement n-ary implicit arguments
correctly and easily we need the pre-processor to expand them to the
right arity, similar to
Matita\citep[chap.~5]{tassiBiDirectionalRefinementAlgorithm2012} and
others
\citep{serranoQuickLookImpredicativity2020, kovacsElaborationFirstclassImplicit2020}.

\hypertarget{eager-reduction-and-performance}{%
\subsection{Eager reduction and
performance}\label{eager-reduction-and-performance}}

As in the pre-processing step we would have to insert a fair number of
wrappers and un-wrappers for all implicits and even more for coercions,
a performance regression would be expected due a lot of spurious
computation steps.

We expect this to be a major concern for coercive subtyping. As one way
to mitigate we suggest a discipline with constraint solvers latching
onto non-reduced types and terms in constraints. In that case, we can
get around with a trick borrowed from Coq, where the wrappers and
unwrappers are identity functions and \texttt{coerce\ f\ t} computes to
\texttt{f\ t}. This also means that constraints can/have to match on
unreduced types in the e.g.~\texttt{FillInTheTherm}. In fact, we already
do this to an extent for a different reason -- since the calculus allows
only fully applied type constructors, we have to wrap each type class
constructor in a lambda-abstraction for it to appear as an argument to
\texttt{TypeClassT\ typeClassName\ argType}. Or, concretely, we have to
define and use
\texttt{PlusOperation\textquotesingle{}\ =\ \textbackslash{}A\ .\ PlusOperation\ A}
in place of \texttt{PlusOperation} in the elaboration example in Section
\ref{sec:case-typeclasses}.

\hypertarget{proving-correctness}{%
\subsection{Proving correctness}\label{proving-correctness}}

As soon as we allow the users to implement their own solvers, there is
little we can say about correctness of the system as a whole without
imposing proof obligations on the plugin writers. However, if one is
willing to do that we can conjecture solution and type preservation --
Theorems 2 and 3, respectively \citep{abelHigherOrderDynamicPattern2011}
-- assuming that each solver on its own satisfies these properties. For
termination it is similar -- we can conjecture that if every solver in
the system ``reduces the weight'' of the unification problem, in terms
of Theorem 1 by \citet{abelHigherOrderDynamicPattern2011}, we can hope
for termination guarantees. This is simply a consequence of the fact
that we do not invent a new unification algorithm, but rather provide
means of easier implementation for it.

\hypertarget{sec:related_work}{%
\section{Related work}\label{sec:related_work}}

We are certainly not the first ones to try to tackle the extensibility
of a language implementation. Dependently typed language implementations
usually consist of at least four parts: parser, elaborator, core
type-checker, and proper back-end. The back-end part is currently
irrelevant to our interests, since for a language to be specified
usually means for specification of the core, anything that happens after
the core does not extend the language, but rather tries to preserve its
semantics in some form. Therefore we're left with three parts: parser,
elaborator, and core type-checker.

We see parser or syntax extensibility as a necessary part of an
extensible language. This problem has been studied extensively in the
past and has a multitude of existing solutions. Macros are one of them
and are utilised heavily in various forms in almost all established
languages
\citep{thecoqdevelopmentteamCoqProofAssistant2022, theagdateamAgdaUserManual2023a, ullrichNotationsHygienicMacro2020}
and can be powerful enough to build a whole language around
\citep{changDependentTypeSystems2019}.

Core extensibility, on the other hand, appears to be a problem with too
many degrees of freedom. Andromeda
\citep{bauerDesignImplementationAndromeda2018, bauerEqualityCheckingGeneral2020}
made an attempt at extensible definitional equality but is quite far
from a usable dependently typed language. Agda's philosophy allows
developers to experiment with the core but also results in a larger
amount of unexpected behaviours. In general, modifications of the core
rules will result in fundamental changes in the type theory, which can
break plenty of important properties like soundness or subject
reduction.

This leaves us with the question about the extensibility of an
elaborator. We will make a division here between syntax traversals,
constraint solving and all other features. The syntax traversal part of
the elaborator is relatively stable and commonly implemented following a
bidirectional discipline to some extent
\citep{norellPracticalProgrammingLanguage2007, tassiBiDirectionalRefinementAlgorithm2012, ferreiraBidirectionalElaborationDependently2014},
so there seems little reason to make it extensible.

GHC has a plugin system that allows users to dynamically add custom
constraint solvers, but the type of constraints itself is not
extensible\footnote{\href{https://gitlab.haskell.org/ghc/ghc/-/wikis/plugins/type-checker/notes}{gitlab.haskell.org/ghc/ghc/-/wikis/plugins/type-checker/notes}}
\citep{peytonjonesTypeInferenceConstraint2019, vytiniotisOutsideInModularType2011, peytonjonesPracticalTypeInference2007}.

Coq \citep{thecoqdevelopmentteamCoqProofAssistant2022}, being one of the
most popular proof assistants, invested a lot effort into user-facing
features: work on tactics like a new tactic engine
\citep{spiwackVerifiedComputingHomological2011} and tactic languages
(Ltac2 \citep{pedrotLtac2TacticalWarfare2019}, SSReflect
\citep{gonthierSmallScaleReflection2008}, etc.), the introduction of a
virtual machine for performance
\citep{gregoireCompiledImplementationStrong2002} and others. However,
the implementation is quite hard to extend. One either has to modify the
source code, which is mostly limited to the core development team, as
seen from the
\href{https://github.com/coq/coq/graphs/contributors}{graph}. Or one has
to use Coq plugins system, which is rather challenging, and in the end
the complexity of it gave rise to TemplateCoq
\citep{malechaExtensibleProofEngineering2014}.

Agda has historically experimented a lot with different extensions for
both type system and the elaborator, even though the design doesn't
accomodate these changes naturally. Instead, each of these extensions is
spread throughout many different parts of the code-base\footnote{some
  recent examples:
  \href{https://github.com/agda/agda/pull/6385}{github.com/agda/agda/pull/6385},
  \href{https://github.com/agda/agda/pull/6354}{github.com/agda/agda/pull/6354}.}.

Lean introduced elaborator extensions
\citep{leonardodemouraLeanMetaprogramming2021, ullrichNotationsHygienicMacro2020}.
They allow the user to overload the commands, but if one defines a
particular elaborator it becomes hard to interleave with others. In a
way, this is an imperative view on extensibility.

Idris
\citep{bradyIdrisGeneralpurposeDependently2013, christiansenElaboratorReflectionExtending2016}
appeared as a programming language first and proof-assistant second and
does not provide either a plugin or hook system at all, except for
reflection. Idris also focuses on tactics as the main mechanism for
elaboration.

Turstile+ by \citet{changDependentTypeSystems2019} uses macros to
elaborate surface syntax to a smaller core. Macros allow them to
modularly implement individual features, however, exactly as for Lean,
combining different features which happens a lot in practise requires
the user to re-define all macros from scratch.

TypOS \citep{allaisTypOSOperatingSystem2022a, guillaumeallaisTypOS2022}
is perhaps the closest to our work, but there are two important
differences. First, they are building a domain-specific language for
building type-checkers, while our design is language-agnostic, as long
as it can model extensible datatypes in some capacity. Second, their
approach settles features of the language as they are decided by the
main developer and doesn't concern future changes and evolution.
Finally, we try to stay close to the designs of existing dependently
typed languages, while TypOS doesn't showcase support for them.

\hypertarget{future-work}{%
\section{Future work}\label{future-work}}

We see three main prospects for future work:

\begin{itemize}
\tightlist
\item
  \textbf{Exploration of different kinds of metavariables.} Currently we
  implement metavariables only for terms, while for some applications
  such as erasure \citep{tejiscakDependentlyTypedCalculus2020} or
  irrelevance inference, it would be beneficial to have metavariables
  representing erasure and relevance annotations. Alternatively, we can
  introduce metavariables for names and implement data constructor
  disambiguation in a simpler manner, since as of now we simply block on
  expected type to disambiguate.
\item
  \textbf{Rendering more elements of the elaborator as constraints} --
  currently components such as occurs-checker and reduction are simple
  functions. Including them into the constraint machinery would make the
  implementation more uniform and allow users to extend them.
\item
  \textbf{Potential optimisations}. Currently our system has a lot of
  room for potential optimisations. The first step would be to allow
  handlers to pass some information to their respective solvers, which
  require introduction an existential type in the plugin. Additionally
  some expandable per-plugin store for the solvers would be handy, for
  example to not recompute type class instances on every invocation.
  Somewhat more challenging, one can imagine a caching system for
  constraints, allowing to avoid solving the same constraint twice. This
  might be beneficial for reduction, since as of now we do a lot of
  redundant computations. However, the memory usage might be
  prohibitive. Finally, we would also like to explore possibilities for
  concurrent solving, similar to the future plans of
  \citet{allaisTypOSOperatingSystem2022a} with LVars for metavariables
  \citep{kuperLatticebasedDataStructures2015}.
\end{itemize}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\end{CSLReferences}

  \bibliographystyle{ACM-Reference-Format}
  \bibliography{bib.bib}

\end{document}

### 2023-11-30 exel/src/TypeCheck/Elaborator.hs
simplify local var lookup
don't forget to reduce substituted metavar
add Logic and HW to test suite
move others to the old directory

@@ -577,7 +577,7 @@ checkType (S.Case scrut alts) ty = do
         ensureTCon term = Env.err [DS "can't verify that",
                                      DD term,
                                      DS "has TCon as head-symbol"]
-    (c, args) <- ensureTCon =<< SA.substMetas sty
+    (c, args) <- ensureTCon =<< fmap fst . whnf =<< SA.substMetas sty
     let checkAlt :: (MonadElab c m) => S.Match -> m I.Match
         checkAlt (S.Match bnd) = do
           (pat, body) <- Unbound.unbind bnd
@@ -868,8 +868,7 @@ elabEntry (S.Def n term) = do
                     DD sig
                   ]
            in do
-                elabterm <- Env.extendCtx (I.TypeSig sig) $
-                              checkType term (I.sigType sig) `catchError` handler
+                elabterm <- Env.extendCtx (I.TypeSig sig) $ checkType term (I.sigType sig) `catchError` handler
                 --Env.warn [ DS "elaborated term before substitution"
                 --         , DD elabterm
                 --         , DS $ show elabterm

### 2023-11-30 exel/src/TypeCheck/StateActions.hs

@@ -260,11 +260,12 @@ getGlobalVars = do
   let vars = decls >>= (\x -> case x of (TypeSig (Sig n _ _)) -> [n]; _ -> [])
   return vars
 
-getLocalFreeVars :: (MonadTcReader m) => Term -> m [TName]
+getLocalFreeVars :: (MonadTcReader m, MonadTcReaderEnv m) => Term -> m [TName]
 getLocalFreeVars t = do
   let lfvs = freeVarList t
-  globals <- names <$> askDecls
-  return . filter ((/= "?") . Unbound.name2String) . filter (`notElem` globals) $ lfvs
+--  globals <- names <$> askDecls
+  locals <- names <$> Env.getLocalCtx
+  return . filter ((/= "?") . Unbound.name2String) . filter (`elem` locals) $ lfvs
   where
     names :: [Decl] -> [TName]
     names [] = []

### 2023-11-30 exel/test/Main.hs

@@ -11,5 +11,5 @@ main = do
    defaultMain $ testGroup "All tests" [
        printparse
      , testtypecheckfiles
---     , exampletypecheckfiles
+     , exampletypecheckfiles
      ]

### 2023-11-30 exel/test/Tests/Full.hs

@@ -61,28 +61,29 @@ testtypecheckfiles = testGroup "test files typechecked" [
 
 exampletypecheckfiles :: TestTree
 exampletypecheckfiles = testGroup "example files typechecked" [
-  testFile "BoolLib.pi",
-  testFile "Equality.pi",
-  testFile "Equal.pi",
-  testFile "FinHw.pi",
-  testFile "Fin.pi",
-  testFile "Fix.pi",
-  testFile "Hurkens.pi",
-  testFile "Hw1.pi",
-  testFile "Hw2.pi",
-  testFile "Lambda0.pi",
-  testFile "Lambda1.pi",
-  testFile "Lambda2.pi",
-  testFile "Lambda.pi",
-  testFile "Lec1.pi",
-  testFile "Lec2.pi",
-  testFile "Lec3.pi",
-  testFile "Lec4.pi",
-  testFile "Lennart.pi",
-  testFile "List.pi",
-  testFile "Logic.pi",
-  testFile "NatChurch.pi",
-  testFile "Nat.pi",
-  testFile "Product1.pi",
-  testFile "Product.pi",
-  testFile "Vec.pi"]
+--  , testFile "BoolLib.pi"
+--  , testFile "Equality.pi"
+--  , testFile "Equal.pi"
+--  , testFile "FinHw.pi"
+--  , testFile "Fin.pi"
+--  , testFile "Fix.pi"
+--  , testFile "Hurkens.pi"
+    testFile "Hw1.pi"
+--  , testFile "Hw2.pi"
+--  , testFile "Lambda0.pi"
+--  , testFile "Lambda1.pi"
+--  , testFile "Lambda2.pi"
+--  , testFile "Lambda.pi"
+--  , testFile "Lec1.pi"
+--  , testFile "Lec2.pi"
+--  , testFile "Lec3.pi"
+--  , testFile "Lec4.pi"
+--  , testFile "Lennart.pi"
+--  , testFile "List.pi"
+  , testFile "Logic.pi"
+--  , testFile "NatChurch.pi"
+--  , testFile "Nat.pi"
+--  , testFile "Product1.pi"
+--  , testFile "Product.pi"
+--  , testFile "Vec.pi"
+  ]

### 2023-12-01 exel/src/Reduction.hs
fix reduction of subst

@@ -80,13 +80,13 @@ whnfc ustr (LetPair a bnd) = do
 
 
 whnfc ustr (Subst tm pf) = do
-  (tm', mb) <- whnfc ustr tm
+  (pf', mb) <- whnfc ustr pf
   case mb of
-    Just b -> return (Subst tm' pf, Just b)
+    Just b -> return (Subst tm pf', Just b)
     Nothing -> do
-      case tm' of
-        Refl -> whnfc ustr pf
-        _ -> return (Subst tm' pf, Nothing)
+      case pf' of
+        Refl -> whnfc ustr tm
+        _ -> return (Subst tm pf', Nothing)
 
 
 whnfc ustr (Case scrut mtchs) = do

### 2023-12-01 exel/test/Tests/Full.hs
add equality test
comment out broken ones

@@ -62,13 +62,13 @@ testtypecheckfiles = testGroup "test files typechecked" [
 exampletypecheckfiles :: TestTree
 exampletypecheckfiles = testGroup "example files typechecked" [
 --  , testFile "BoolLib.pi"
---  , testFile "Equality.pi"
+      testFile "Equality.pi"
 --  , testFile "Equal.pi"
 --  , testFile "FinHw.pi"
 --  , testFile "Fin.pi"
 --  , testFile "Fix.pi"
 --  , testFile "Hurkens.pi"
-    testFile "Hw1.pi"
+    , testFile "Hw1.pi"
 --  , testFile "Hw2.pi"
 --  , testFile "Lambda0.pi"
 --  , testFile "Lambda1.pi"

### 2023-12-01 exel/test/Tests/Full.hs
add nat tests

@@ -68,7 +68,7 @@ exampletypecheckfiles = testGroup "example files typechecked" [
 --  , testFile "Fin.pi"
 --  , testFile "Fix.pi"
 --  , testFile "Hurkens.pi"
-    , testFile "Hw1.pi"
+    ,  testFile "Hw1.pi"
 --  , testFile "Hw2.pi"
 --  , testFile "Lambda0.pi"
 --  , testFile "Lambda1.pi"
@@ -80,9 +80,9 @@ exampletypecheckfiles = testGroup "example files typechecked" [
 --  , testFile "Lec4.pi"
 --  , testFile "Lennart.pi"
 --  , testFile "List.pi"
-  , testFile "Logic.pi"
+    , testFile "Logic.pi"
 --  , testFile "NatChurch.pi"
---  , testFile "Nat.pi"
+    , testFile "Nat.pi"
 --  , testFile "Product1.pi"
 --  , testFile "Product.pi"
 --  , testFile "Vec.pi"

### 2023-12-01 exel/src/TypeCheck/Elaborator.hs
partial fix to contra
add a couple more test files to typecheck

@@ -496,8 +496,8 @@ checkType (S.Contra p) typ = do
           _ -> return Nothing
   ret <- SA.createMetaTerm typ
   blockAndReblockUntil blockboth detector $ do
-    (a', ba) <- whnf a
-    (b', bb) <- whnf b
+    (a', ba) <- whnf =<< SA.substAllMetas a
+    (b', bb) <- whnf =<< SA.substAllMetas b
     case (a, b) of
       (I.DCon da _, I.DCon db _)
         | da /= db ->
@@ -508,9 +508,9 @@ checkType (S.Contra p) typ = do
       (_, _) ->
         Env.err
           [ DS "I can't tell that",
-            DD a,
+            DD a',
             DS "and",
-            DD b,
+            DD b',
             DS "are contradictory",
             DS "the reduction is blocked on",
             DD ba,

### 2023-12-01 exel/test/Tests/Full.hs

@@ -68,8 +68,8 @@ exampletypecheckfiles = testGroup "example files typechecked" [
 --  , testFile "Fin.pi"
 --  , testFile "Fix.pi"
 --  , testFile "Hurkens.pi"
-    ,  testFile "Hw1.pi"
---  , testFile "Hw2.pi"
+    , testFile "Hw1.pi"
+    , testFile "Hw2.pi"
 --  , testFile "Lambda0.pi"
 --  , testFile "Lambda1.pi"
 --  , testFile "Lambda2.pi"

### 2024-02-28 paper/main.tex
switch to tex build by default
add a comment explaining what the changes are to the tex file
grammar
don't forget to actually add the header

@@
 % Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
 \PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
 %
-\documentclass[
-  sigconf,
-  screen,
-  review]{acmart}
+\documentclass[sigconf, screen]{acmart}
 \usepackage{amsmath,amssymb}
@@
 }{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
 \usepackage{xcolor}
@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\widowpenalty10000
-\clubpenalty10000
+\widowpenalty=10000
+\clubpenalty=10000
+\displaywidowpenalty=10000
 \author{Bohdan Liesnikov}
@@
 As a consequence the developers have to make a lot of adaptations in the implementations, which leads to codebases that are hard to maintain.
-In this work we focus on a part of the type-checker called the elaborator that transforms user-friendly surface syntax to a small well-behaved core theory.
-Our core idea is to use an open datatype for constraints and plugin system for solvers.
+In this work we focus on a part of the type-checker called the elaborator that transforms user-friendly surface syntax into a small well-behaved core theory.
+Our core idea is to use an open datatype for constraints and a plugin system for solvers.
 This allows for a more compact base elaborator implementation while enabling extensions to the type system.
@@
 \usepackage[]{natbib}
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+%\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
   pdftitle={ExEl: Building an Elaborator Using Extensible Constraints},
-  colorlinks=true,
-  linkcolor={Maroon},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
+  hidelinks,
   pdfcreator={LaTeX via pandoc}}
@@
 
-\hypertarget{sec:introduction}{%
-\section{Introduction}\label{sec:introduction}}
-
-Strongly-typed languages allow us to catch certain classes of bugs at
-compile-time by checking the implementation against the type-signature.
-When the types are provided by the user they can be viewed as a form of
-specification, constraining the behaviour of the programs. This comes
-with the benefit of more static guarantees but with an increased toll on
-the user to supply more precise information about the program. Many
-languages choose to infer types, but another option is to use the
-information in the types to infer parts of the program. This idea was
-aptly worded by Conor McBride as ``Write more types and fewer
-programs.''
-\citep[chap.~2.1]{ptoopTypeInferenceThought2022, mcbrideEpigramPracticalProgramming2005}
-Some examples of this include overloaded functions in Java, implicits in
-Scala, and type classes in Haskell.
-
-In dependently typed languages like Agda
-\citep{norellPracticalProgrammingLanguage2007, theagdateamAgdaUserManual2023a},
-Coq \citep{thecoqdevelopmentteamCoqProofAssistant2022} or Idris
-\citep{bradyIdrisGeneralpurposeDependently2013} the types can be much
-more precise. This gives us an opportunity to infer even larger parts of
-the program from the type in the process of elaboration. Examples
-include implicit arguments in Agda, implicit coercions in Coq, and
-tactic arguments in Idris. The inference or ``solving'' can be not only
-automatic but also interactive or partially automatic. Holes in Agda and
-proof obligations in Coq are examples of interactive, while Canonical
-Structures \citep{mahboubiCanonicalStructuresWorking2013} in Coq and
-program-synthesis for holes in Haskell
-\citep{koppelSearchingEntangledProgram2022} are partially automatic. All
-these mechanisms use different algorithms and have various degrees of
-extensibility. These are typically not very well isolated from each
-other and can therefore interact in unexpected ways, for example in case
-between implicits and instance search
-\citep{agdausersPerformanceRegressionIssue2018}. This puts a toll on the
-language developer to modulate the interactions and for the user to
-understand these features.
-
-In all of these examples, solvers evolved organically over time together
-with the language. Coq historically struggled with similar issues in the
-elaborator: for example, Canonical Structures was not properly
-documented for 15 years \citep{mahboubiCanonicalStructuresWorking2013}.
-Agda experimented with features baked into the core of the type system
--- like sized types -- which brought their own solver infrastructure
-\citep{abelExtensionMartinLofType2016}. Lean 4 is aiming to allow the
-users to develop new surface-level features
-\citep{leonardodemouraLeanMetaprogramming2021} using elaboration monads
-\citep{mouraLeanTheoremProver2021}, somewhat akin to elaborator
-reflection in Idris
-\citep{christiansenElaboratorReflectionExtending2016}, but Lean 3 was
-built in a more conventional way \citep{demouraLeanTheoremProver2015a}.
-
-One common piece of infrastructure needed by these inference algorithms
-are metavariables, also known as ``existential variables''
-\citep[chap.~2.2.1]{thecoqdevelopmentteamCoqProofAssistant2022}, which
-represent as-of-yet unknown parts of the program. Another one is the
-need is to constrain two terms to be equal, known as unification.
-Metavariables and unification are heavily used throughout the compiler,
-for inference of implicit arguments and for general type-checking,
-making the compiler sensitive towards changes in unification algorithms.
-Because of the complexity unification breaking changes are often
-discovered only when ran against a large existing project on CI, like
-\texttt{cubical} or \texttt{stdlib} for Agda or \texttt{unimath} for
-Coq.
-
-In this paper we propose a new architecture for an extensible elaborator
-for dependently typed languages. The idea is to provide an API for
-developers to tap into the elaboration procedure with custom solvers
-that can manipulate metavariables and constraints placed on them. This
-design separates the `what' the solvers are doing from the `when',
-making the interaction points between different parts of the
-type-checker explicit. As a result, this allows the developer to reason
-more easily about exceptions and asynchronicity in the type-checker and
-add new features in a more contained fashion. Practically, this means
-that each feature is contained within one module, as opposed to being
-spread around the codebase.
-
-Contributions:
-
-\begin{itemize}
-\tightlist
-\item
-  We propose a new design blueprint for implementing a language that is
-  extensible with new constraints and new solvers. It supports type
-  classes (Section \ref{sec:case-typeclasses}), implicit arguments
-  (Section \ref{sec:case-implicits}), implicit coercions and tactic
-  arguments (Section \ref{sec:coercion-tactics}).
-\item
-  We propose a view on metavariables as communication channels for the
-  solvers, drawing parallels with asynchronous programming primitives
-  (Section \ref{sec:solvers-implementation}).
-\item
-  We decompose the usual components of a type-checker, like the unifier
-  in Agda, into a suite of solvers which can be extended and interleaved
-  by user-provided plugins (Section
-  \ref{sec:constraints_and_unification}).
-\item
-  Following the blueprint, we present a prototype implementation of a
-  dependently typed language available at
-  \href{https://github.com/liesnikov/extensible-elaborator}{github.com/liesnikov/extensible-elaborator}.
-\end{itemize}
-
-\hypertarget{sec:unification_constraint_based_elaboration_and_design_challanges}{%
-\section{Unification, constraint-based elaboration and design
-challenges}\label{sec:unification_constraint_based_elaboration_and_design_challanges}}
-
-Constraints have been an integral part of compilers for strongly-typed
-languages for a while \citep{oderskyTypeInferenceConstrained1999}. For
-example, both Haskell \citep{vytiniotisOutsideInModularType2011} and
-Agda \citep[ chap.~3]{norellPracticalProgrammingLanguage2007} use
-constraints extensively. In the former case, they are even reflected and
-can be manipulated by the user
-\citep[chap.~6.10.3]{orchardHaskellTypeConstraints2010a, ghcdevelopmentteamGHCUserGuide2022}.
-This has proved to be a good design decision for GHC, as is reflected,
-for example in a talk by \citet{peytonjonesTypeInferenceConstraint2019},
-as well as in a few published sources
-\citep{vytiniotisOutsideInModularType2011, peytonjonesPracticalTypeInference2007}.
-
-In the land of dependently typed languages constraints are much less
-principled.
-Agda has a family of constraints\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Monad/Base.hs\#L1157-L1191}{./src/full/Agda/TypeChecking/Monad/Base.hs\#L1157-L1191}.
-  Here and hencefoth we shorten the links in footnotes to paths in the
-  repository, the source code can be found at
-  \href{https://github.com/agda/agda/blob/v2.6.4/}{github.com/agda/agda/blob/v2.6.4/}.}
-that grew organically, currently counting 19 constructors. Idris
-technically has constraints\footnote{\href{https://github.com/idris-lang/Idris2/blob/e673d05a67b82591131e35ccd50fc234fb9aed85/src/Core/UnifyState.idr}{./src/Core/UnifyState.idr}
-  at
-  \href{https://github.com/idris-lang/Idris2/blob/e673d05a67b82591131e35ccd50fc234fb9aed85}{github.com/idris-lang/Idris2/blob/e673d0}},
-with the only two constructors being equality constraints for two terms
-and for two sequences of terms.
-The same\footnote{\href{https://github.com/leanprover/lean4/blob/0a031fc9bbb43c274bb400f121b13711e803f56c/src/Lean/Meta/Match/Basic.lean\#L161}{./src/Lean/Meta/Match/Basic.lean\#L161}
-  at
-  \href{https://github.com/leanprover/lean4/blob/0a031fc9bbb43c274bb400f121b13711e803f56c/}{github.com/leanprover/lean4/blob/0a031f}}
-holds for Lean. These languages either use constraints in a restricted,
-single use-case manner -- namely, for unification -- or in an ad-hoc
-manner.
-
-In this section we will demonstrate why a more methodical approach to
-constraints will result in more robust elaborators across the board. We
-go over three typical challenges that come up when building a compiler
-for a dependently typed language and the way they are usually solved. We
-cover unification of the base language and the complexity of managing
-the state of the unifier in Section \ref{sec:conversion_checking}. Then
-we take a look at different kinds of implicit arguments and their
-implementation in Section \ref{sec:implicit-arguments}. We briefly touch
-on the problem of extending the unifier in Section
-\ref{sec:extending-unification}. Finally, we summarize the issues in
-Section \ref{sec:summary-of-the-issues}.
-
-\hypertarget{sec:conversion_checking}{%
-\subsection{Unification in presence of
-meta-variables}\label{sec:conversion_checking}}
-
-As mentioned in the introduction, in the process of type-checking we use
-unification to compare terms, which is notoriously hard to implement.
-The complexity stems from the desire of compiler writers to implement
-the most powerful unifier, while being limited by the fact that
-higher-order unification is undecidable in general. Some of this
-complexity is unavoidable, but we can manage it better by splitting up
-the unifier into smaller modular parts. In practice, this means that one
-does not have to fit together an always-growing unifier but can instead
-write different cases separately.
-
-An example from Agda's elaborator is type-driven unification, which is
-spread between about a hundred functions and 2200 lines of
-code\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs}{./src/full/Agda/TypeChecking/Conversion.hs}}.
-Each of the functions implements part of the ``business logic'' of the
-unifier. But all of them contain a lot of code that deals with
-bookkeeping related to metavariables and constraints:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  They have to throw and catch exceptions, driving the control flow of
-  the unification.
-\item
-  They have to compute blockers that determine when a postponed
-  constraint is retried.
-\item
-  They have to deal with cases where either or both of the sides
-  equation or its type are either metavariables or terms whose
-  evaluation is blocked on some metavariables.
-\end{enumerate}
-
-Concretely, this code uses functions like \texttt{noConstraints} and
-\texttt{dontAssignMetas} which rely on specific behaviour of the
-constraint solver system. Other functions like \texttt{abortIfBlocked},
-\texttt{reduce} and \texttt{catchConstraint}/\texttt{patternViolation}
-force the programmer to make a choice between letting the constraint
-system handle blockers or doing it manually. These things are known to
-be brittle and pose an increased mental overhead when making changes.
-
-The unifier is heavily used throughout the type-checker: either as
-function calls to \texttt{leqType} when type-checking terms and
-\texttt{compareType} when type-checking applications, or as raised
-constraints \texttt{ValueCmp} and \texttt{SortCmp} from
-\texttt{equalTerm} while checking applications or definitions,
-\texttt{ValueCmpOnFace} from \texttt{equalTermOnFace} again while
-checking applications. At the same it is unintuitive and full of
-intricacies as indicated by multiple comments\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs\#L484-L485}{./src/full/Agda/TypeChecking/Conversion.hs\#\#L484-L485},
-  \href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs\#L541-L549}{./src/full/Agda/TypeChecking/Conversion.hs\#L541-L549}}.
-
-We would like the compiler-writer to separate the concerns of managing
-constraints and blockers from the actual logic of the comparison
-function. In fact, if we zoom in on the \texttt{compareAtom} function,
-the core can be expressed in about 20 lines\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs\#L550-L599}{./src/full/Agda/TypeChecking/Conversion.hs\#L550-L599}}
-of simplified code, stripping out size checks, cumulativity, polarity,
-and forcing. This is precisely what we would like the developer to
-write.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{case}\NormalTok{ (m, n) }\KeywordTok{of}
-\NormalTok{  (}\DataTypeTok{Lit}\NormalTok{ l1, }\DataTypeTok{Lit}\NormalTok{ l2) }\OperatorTok{|}\NormalTok{ l1 }\OperatorTok{==}\NormalTok{ l2 }\OtherTok{{-}\textgreater{}} \FunctionTok{return}\NormalTok{ ()}
-\NormalTok{  (}\DataTypeTok{Var}\NormalTok{ i es, }\DataTypeTok{Var}\NormalTok{ i\textquotesingle{} es\textquotesingle{}) }\OperatorTok{|}\NormalTok{ i }\OperatorTok{==}\NormalTok{ i\textquotesingle{} }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
-\NormalTok{      a }\OtherTok{\textless{}{-}}\NormalTok{ typeOfBV i}
-\NormalTok{      compareElims [] [] a (var i) es es\textquotesingle{}}
-\NormalTok{  (}\DataTypeTok{Con}\NormalTok{ x ci xArgs, }\DataTypeTok{Con}\NormalTok{ y \_ yArgs) }\OperatorTok{|}\NormalTok{ x }\OperatorTok{==}\NormalTok{ y }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
-\NormalTok{      t\textquotesingle{} }\OtherTok{\textless{}{-}}\NormalTok{ conType x t}
-\NormalTok{      compareElims t\textquotesingle{} (}\DataTypeTok{Con}\NormalTok{ x ci []) xArgs yArgs}
-  \OperatorTok{...}
-\end{Highlighting}
-\end{Shaded}
-
-The functions described above are specific to Agda but in other major
-languages we can find similar problems with unifiers being large modules
-that are hard to understand. The sizes of modules with unifiers are as
-follows: Idris (1.5kloc\footnote{\href{https://github.com/idris-lang/Idris2/blob/102d7ebc18a9e881021ed4b05186cccda5274cbe/src/Core/Unify.idr}{./src/Core/Unify.idr}}),
-Lean (1.8kloc\footnote{\href{https://github.com/leanprover/lean4/blob/75252d2b85df8cb9231020a556a70f6d736e7ee5/src/Lean/Meta/ExprDefEq.lean}{./src/Lean/Meta/ExprDefEq.lean}}),
-Coq (1.8kloc\footnote{\href{https://github.com/coq/coq/blob/b35c06c3ab3ed4911311b4a9428a749658d3eff1/pretyping/evarconv.mli}{./pretyping/evarconv.mli}
-  at
-  \href{github.com/coq/coq/blob/b35c06c3ab3ed4911311b4a9428a749658d3eff1/}{github.com/coq/coq/blob/b35c06}}).
-For Haskell, which is not a dependently typed language yet, but does
-have a constraints system
-\citep{peytonjonesTypeInferenceConstraint2019}, this number is at
-2kloc\footnote{\href{https://gitlab.haskell.org/ghc/ghc/-/blob/b81cd709df8054b8b98ac05d3b9affcee9a8b840/compiler/GHC/Core/Unify.hs}{./compiler/GHC/Core/Unify.hs}
-  at
-  \href{https://gitlab.haskell.org/ghc/ghc/-/blob/b81cd709df8054b8b98ac05d3b9affcee9a8b840}{gitlab.haskell.org/ghc/ghc/-/blob/b81cd709d}}.
-
-\hypertarget{sec:implicit-arguments}{%
-\subsection{Type-checking function application in presence of implicit
-arguments}\label{sec:implicit-arguments}}
-
-During the type-checking of function application there may be different
-kinds of implicit arguments to infer, for example, instance arguments,
-or tactic arguments. If we start from a simple case of type-checking an
-application of a function symbol to regular arguments, every next
-extension needs to be handled as a special case.
-
-Take Agda -- when checking an application of implicit
-arguments\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Implicit.hs\#L96-L128}{./src/full/Agda/TypeChecking/Implicit.hs\#L96-L128}}
-we already have to carry the information on how the argument will be
-resolved and then create a specific kind of metavariable\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Implicit.hs\#L130-L149}{./src/full/Agda/TypeChecking/Implicit.hs\#L130-L149}}
-\citep[chap.~3]{norellPracticalProgrammingLanguage2007} for each of
-those cases.
-
-For handling \texttt{auto} variables, Idris 2
-\citep[chap.~13.1]{theidristeamIdrisTutorial2021} has to essentially
-inline the search procedure through a chain of elaboration function
-calls (\texttt{checkApp} to \texttt{checkAppWith} to
-\texttt{checkAppWith\textquotesingle{}}) to
-\texttt{makeAutoImplicit}\footnote{\href{https://github.com/idris-lang/Idris2/blob/870bc824371d504a03af937f326216302210a875/src/TTImp/Elab/App.idr\#L224-L241}{./src/TTImp/Elab/App.idr\#L224-L241}}.
-This can accommodate interfaces (or type classes), but one can imagine
-that if a different kind of implicit was added, like tactic arguments or
-Canonical Structures, we would have to inline the search again,
-requiring a non-trivial modification to the elaboration mechanism.
-
-While the codebases above show that it is certainly possible to extend
-languages with new features if the language was not written with
-extensibility in mind, this can lead to rather ad hoc solutions.
-
-\hypertarget{sec:extending-unification}{%
-\subsection{Extending unification}\label{sec:extending-unification}}
-
-While writing a unifier is hard enough as it is, at times the developers
-might want to give their users ability to extend the unification
-procedure.
-
-Canonical Structures
-\citep{saibiOutilsGeneriquesModelisation1999, mahboubiCanonicalStructuresWorking2013}
-was already mentioned as it is in the overlap between type classes and
-unification hints. Adding it to a language that does not support them
-requires an extension of the unification algorithm with a rule that says
-that projection from a canonical structure is an injective function
-\citep[eq. 1]{mahboubiCanonicalStructuresWorking2013}.
-
-One could also provide means to do so manually in a more general case,
-by allowing users to declare certain symbols as injective. This is one
-of the features requested by the Agda users
-\citep{agdausersInjectiveUnificationPragma2023}.
-
-Another example of this can be adding rules of associativity and
-commutativity to the unifier, as described in the thesis by
-\citet{holtenDependentTypeCheckingModulo2023}. It required 2000 lines of
-code\footnote{\href{https://github.com/LHolten/agda-commassoc/tree/defenitional-commutativity}{github.com/LHolten/agda-commassoc/tree/defenitional-commutativity}}.
-We would like to make changes such as this more feasible.
-
-\hypertarget{sec:summary-of-the-issues}{%
-\subsection{Summary of the issues}\label{sec:summary-of-the-issues}}
-
-The examples above show that when building a dependently typed language
-the core might be perfectly elegant and simple, but the features that
-appear on top of it complicate the design.
-
-One can also observe that while the code in existing implementations
-might rely on constraints, the design at large does not put them at the
-centre of the picture and instead views them primarily as a gadget.
-Agda's constraint solver\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Constraints.hs\#L247-L298}{src/full/Agda/TypeChecking/Constraints.hs\#L247-L298}}
-relies on the type-checker to call it at the point where it is needed
-and has to be carefully engineered to work with the rest of the
-codebase.
-
-\hypertarget{sec:what-is-our-design-bringing-into-the-picture}{%
-\section{A blueprint for extensible
-elaborators}\label{sec:what-is-our-design-bringing-into-the-picture}}
-
-Our idea for a new design is to shift focus more towards the constraints
-themselves:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  We give a stable API for raising constraints that can be called by the
-  type-checker, essentially creating an ``ask'' to be fulfilled by the
-  solvers. This is not dissimilar to the idea of mapping object-language
-  unification variables to host-language ones as done by
-  \citet{guidiImplementingTypeTheory2017}, view of the ``asks'' as a
-  general effect \citep[chap.~4.4]{bauerEqualityCheckingGeneral2020}, or
-  communication between actors \citep{allaisTypOSOperatingSystem2022a}.
-\item
-  To make the language more modular, we make constraints an extensible
-  data type in the style of \citet{swierstraDataTypesCarte2008} and give
-  an API to define new solvers with the ability to specify what kinds of
-  constraints they can solve.
-\end{enumerate}
-
-In the examples in this paper, we follow the bidirectional style of
-type-checking, but in practice, the design decisions are agnostic of the
-underlying system, as long as it adheres to the principle of stating the
-requirements on terms in terms of raising a constraint and not by, say,
-pattern-matching on a concrete term representation.
-
-From a birds-eye view the architecture looks as depicted in Figure
-\ref{architecture-figure}. The type-checking begins by initializing
-state and doing the syntax traversal. The traversal raises the
-constraints, and for the moment, the constraints are simply stored. As
-soon as we finish the traversal of some block (one declaration in our
-case), the solver dispatcher is called. It goes over the set of
-constraints, and for each active constraint calls different plugins to
-make an attempt at solving it. Each plugin, whether user-supplied
-(\texttt{Plugin\ A}) or provided by us (\texttt{unification}) consist of
-a handler and a solver. The handler determines if the plugin can
-potentially solve a constraint, if so the dispatcher runs the
-corresponding solver. All components have some read access to the state,
-including handlers which might for example verify that there are no
-extra constraints on the metavariable. For the write access: syntax
-traversal writes new metavariables to the state and elaborated
-definitions; the solver dispatcher writes updated meta-information;
-solvers write solutions to the metavariables and can raise new
-constraints.
-
-\begin{figure*}
-  \center
-
-  \includegraphics[width=\textwidth]{architecture-diagram.png}
-
-  \caption{Architecture diagram}
-  \label{architecture-figure}
-\end{figure*}
-
-For the moment we need to recompile the project in order to include new
-plugins, but this is not necessary and a system that dynamically loads
-plugins is possible to implement in a way that is similar to GHC
-Plugins\footnote{\href{https://mpickering.github.io/plugins.html}{mpickering.github.io/plugins.html}}
-or Accelerate \footnote{\href{https://github.com/tmcdonell/accelerate-llvm/blob/master/accelerate-llvm-native/src/Data/Array/Accelerate/LLVM/Native/Link/Runtime.hs\#L40}{github.com/tmcdonell/accelerate-llvm/blob/master/accelerate-llvm-native/src/Data/Array/Accelerate/LLVM/Native/Link/Runtime.hs}}
-\citep{mcdonellTypesafeRuntimeCode2015}.
-
-\hypertarget{sec:bidirectional}{%
-\section{Dependently typed calculus and bidirectional
-typing}\label{sec:bidirectional}}
-
-In this section, we describe the core of the type system we implement.
-We take pi-forall \citep{weirichImplementingDependentTypes2022} as a
-basis for the system and extend it with metavariables in core syntax and
-implicit arguments in the surface syntax. However, for all other
-purposes, we leave the core rules intact and therefore, the core
-calculus too.
-
-\hypertarget{basic-language-and-rules}{%
-\subsection{Basic language and rules}\label{basic-language-and-rules}}
-
-This is a dependently typed calculus that includes Pi, Sigma and indexed
-inductive types.
-
-Here is the internal syntax term data type, apart from service
-constructor omissions, like a placeholder \texttt{TRUSTME}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{data} \DataTypeTok{Term} \OtherTok{=}
-  \CommentTok{{-}{-} type of types Type}
-    \DataTypeTok{Type}
-  \CommentTok{{-}{-} variables x}
-  \OperatorTok{|} \DataTypeTok{Var} \DataTypeTok{TName}
-  \CommentTok{{-}{-} abstraction \textbackslash{}x. a}
-  \OperatorTok{|} \DataTypeTok{Lam}\NormalTok{ (}\DataTypeTok{Bind} \DataTypeTok{TName} \DataTypeTok{Term}\NormalTok{)}
-  \CommentTok{{-}{-} application a b}
-  \OperatorTok{|} \DataTypeTok{App} \DataTypeTok{Term} \DataTypeTok{Arg}
-  \CommentTok{{-}{-} function type (x : A) {-}\textgreater{} B}
-  \OperatorTok{|} \DataTypeTok{Pi} \DataTypeTok{Epsilon} \DataTypeTok{Type}\NormalTok{ (}\DataTypeTok{Bind} \DataTypeTok{TName} \DataTypeTok{Type}\NormalTok{)}
-  \CommentTok{{-}{-} Sigma{-}type \{ x : A | B \}}
-  \OperatorTok{|} \DataTypeTok{Sigma} \DataTypeTok{Term}\NormalTok{ (}\DataTypeTok{Bind} \DataTypeTok{TName} \DataTypeTok{Term}\NormalTok{)}
-  \OperatorTok{|} \DataTypeTok{Prod} \DataTypeTok{Term} \DataTypeTok{Term}
-  \OperatorTok{|} \DataTypeTok{LetPair} \DataTypeTok{Term}\NormalTok{ (}\DataTypeTok{Bind}\NormalTok{ (}\DataTypeTok{TName}\NormalTok{, }\DataTypeTok{TName}\NormalTok{) }\DataTypeTok{Term}\NormalTok{)}
-  \CommentTok{{-}{-} Equality type  a = b}
-  \OperatorTok{|} \DataTypeTok{TyEq} \DataTypeTok{Term} \DataTypeTok{Term}
-  \OperatorTok{|} \DataTypeTok{Refl}
-  \OperatorTok{|} \DataTypeTok{Subst} \DataTypeTok{Term} \DataTypeTok{Term}
-  \OperatorTok{|} \DataTypeTok{Contra} \DataTypeTok{Term}
-  \CommentTok{{-}{-} inductive datatypes}
-  \OperatorTok{|} \DataTypeTok{TCon} \DataTypeTok{TCName}\NormalTok{ [}\DataTypeTok{Arg}\NormalTok{] }\CommentTok{{-}{-} types (fully applied)}
-  \OperatorTok{|} \DataTypeTok{DCon} \DataTypeTok{DCName}\NormalTok{ [}\DataTypeTok{Arg}\NormalTok{] }\CommentTok{{-}{-} terms (fully applied)}
-  \OperatorTok{|} \DataTypeTok{Case} \DataTypeTok{Term}\NormalTok{ [}\DataTypeTok{Match}\NormalTok{]}
-    \CommentTok{{-}{-} metavariables}
-  \OperatorTok{|} \DataTypeTok{MetaVar} \DataTypeTok{MetaClosure}
-\end{Highlighting}
-\end{Shaded}
-
-Equality is not defined as a regular inductive type, but is instead
-built-in. The user has access to the type and term constructor, but not
-the ability to pattern-matching on it. Instead the language provides a
-\texttt{subst} primitive of type
-\texttt{(A\ x)\ -\textgreater{}\ (x=y)\ -\textgreater{}\ A\ y} and
-\texttt{contra} that takes an equality of two different inductive type
-constructors and produces an element of any type.
-
-On top of the above the language includes indexed inductive datatypes
-and case-constructs for their elimination. Indexed inductive datatypes
-are encoded as parameterised datatypes with an equality argument
-constraining the index.
-
-As for metavariables \texttt{MetaVar}: as mentioned in the introduction,
-they are placeholders in the syntax tree (AST) that are produced in the
-process of elaboration. Metavariables do not appear in the surface
-syntax as they are not created by the user. In this paper we implement
-metavariables in the contextual style, as described by
-\citet{abelHigherOrderDynamicPattern2011}, therefore they have to carry
-around a closure of type \texttt{MetaClosure}.
-
-\hypertarget{syntax-traversal}{%
-\subsection{Syntax traversal}\label{syntax-traversal}}
-
-We implement the core of the elaborator as a bidirectional syntax
-traversal, raising a constraint every time we need to assert something
-about the type.
-
-This includes the expected use of unification constraints, like in case
-we enter checking mode while the term should be inferred:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{checkType tm ty }\OtherTok{=} \KeywordTok{do}
-\NormalTok{  (etm, ty\textquotesingle{}) }\OtherTok{\textless{}{-}}\NormalTok{ inferType tm}
-\NormalTok{  constrainEquality ty\textquotesingle{} ty }\DataTypeTok{I.Type}
-  \FunctionTok{return}\NormalTok{ etm}
-\end{Highlighting}
-\end{Shaded}
-
-Any time we want to decompose the type provided in checking mode we pose
-a constraint that guarantees the type being in a specific shape:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{checkType (}\DataTypeTok{S.Lam}\NormalTok{ lam) ty }\OtherTok{=} \KeywordTok{do}
-\NormalTok{  mtyA }\OtherTok{\textless{}{-}}\NormalTok{ createMetaTerm}
-\NormalTok{  mtx }\OtherTok{\textless{}{-}}\NormalTok{ createUnknownVar}
-\NormalTok{  mtyB }\OtherTok{\textless{}{-}}\NormalTok{ extendCtx (}\DataTypeTok{I.TypeSig}\NormalTok{ (}\DataTypeTok{I.Sig}\NormalTok{ mtx mtyA))}
-\NormalTok{                    (createMetaTerm)}
-  \KeywordTok{let}\NormalTok{ mbnd }\OtherTok{=}\NormalTok{ bind mtx mtyB}
-  \KeywordTok{let}\NormalTok{ metaPi }\OtherTok{=} \DataTypeTok{I.Pi}\NormalTok{ mtyA mbnd}
-
-\NormalTok{  constrainEquality ty metaPi }\DataTypeTok{I.Type}
-  \CommentTok{{-}{-} rest of the traversal can now use mtyA and mbnd}
-  \OperatorTok{...}
-\end{Highlighting}
-\end{Shaded}
-
-At certain points we have to raise a constraint which has an associated
-continuation. Like for checking the type of a data constructor -- the
-part of the program that comes as an argument to
-\texttt{constrainTConAndFreeze} will be suspended (or ``blocked'') until
-the meta is solved with something of shape \texttt{TCon\ \_\ \_}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{checkType t}\OperatorTok{@}\NormalTok{(}\DataTypeTok{S.DCon}\NormalTok{ c args) ty }\OtherTok{=} \KeywordTok{do}
-\NormalTok{  elabpromise }\OtherTok{\textless{}{-}}\NormalTok{ createMetaTerm}
-\NormalTok{  constrainTConAndFreeze ty }\OperatorTok{$} \KeywordTok{do}
-\NormalTok{    mty }\OtherTok{\textless{}{-}}\NormalTok{ substMetas ty}
-    \KeywordTok{case}\NormalTok{ mty }\KeywordTok{of}
-\NormalTok{      (}\DataTypeTok{I.TCon}\NormalTok{ tname params) }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
-        \OperatorTok{...}
-\NormalTok{      \_ }\OtherTok{{-}\textgreater{}} \FunctionTok{error} \StringTok{"impossible"}
-\end{Highlighting}
-\end{Shaded}
-
-We add one more case to the elaborator for implicit arguments.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{checkType (}\DataTypeTok{Implicit}\NormalTok{) ty }\OtherTok{=} \KeywordTok{do}
-\NormalTok{  m }\OtherTok{\textless{}{-}}\NormalTok{ createMetaTerm}
-\NormalTok{  raiseConstraint }\OperatorTok{$} \DataTypeTok{FillInTheMeta}\NormalTok{ m ty}
-  \FunctionTok{return}\NormalTok{ m}
-\end{Highlighting}
-\end{Shaded}
-
-This is used for implicit arguments as will be described in more detail
-in Section \ref{sec:case-implicits}. The syntax traversal does not need
-to know anything at all about the implicits. The only thing we require
-is that the elaboration of the argument is called with the type
-information available. This corresponds to how in bidirectional typing
-function application is done in the inference mode but the arguments are
-processed in checking mode.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{inferType (}\DataTypeTok{App}\NormalTok{ t1 t2) }\OtherTok{=} \KeywordTok{do}
-\NormalTok{  (et1, }\DataTypeTok{Pi}\NormalTok{ tyA tyB) }\OtherTok{\textless{}{-}}\NormalTok{ inferType t1}
-\NormalTok{  et2 }\OtherTok{\textless{}{-}}\NormalTok{ checkType t2 tyA}
-  \FunctionTok{return}\NormalTok{ (}\DataTypeTok{App}\NormalTok{ et1 et2, subst tyB et2)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{sec:constraints_and_unification}{%
-\section{Constraints and
-unification}\label{sec:constraints_and_unification}}
-
-In Section \ref{sec:bidirectional} we described the syntax traversal
-part of the elaborator, which generates the constraints. In this section
-we will go over the constraint datatype needed for the base language,
-how unification is implemented and how we can extend the unification
-procedure.
-
-\hypertarget{base-constraints}{%
-\subsection{Base constraints}\label{base-constraints}}
-
-The datatype of constraints is open which means the user can write a
-plugin to extend it. However, we offer a few constraints out of the box
-to be able to type-check the base language.
-
-For the purposes of the base language it suffices to have the following.
-
-\begin{itemize}
-\item
-  A constraint that enforces equality of two terms of a given type
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{{-}{-} two terms given should be equal}
-\KeywordTok{data} \DataTypeTok{EqualityConstraint}\NormalTok{ e }\OtherTok{=}
-     \DataTypeTok{EqualityConstraint} \DataTypeTok{Syntax.Term} \DataTypeTok{Syntax.Term}
-                        \DataTypeTok{Syntax.Type}
-                        \DataTypeTok{Sytax.MetaVarId}
-\end{Highlighting}
-\end{Shaded}
-\item
-  A constraint that ensures that a metavariable is resolved eventually:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{{-}{-} this terms has to be filled in}
-\KeywordTok{data} \DataTypeTok{FillInTheTerm}\NormalTok{ e }\OtherTok{=}
-     \DataTypeTok{FillInTheTerm} \DataTypeTok{Syntax.Term}
-\NormalTok{                   (}\DataTypeTok{Maybe} \DataTypeTok{Syntax.Type}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-\item
-  Lastly, a constraint which ensures that a term is a type constructor:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{{-}{-} the term passed to the constraint}
-\CommentTok{{-}{-} should be a type constructor}
-\KeywordTok{data} \DataTypeTok{TypeConstructorConstraint}\NormalTok{ e }\OtherTok{=}
-     \DataTypeTok{TypeConstructorConstraint} \DataTypeTok{Syntax.Type}
-\end{Highlighting}
-\end{Shaded}
-\end{itemize}
-
-The type-checker raises them supplying the information necessary, but
-agnostic of how they will be solved.
-
-\hypertarget{sec:solvers-interface}{%
-\subsection{Interface of the solvers}\label{sec:solvers-interface}}
-
-On the solver side we provide a suite\footnote{In the prototype we
-  implement a subset of all unification rules, here are they listed:
-  identityPlugin, propagateMetasEqPlugin, reduceLeftPlugin,
-  reduceRightPlugin, leftMetaPlugin, rightMetaPlugin,
-  typeConstructorPlugin, typeConstructorWithMetasPlugin,
-  piEqInjectivityPlugin, tyEqInjectivityPlugin, consInjectivityPlugin,
-  typeInjectivityPlugin, unificationStartMarker, unificationEndMarker}
-of solvers for unification that handle different cases of the problem.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{{-}{-} solves syntactically equal terms}
-\OtherTok{syntacticHandler ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ c)}
-                 \OtherTok{=\textgreater{}} \DataTypeTok{HandlerType}\NormalTok{ c}
-\OtherTok{syntacticSolver ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ c)}
-                \OtherTok{=\textgreater{}} \DataTypeTok{SolverType} \DataTypeTok{Bool}
-\OtherTok{syntactic ::} \DataTypeTok{Plugin}
-\NormalTok{syntactic  }\OtherTok{=} \DataTypeTok{Plugin}\NormalTok{ \{ solver  }\OtherTok{=}\NormalTok{ syntacticSolver}
-\NormalTok{                    , handler }\OtherTok{=}\NormalTok{ syntacticHandler}
-                    \OperatorTok{...}
-\NormalTok{                    \}}
-\end{Highlighting}
-\end{Shaded}
-
-We first define the class of constraints that will be handled by the
-solver via providing a ``handler'' -- function that decides whether a
-given solver has to fire.\footnote{The code shown above and in the rest
-  of the paper is close to the actual implementation, but has been
-  simplified for presentation purposes.} In this case, this amounts to
-checking that the constraint given is indeed an
-\texttt{EqualityConstraint} and that the two terms given to it are
-syntactically equal. Then we define the solver itself, which in this
-case does not have to do anything except \texttt{return\ True} to
-indicate that the constraint is solved, as we it only fires once it has
-been cleared to do so by the handler and the equality has already been
-checked. Finally, we register the solver by declaring it using a plugin
-interface.
-
-The reason for this separation between a decision procedure and the
-execution of solver is to ensure separation between pontentially slow,
-effectful solving and fast read-only decision-making in the handler. We
-opt for this division since handlers will be run on many constraints
-that do not fit, therefore any write effects would have to be rolled
-back, while solvers should be fired only in cases when we can reasonably
-hope that the constraint will be solved and the effects will not have to
-be rolled back.
-
-Similarly, we can define \texttt{leftMetaSolver} and
-\texttt{rightMetaSolver} that only work on problems where only one of
-the sides is a metavariable. Here the job of the solver is not as
-trivial -- it has to check that the type of the other side indeed
-matches the needed one and then register the instantiation of the
-metavariable in the state.
-
-Since the constraints they match on overlap we can provide priority
-preferences, using the \texttt{pre} and \texttt{suc} fields of the
-plugin interface. They are used to indicate whether the currently
-defined plugin should run before or after, respectively, which other
-plugins. At the time of running the compiler, these preferences are
-loaded into a big pre-order relation for all the plugins, which is then
-linearised and used to guide the solving procedure.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\OtherTok{rightMetaPlugin ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
-                \OtherTok{=\textgreater{}} \DataTypeTok{Plugin}\NormalTok{ cs}
-\NormalTok{rightMetaPlugin }\OtherTok{=}
-  \DataTypeTok{Plugin}\NormalTok{ \{ handler }\OtherTok{=}\NormalTok{ rightMetaHandler}
-\NormalTok{         , solver  }\OtherTok{=}\NormalTok{ rightMetaSolver}
-\NormalTok{         , symbol  }\OtherTok{=}\NormalTok{ rightMetaSymbol}
-\NormalTok{         , pre }\OtherTok{=}\NormalTok{ []}
-\NormalTok{         , suc }\OtherTok{=}\NormalTok{ [leftMetaSymbol]}
-\NormalTok{         \}}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{sec:solvers-implementation}{%
-\subsection{Implementation of the solvers and unification
-details}\label{sec:solvers-implementation}}
-
-We implement a system close to the one described by
-\citet{abelHigherOrderDynamicPattern2011}. We modularise the
-implementation by mapping every function call in the simplification
-procedure to a raised constraint and every simplification rule to a
-separate solver. For example, the ``decomposition of functions''
-\citep[fig.~2]{abelHigherOrderDynamicPattern2011} rule is translated to
-the following implementation.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\OtherTok{piEqInjectivityHandler ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
-                       \OtherTok{=\textgreater{}} \DataTypeTok{HandlerType}\NormalTok{ cs}
-\NormalTok{piEqInjectivityHandler constr }\OtherTok{=} \KeywordTok{do}
-  \KeywordTok{let}\NormalTok{ eqcm }\OtherTok{=}\NormalTok{ match }\OperatorTok{@}\DataTypeTok{EqualityConstraint}\NormalTok{ constr}
-  \KeywordTok{case}\NormalTok{ eqcm }\KeywordTok{of}
-    \DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{EqualityConstraint}\NormalTok{ pi1 pi2 \_ \_) }\OtherTok{{-}\textgreater{}}
-      \KeywordTok{case}\NormalTok{ (pi1, pi2) }\KeywordTok{of}
-\NormalTok{        (}\DataTypeTok{I.Pi}\NormalTok{ \_ \_ \_, }\DataTypeTok{I.Pi}\NormalTok{ \_ \_ \_) }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{True}
-\NormalTok{        \_ }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
-\NormalTok{    \_ }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
-\end{Highlighting}
-\end{Shaded}
-
-The handler is simply checking that both sides of the equality are
-indeed Pi-types, and in case either of the matches fails, failure will
-be reported and the solver will not be fired. The \texttt{match}
-function above comes from the open datatype of constraints, checking if
-\texttt{constr} can be projected from \texttt{cs} to
-\texttt{EqualityConstraint}.
-
-Now let us take a look at the solver.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\OtherTok{piEqInjectivitySolver ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
-                      \OtherTok{=\textgreater{}} \DataTypeTok{SolverType}\NormalTok{ cs}
-\NormalTok{piEqInjectivitySolver constr }\OtherTok{=} \KeywordTok{do}
-  \KeywordTok{let}\NormalTok{ (}\DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{EqualityConstraint}\NormalTok{ (}\DataTypeTok{I.Pi}\NormalTok{ a1 b1)}
-\NormalTok{                                (}\DataTypeTok{I.Pi}\NormalTok{ a2 b2) \_ m)) }\OtherTok{=}
-\NormalTok{        match }\OperatorTok{@}\DataTypeTok{EqualityConstraint}\NormalTok{ constr}
-\NormalTok{  ma }\OtherTok{\textless{}{-}}\NormalTok{ constrainEquality a1 a2 }\DataTypeTok{I.Type}
-\NormalTok{  (x, tyB1) }\OtherTok{\textless{}{-}}\NormalTok{ Unbound.unbind b1}
-\NormalTok{  (y, tyB2\textquotesingle{}) }\OtherTok{\textless{}{-}}\NormalTok{ Unbound.unbind b2}
-  \KeywordTok{let}\NormalTok{ tyB2 }\OtherTok{=}\NormalTok{ Unbound.subst y (}\DataTypeTok{I.Var}\NormalTok{ x) tyB2\textquotesingle{}}
-\NormalTok{      mat }\OtherTok{=}\NormalTok{ I.identityClosure ma}
-\NormalTok{  mb }\OtherTok{\textless{}{-}}\NormalTok{ extendCtx (}\DataTypeTok{I.TypeSig}\NormalTok{ (}\DataTypeTok{I.Sig}\NormalTok{ x e1 mat)) }\OperatorTok{$}
-\NormalTok{                  constrainEquality tyB1 tyB2 }\DataTypeTok{I.Type}
-  \KeywordTok{let}\NormalTok{ mbt }\OtherTok{=}\NormalTok{ bind x }\OperatorTok{$}\NormalTok{ I.identityClosure mb}
-\NormalTok{  solveMeta m (}\DataTypeTok{I.Pi}\NormalTok{ mat mbt)}
-  \FunctionTok{return} \DataTypeTok{True}
-\end{Highlighting}
-\end{Shaded}
-
-Following pi-forall, we use the unbound-generics\footnote{\href{https://hackage.haskell.org/package/unbound-generics-0.4.3}{hackage.haskell.org/package/unbound-generics-0.4.3}}
-library to deal with the names and binders. As it will fire after a
-handler returns \texttt{True}, we can assume the pattern-matches will
-not fail.
-
-First we constrain the equality of the domain of the Pi-type:
-\texttt{a1} and \texttt{a2}. The seemingly spurious metavariable
-\texttt{ma} returned from this call serves as an anti-unification
-\citep{pfenningUnificationAntiunificationCalculus1991} communication
-channel. Every time an equality constraint is created we return a
-metavariable that stands for the unified term. This metavariable is used
-for unification problems that are created in the extended contexts -- in
-this case second argument of the Pi-type, but also when solving
-equalities concerning two data constructors. We do this to tackle the
-``spine problem''
-\citep[sec.~1.4]{victorlopezjuanPracticalHeterogeneousUnification2021}
--- as we operate according to the ``well-typed modulo constraints''
-principle, essentially providing a placeholder that is guaranteed to
-preserve well-typedness in the extended context. \footnote{\(\text{Tog}^{+}\)
-  \citep{victorlopezjuanTog2020, victorlopezjuanPracticalHeterogeneousUnification2021}
-  focuses on extending unification algorithm for the case where two
-  sides of equality might not be of the same type, which is also a
-  problem relevant for us. Their main argument against the usage of
-  anti-unification in Agda provided there is that it is bug-prone. We
-  think that in Agda the problems were stemming from the fact that
-  anti-unification were implemented separately from unification, in
-  which case it is indeed hard to keep the two in sync. There is no such
-  duplication in our case since unification and anti-unification are
-  one.} Finally \texttt{ma} has to be applied to a closure which will
-keep track of the delayed substitution.
-
-Then we can constrain the co-domains of Pi-types in an extended context.
-In case one of solvers the constraints created in the extended context
-might need to know the exact shape of \texttt{ma}, we can block on the
-metavariable later, freezing the rest of the problem until it is
-instantiated.
-
-As for the actual unification steps, we implement them in a similar
-fashion to simplification procedure. Take a look at the following
-example, when only the left hand side of the constaint is an unsolved
-meta:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\OtherTok{leftMetaSolver ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
-               \OtherTok{=\textgreater{}} \DataTypeTok{SolverType}\NormalTok{ cs}
-\NormalTok{leftMetaSolver constr }\OtherTok{=} \KeywordTok{do}
-  \KeywordTok{let}\NormalTok{ (}\DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{EqualityConstraint}\NormalTok{ t1 t2 \_ m)) }\OtherTok{=}
-\NormalTok{        match }\OperatorTok{@}\DataTypeTok{EqualityConstraint}\NormalTok{ constr}
-\NormalTok{      (}\DataTypeTok{MetaVar}\NormalTok{ (}\DataTypeTok{MetaVarClosure}\NormalTok{ m1 c1)) }\OtherTok{=}\NormalTok{ t1}
-\NormalTok{  mt2 }\OtherTok{\textless{}{-}}\NormalTok{ occursCheck m1 t2}
-  \KeywordTok{case}\NormalTok{ mt2 }\KeywordTok{of}
-    \CommentTok{{-}{-} indicates a failure in occurs{-}check}
-    \DataTypeTok{Left}\NormalTok{ e }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
-    \CommentTok{{-}{-} indicates a passed occurs{-}check}
-    \DataTypeTok{Right}\NormalTok{ t2 }\OtherTok{{-}\textgreater{}}
-      \KeywordTok{case}\NormalTok{ (invertClosureOn c1 (freeVarList t2)) }\KeywordTok{of}
-        \DataTypeTok{Just}\NormalTok{ s }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
-          \KeywordTok{let}\NormalTok{ st2 }\OtherTok{=}\NormalTok{ Unbound.substs s t2}
-          \CommentTok{{-}{-} apply the subsitution}
-\NormalTok{          solveMeta m1 st2}
-          \CommentTok{{-}{-} instantiate the anti{-}unification variable}
-\NormalTok{          solveMeta m st2}
-          \FunctionTok{return} \DataTypeTok{True}
-        \DataTypeTok{Nothing} \OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
-\end{Highlighting}
-\end{Shaded}
-
-Once the occurs-check returns and if it was successful we apply the
-inverted closure to the right-hand side of the equality.
-
-By splitting up the rules into individual, simple solvers we can
-compartmentalise the complexity of the unifier, making sure that each
-rule is as decoupled from the others as possible. This does not
-deteriorate the properties of the system, but does not help to guarantee
-them either. We talk more about the challenge of proving correctness in
-Section \ref{sec:limitations}.
-
-\hypertarget{extending-unification}{%
-\subsection{Extending unification}\label{extending-unification}}
-
-Now that unification is implemented let us create a simple plugin that
-makes certain symbols declared by the user injective
-\citep{agdausersInjectiveUnificationPragma2023}. The actual
-implementation is relatively simple and is not dissimilar to the
-Pi-injectivity solver we showed above.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\OtherTok{userInjectivitySolver ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
-                      \OtherTok{=\textgreater{}} \DataTypeTok{SolverType}\NormalTok{ cs}
-\NormalTok{userInjectivitySolver constr }\OtherTok{=} \KeywordTok{do}
-  \KeywordTok{let}\NormalTok{ (}\DataTypeTok{Just} \DataTypeTok{EqualityConstraint}
-\NormalTok{              (}\DataTypeTok{I.App}\NormalTok{ (}\DataTypeTok{I.Var}\NormalTok{ f) a)}
-\NormalTok{              (}\DataTypeTok{I.App}\NormalTok{ (}\DataTypeTok{I.Var}\NormalTok{ g) b) \_ m)) }\OtherTok{=}
-\NormalTok{        match }\OperatorTok{@}\DataTypeTok{EqualityConstraint}\NormalTok{ constr}
-  \KeywordTok{if}\NormalTok{ f }\OperatorTok{==}\NormalTok{ g}
-  \KeywordTok{then} \KeywordTok{do}
-\NormalTok{    ifM (queryInjectiveDeclarations f)}
-\NormalTok{        (}\KeywordTok{do}
-\NormalTok{           ms }\OtherTok{\textless{}{-}}\NormalTok{ constrainEquality a b }\DataTypeTok{I.Type}
-\NormalTok{           solveMeta m ms}
-           \FunctionTok{return} \DataTypeTok{True}\NormalTok{)}
-\NormalTok{        (}\FunctionTok{return} \DataTypeTok{False}\NormalTok{)}
-  \KeywordTok{else} \FunctionTok{return} \DataTypeTok{False}
-\end{Highlighting}
-\end{Shaded}
-
-where \texttt{queryInjectiveDeclarations} simply scans the available
-declarations for a marker that \texttt{f} has been declared injective.
-
-The only big thing left is to make sure that this solver fires at the
-right time. This can only conflict with the ``decomposition of
-neutrals'' rule, so we indicate to the solver dispatcher that our plugin
-should run before the it:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\OtherTok{userInjectivityPlugin ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
-                      \OtherTok{=\textgreater{}} \DataTypeTok{Plugin}\NormalTok{ cs}
-\NormalTok{userInjectivityPlugin }\OtherTok{=}
-  \DataTypeTok{Plugin}\NormalTok{ \{ }\OperatorTok{...}
-\NormalTok{         , solver }\OtherTok{=}\NormalTok{ userInjectivitySolver}
-\NormalTok{         , symbol }\OtherTok{=} \DataTypeTok{PluginId} \OperatorTok{$} \StringTok{"userInjectivity"}
-\NormalTok{         , pre }\OtherTok{=}\NormalTok{ [ unifyNeutralsDecomposition}
-\NormalTok{                 , unificationEndMarkerSymbol]}
-\NormalTok{         , suc }\OtherTok{=}\NormalTok{ [unificationStartMarkerSymbol]}
-\NormalTok{         \}}
-\end{Highlighting}
-\end{Shaded}
-
-This modification does not alter the core of the language.
-
-\hypertarget{sec:casestudies}{%
-\section{Case studies}\label{sec:casestudies}}
-
-Once we implement basic elaboration and unification we can extend the
-language. This is where we make use of the fact that the constraints
-datatype is open.
-
-We saw before in Section \ref{sec:implicit-arguments} that conventional
-designs require separate handling of different kinds of implicit
-variables. To simplify the design we would like to uniformly dispatch a
-search for the solution, which would be handled by the a fitting solver.
-We can achieve this by communicating the kind of the meta through its
-type in the second argument of \texttt{FillInTheMeta\ m\ ty}. The
-solvers then match on the shape of the type of the metavariable and
-handle it in a case-specific manner: instance-search for type classes,
-tactic execution for a tactic argument, waiting for regular unification
-to solve the metavariable for regular implicit arguments.
-
-In this section we will describe the implementation details of regular
-implicit arguments (Section \ref{sec:case-implicits}), the
-implementation of type classes added on top of the implicit arguments
-(Section \ref{sec:case-typeclasses}). And finally (Section
-\ref{sec:coercion-tactics}) we sketch the implementation of coercive
-subtyping and tactic arguments.
-
-\hypertarget{sec:case-implicits}{%
-\subsection{Implicit arguments}\label{sec:case-implicits}}
-
-As we mentioned, the rule for \texttt{Implicit} had to be added to the
-syntax traversal part of the elaborator. In fact, we require not one but
-two modifications that lie outside of the solvers-constraints part of
-the system. The first one is, indeed, the addition of a separate case in
-the syntax traversal, however contained. The second one lies in the
-purely syntactical part of the compiler. We need the pre-processor to
-insert the placeholder terms in the surface syntax. Particularly, we
-need to desugar declarations of functions in the following way:
-
-For any declaration of a function \texttt{f} with some implicit argument
-\texttt{a}:
-
-\begin{verbatim}
-def f : {A : Type} -> {a : A} -> B a -> C
-\end{verbatim}
-
-Which should desugar to the following:
-
-\begin{verbatim}
-def f : (A : Implicit Type)
-     -> (a : Implicit (deImp A))
-     -> (b : B (deImp a)) -> C
-\end{verbatim}
-
-Then, for each function call \texttt{f\ b1} we need to insert the
-corresponding number of implicit arguments, transforming it to
-\texttt{f\ \_\ \_\ b1}.
-
-For simple implicits this suffices -- as soon as we have the
-placeholders in the surface syntax we create the constraints in the
-\texttt{Implicit} case of the syntax traversal. The only solver that is
-needed in this case is a trivial one that checks that the metavariable
-has been instantiated indeed. This is because a regular implicit should
-be instantiated by the unifier at some point later. This constraint
-simply serves as a guarantee that all implicits have been instantiated.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\OtherTok{fillInImplicitSymbol ::} \DataTypeTok{PluginId}
-
-\OtherTok{fillInImplicitHandler ::}\NormalTok{ (}\DataTypeTok{FillInImplicit} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
-                      \OtherTok{=\textgreater{}} \DataTypeTok{HandlerType}\NormalTok{ cs}
-\NormalTok{fillInImplicitHandler constr }\OtherTok{=} \KeywordTok{do}
-  \KeywordTok{let}\NormalTok{ ficm }\OtherTok{=}\NormalTok{ match }\OperatorTok{@}\DataTypeTok{FillInImplicit}\NormalTok{ constr}
-  \KeywordTok{case}\NormalTok{ ficm }\KeywordTok{of}
-    \DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{FillInImplicit}\NormalTok{ term ty) }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
-      \KeywordTok{case}\NormalTok{ term }\KeywordTok{of}
-        \DataTypeTok{MetaVar}\NormalTok{ (}\DataTypeTok{MetaVarClosure}\NormalTok{ mid \_) }\OtherTok{{-}\textgreater{}}
-\NormalTok{          isMetaSolved mid}
-\NormalTok{        \_ }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
-    \DataTypeTok{Nothing} \OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
-
-\OtherTok{fillInImplicitPlugin ::}\NormalTok{ (}\DataTypeTok{FillInImplicit} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
-                     \OtherTok{=\textgreater{}} \DataTypeTok{Plugin}\NormalTok{ cs}
-\NormalTok{fillInImplicitPlugin }\OtherTok{=} \DataTypeTok{Plugin}\NormalTok{ \{}
-\NormalTok{  solver }\OtherTok{=}\NormalTok{ fillInImplicitSolver,}
-\NormalTok{  handler }\OtherTok{=}\NormalTok{ fillInImplicitHandler,}
-\NormalTok{  symbol }\OtherTok{=}\NormalTok{ fillInImplicitSymbol,}
-\NormalTok{  suc }\OtherTok{=}\NormalTok{ [],}
-\NormalTok{  pre }\OtherTok{=}\NormalTok{ []}
-\NormalTok{  \}}
-\end{Highlighting}
-\end{Shaded}
-
-Additionally, there is a design choice to be made the implementation of
-\texttt{Implicit\ A} and \texttt{deImp}. One option is to turn them into
-a constructor and a projection of a record type, the other is to make
-them computationally equivalent to \texttt{id}. In the former case we
-need to manually unwrap them both in the pre-processor and during the
-constraint-solving, but since the head symbol is distinct we can
-guarantee that other solvers will not match on it, unless explicitly
-instructed to. In the latter case, one has to be cautious of the order
-in which the solvers are activated, particularly in the case of
-different search procedures, should they be implemented. However, in the
-example above it does not make a difference.
-
-\hypertarget{sec:case-typeclasses}{%
-\subsection{Type classes}\label{sec:case-typeclasses}}
-
-Next, let us implement a plugin for type classes. Same as for the
-implicit arguments in general, we rely here on a pre-processor to
-transform user-friendly syntax to simple declarations. We do not focus
-on this part, since such a pre-processor does a simple local
-transformation.
-
-Let us now go through an example of the elaboration process for a simple
-term. For typeclasses we need a few declarations, listed below in
-Agda-like syntax:
-
-\begin{verbatim}
-plus  :  {A : Type} -> {{PlusOperation A}}
-     -> (a : A) -> (b : A) -> A
-
--- a semigroup on booleans, addition is OR
-instance BoolPlus : PlusOperation Bool where
-  plus = orb
-\end{verbatim}
-
-And the exemplary term itself is:
-
-\begin{verbatim}
-m = plus True False
-\end{verbatim}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  First, the pre-processor eliminates the implicits and type class
-  arguments. We end with the following declarations:
-
-\begin{verbatim}
-plus : (impA : Implicit Type)
-    -> Instance PlusOperation (deImp impA)
-    -> (a : deImp impA) -> (b :  deImp impA)
-    -> deImp impA
-
-instanceBoolPlus : InstanceT PlusOperation Bool
-instanceBoolPlus =
-  InstanceC (TypeClassC (Plus orb))
-
-m = plus _ _ True False
-\end{verbatim}
-
-  Declaration \texttt{instance\ BoolPlus} turning into a constructor of
-  \texttt{InstanceC} is precisely the part we need the pre-processor to
-  do.
-\item
-  We go into the elaboration of \texttt{m} now. The elaborator applies
-  \texttt{inferType\ (App\ t1\ t2)} rule four times and
-  \texttt{checkType\ (Implicit)\ ty} twice on the two placeholders. The
-  output of the elaborator is
-
-\begin{verbatim}
-m = plus ?_1 ?_2 True False
-\end{verbatim}
-
-  And the state of the elaborator contains four more constraints:
-
-\begin{verbatim}
-C1: FillInTheTerm ?_1 (Implicit Type)
-C2: FillInTheTerm ?_2 (InstanceT PlusOperation
-                                 (deImp ?_1))
-C3: EqualityConstraint (deImp ?_1) Bool Type
-C4: EqualityConstraint (deImp ?_1) Bool Type
-\end{verbatim}
-
-  The first two correspond to implicit arguments. The latter two are
-  unification problems rendered into constraints.
-\item
-  Now we step into the constraint-solving world. First, the unifier
-  solves the latter two constraints, instantiating \texttt{?\_1} to
-  \texttt{Implicit\ Bool}. \texttt{C1} is then discarded as solved since
-  \texttt{?\_1} is already instantiated to \texttt{Implicit\ Bool}.
-  Next, the typeclass resolution launches a search for the instance of
-  type \texttt{Instance\ PlusOperation\ Bool}.
-\item
-  This is where the type class plugin can take over. It transforms
-  \texttt{C2:\ FillInTheTerm\ ?\_2\ (InstanceT\ PlusOperation\ Bool)} to
-  \texttt{C5:\ InstanceSearch\ PlusOperation\ Bool\ ?\_2}. \texttt{C5}
-  then gets matched by the search plugin for concrete instances, simply
-  weeding through available declarations, looking for something of the
-  shape \texttt{InstanceT\ PlusOperation\ Bool}. Such a declaration
-  exists indeed and we can instantiate \texttt{?\_2} to
-  \texttt{instanceBoolPlus}.
-\end{enumerate}
-
-Now let us take a look at the plugin for the constraint system. It is
-contained in a single file
-\href{https://github.com/liesnikov/extensible-elaborator/blob/elaborator-experiments/exel/src/Plugins/Typeclasses.hs}{\texttt{./exel/src/Plugins/Typeclasses.hs}}.
-In it we define a new constraint type \texttt{InstanceSearch}, a solver
-that transforms constraints of the shape
-\texttt{FillInTheType\ ?\ (InstanceT\ \_)} to the new constraint, and
-finally, a solver for instance search problems. The reason to transform
-the original constraint to the new type is to ensure that no other
-solver will make an attempt at this problem, therefore modulating
-interactions with other plugins.
-
-Finally, the implementation of search for concrete instances is quite
-simple:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{instanceConcreteSolver constr }\OtherTok{=} \KeywordTok{do}
-  \KeywordTok{let}\NormalTok{ (}\DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{InstanceSearch}\NormalTok{ tcn ty m)) }\OtherTok{=}
-\NormalTok{        match }\OperatorTok{@}\DataTypeTok{InstanceSearch}\NormalTok{ constr}
-\NormalTok{  alldecls }\OtherTok{\textless{}{-}}\NormalTok{ collectInstancesOf tcn }\OperatorTok{\textless{}$\textgreater{}}\NormalTok{ SA.getDecls}
-\NormalTok{  sty }\OtherTok{\textless{}{-}}\NormalTok{ SA.substAllMetas ty}
-  \KeywordTok{case}\NormalTok{ Map.lookup sty alldecls }\KeywordTok{of}
-    \DataTypeTok{Just}\NormalTok{ i }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
-\NormalTok{      SA.solveMeta m (}\DataTypeTok{I.Var}\NormalTok{ i)}
-      \FunctionTok{return} \DataTypeTok{True}
-    \DataTypeTok{Nothing} \OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
-\end{Highlighting}
-\end{Shaded}
-
-We conjecture that implementation of Canonical Structures
-\citep{mahboubiCanonicalStructuresWorking2013} would be relatively
-simple in such a system due to openness of both unification procedure
-and instance search.
-
-\hypertarget{sec:coercion-tactics}{%
-\subsection{Tactic arguments and coercions}\label{sec:coercion-tactics}}
-
-In a similar fashion to the transformation of \texttt{FillInImplicit}
-constraints to \texttt{InstanceSearch}, we could implement tactic
-arguments and coercive subtyping.
-
-The former would be quite similar to what we saw in the previous section
-\ref{sec:case-typeclasses}, except we would have to resolve
-\texttt{FillInImplicit} to \texttt{RunTactic} constraint or fill in
-directly. The question of actually running the tactics is independent of
-constraints, since it is essentially another type-checking action that
-we accomodate already for blockers.
-
-Coercive subtyping is of a slightly different nature. First, it would
-rely on a pre-processor the heaviest out of all of the examples
-described above, since naively one would insert a (potentially identity)
-coercion in each argument of the application and potentially around
-heads of applications and types of abstractions
-\citep{tassiBiDirectionalRefinementAlgorithm2012}. This incurs not only
-syntactic noise, but also potential performance penalty, which we
-describe further in Section \ref{sec:limitations}. The second challenge
-comes from the fact that unlike the above this feature would be
-anti-modular in a sense that we need to access several constraints at
-once, at least read-only. Consider the following example:
-
-\begin{verbatim}
-f : (A : Type) -> A -> A
-arg : ArgTy
-
-t : ExpTy
-t = f _ arg
-\end{verbatim}
-
-This declaration would desugar to something like
-
-\begin{verbatim}
-t = coerce _ (f _ (coerce _ arg))
-\end{verbatim}
-
-which gives raise to two constraints, where \texttt{Coercion\ A\ B}
-stands for the evidence of \texttt{A} being a subtype of \texttt{B}:
-
-\begin{verbatim}
-C1 : Coercion ArgTy ?1
-C3 : Coercion ?1 ExpTy
-\end{verbatim}
-
-Clearly, we can not solve the first constraint in isolation and need to
-consider all coercion constraints related to a particular type at the
-same time. While we can accommodate such solvers, due to solvers having
-write-access to the state, it becomes much harder to modulate
-interactions between different plugins.
-
-\hypertarget{sec:limitations}{%
-\section{Limitations}\label{sec:limitations}}
-
-While our design offers a lot of flexibility, it does not solve every
-problem. In this section we describe a few examples of extensions that
-do not quite fit in this framework and general limitations.
-
-\hypertarget{handling-of-meta-variables-outside-of-definition-sites}{%
-\subsection{Handling of meta-variables outside of definition
-sites}\label{handling-of-meta-variables-outside-of-definition-sites}}
-
-After we elaborate a definition there can still be unsolved
-metavariables in it. This presents us with a design choice. The first
-option is to freeze the metavariables and instantiate them per-use site,
-essentially allowing for an expression to have multiple types. The
-second option is to leave them up to be solved later, which might make
-elaboration less predictable since now the use sites can influence
-whether a particular definition type-checks. The third option is to
-report them as an error and exit immediately.
-
-In particular, the second option allows us to incorporate more involved
-inference algorithms into the system. For example, if we were to
-implement an erasure inference algorithm as described by
-\citet{tejiscakDependentlyTypedCalculus2020}, we would have to create
-metavariable annotations (described as ``evars'' in the paper) that can
-be instantiated beyond the definition site. The downside is that the
-defintions can change depending on their use and type-checking has to be
-done effectively for the whole program and not per-definition.
-
-\hypertarget{the-language-is-only-as-extensible-as-the-syntax-traversal-is}{%
-\subsection{The language is only as extensible, as the syntax traversal
-is}\label{the-language-is-only-as-extensible-as-the-syntax-traversal-is}}
-
-Extensibility via constraints allows for a flexible user-specified
-control flow as soon as we step into the constraints world. But control
-flow of the syntax traversal for the basic language elaborator is fixed
-by the basic language. For example, consider the following simplified
-lambda-function type-checking function\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Rules/Term.hs\#L430-L518}{./src/full/Agda/TypeChecking/Rules/Term.hs\#L430-L518}}
-from Agda:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{checkLambda\textquotesingle{} cmp b xps typ body target }\OtherTok{=} \KeywordTok{do}
-  \DataTypeTok{TelV}\NormalTok{ tel btyp }\OtherTok{\textless{}{-}}\NormalTok{ telViewUpTo numbinds target}
-  \KeywordTok{if}\NormalTok{ size tel }\OperatorTok{\textless{}}\NormalTok{ numbinds }\OperatorTok{||}\NormalTok{ numbinds }\OperatorTok{/=} \DecValTok{1}
-    \KeywordTok{then}\NormalTok{ dontUseTargetType}
-    \KeywordTok{else}\NormalTok{ useTargetType tel btyp}
-\end{Highlighting}
-\end{Shaded}
-
-Here Agda steps away from the bidirectional discipline and infers a
-(lambda) function if the target type is not fully known. If in our
-design the developer chooses to go only with a pure bidirectional style
-of type-checking inferred lambda functions would be impossible to
-emulate.
-
-The solution in this case is effectively to replicate what Agda is doing
-by implementing each type-checking rule in inference mode, essentially
-factoring out \texttt{dontUseTargetType} in Agda's code snippet above.
-
-Similarly, implementing something like commutativity and associativity
-unifier plugins can require modifications in the core, since we two
-terms that are equal during elaboration have to equal during core
-type-checking too.
-
-\hypertarget{lack-of-backtracking}{%
-\subsection{Lack of backtracking}\label{lack-of-backtracking}}
-
-We do not implement any backtracking in the solver dispatcher as it is
-now. This means that every step taken is committing, which can be a
-limitation in cases where one would like to have backtracking -- for
-example, Agda's instance arguments search (with
-\texttt{-\/-overlapping-instances} flag)
-\citep[chap.~3.18]{theagdateamAgdaUserManual2023a}, as well as type
-classes in Lean \citep{selsamTabledTypeclassResolution2020}.
-
-Backtracking in principle could be achieved by tracking changes to the
-state of the elaborator and the production graph for constraints, but
-such a system would be rather awkward. Alternatively, such a solver can
-be implemented within one solver, removing the extension points, but
-allowing arbitrary control-flow within the algorithm.
-
-\hypertarget{reliance-on-a-pre-processor}{%
-\subsection{Reliance on a
-pre-processor}\label{reliance-on-a-pre-processor}}
-
-This work crucially relies on a pre-processor of some kind, be it macro
-expansion or some other way to extend the parser with custom desugaring
-rules. In particular, in order to implement n-ary implicit arguments
-correctly and easily we need the pre-processor to expand them to the
-right arity, similar to
-Matita\citep[chap.~5]{tassiBiDirectionalRefinementAlgorithm2012} and
-others
-\citep{serranoQuickLookImpredicativity2020, kovacsElaborationFirstclassImplicit2020}.
-
-\hypertarget{eager-reduction-and-performance}{%
-\subsection{Eager reduction and
-performance}\label{eager-reduction-and-performance}}
-
-As in the pre-processing step we would have to insert a fair number of
-wrappers and un-wrappers for all implicits and even more for coercions,
-a performance regression would be expected due a lot of spurious
-computation steps.
-
-We expect this to be a major concern for coercive subtyping. As one way
-to mitigate we suggest a discipline with constraint solvers latching
-onto non-reduced types and terms in constraints. In that case, we can
-get around with a trick borrowed from Coq, where the wrappers and
-unwrappers are identity functions and \texttt{coerce\ f\ t} computes to
-\texttt{f\ t}. This also means that constraints can/have to match on
-unreduced types in the e.g.~\texttt{FillInTheTherm}. In fact, we already
-do this to an extent for a different reason -- since the calculus allows
-only fully applied type constructors, we have to wrap each type class
-constructor in a lambda-abstraction for it to appear as an argument to
-\texttt{TypeClassT\ typeClassName\ argType}. Or, concretely, we have to
-define and use
-\texttt{PlusOperation\textquotesingle{}\ =\ \textbackslash{}A\ .\ PlusOperation\ A}
-in place of \texttt{PlusOperation} in the elaboration example in Section
-\ref{sec:case-typeclasses}.
-
-\hypertarget{proving-correctness}{%
-\subsection{Proving correctness}\label{proving-correctness}}
-
-As soon as we allow the users to implement their own solvers, there is
-little we can say about correctness of the system as a whole without
-imposing proof obligations on the plugin writers. However, if one is
-willing to do that we can conjecture solution and type preservation --
-Theorems 2 and 3, respectively \citep{abelHigherOrderDynamicPattern2011}
--- assuming that each solver on its own satisfies these properties. For
-termination it is similar -- we can conjecture that if every solver in
-the system ``reduces the weight'' of the unification problem, in terms
-of Theorem 1 by \citet{abelHigherOrderDynamicPattern2011}, we can hope
-for termination guarantees. This is simply a consequence of the fact
-that we do not invent a new unification algorithm, but rather provide
-means of easier implementation for it.
-
-\hypertarget{sec:related_work}{%
-\section{Related work}\label{sec:related_work}}
-
-We are certainly not the first ones to try to tackle the extensibility
-of a language implementation. Dependently typed language implementations
-usually consist of at least four parts: parser, elaborator, core
-type-checker, and proper back-end. The back-end part is currently
-irrelevant to our interests, since for a language to be specified
-usually means for specification of the core, anything that happens after
-the core does not extend the language, but rather tries to preserve its
-semantics in some form. Therefore we're left with three parts: parser,
-elaborator, and core type-checker.
-
-We see parser or syntax extensibility as a necessary part of an
-extensible language. This problem has been studied extensively in the
-past and has a multitude of existing solutions. Macros are one of them
-and are utilised heavily in various forms in almost all established
-languages
-\citep{thecoqdevelopmentteamCoqProofAssistant2022, theagdateamAgdaUserManual2023a, ullrichNotationsHygienicMacro2020}
-and can be powerful enough to build a whole language around
-\citep{changDependentTypeSystems2019}.
-
-Core extensibility, on the other hand, appears to be a problem with too
-many degrees of freedom. Andromeda
-\citep{bauerDesignImplementationAndromeda2018, bauerEqualityCheckingGeneral2020}
-made an attempt at extensible definitional equality but is quite far
-from a usable dependently typed language. Agda's philosophy allows
-developers to experiment with the core but also results in a larger
-amount of unexpected behaviours. In general, modifications of the core
-rules will result in fundamental changes in the type theory, which can
-break plenty of important properties like soundness or subject
-reduction.
-
-This leaves us with the question about the extensibility of an
-elaborator. We will make a division here between syntax traversals,
-constraint solving and all other features. The syntax traversal part of
-the elaborator is relatively stable and commonly implemented following a
-bidirectional discipline to some extent
-\citep{norellPracticalProgrammingLanguage2007, tassiBiDirectionalRefinementAlgorithm2012, ferreiraBidirectionalElaborationDependently2014},
-so there seems little reason to make it extensible.
-
-GHC has a plugin system that allows users to dynamically add custom
-constraint solvers, but the type of constraints itself is not
-extensible\footnote{\href{https://gitlab.haskell.org/ghc/ghc/-/wikis/plugins/type-checker/notes}{gitlab.haskell.org/ghc/ghc/-/wikis/plugins/type-checker/notes}}
-\citep{peytonjonesTypeInferenceConstraint2019, vytiniotisOutsideInModularType2011, peytonjonesPracticalTypeInference2007}.
-
-Coq \citep{thecoqdevelopmentteamCoqProofAssistant2022}, being one of the
-most popular proof assistants, invested a lot effort into user-facing
-features: work on tactics like a new tactic engine
-\citep{spiwackVerifiedComputingHomological2011} and tactic languages
-(Ltac2 \citep{pedrotLtac2TacticalWarfare2019}, SSReflect
-\citep{gonthierSmallScaleReflection2008}, etc.), the introduction of a
-virtual machine for performance
-\citep{gregoireCompiledImplementationStrong2002} and others. However,
-the implementation is quite hard to extend. One either has to modify the
-source code, which is mostly limited to the core development team, as
-seen from the
-\href{https://github.com/coq/coq/graphs/contributors}{graph}. Or one has
-to use Coq plugins system, which is rather challenging, and in the end
-the complexity of it gave rise to TemplateCoq
-\citep{malechaExtensibleProofEngineering2014}.
-
-Agda has historically experimented a lot with different extensions for
-both type system and the elaborator, even though the design doesn't
-accomodate these changes naturally. Instead, each of these extensions is
-spread throughout many different parts of the code-base\footnote{some
-  recent examples:
-  \href{https://github.com/agda/agda/pull/6385}{github.com/agda/agda/pull/6385},
-  \href{https://github.com/agda/agda/pull/6354}{github.com/agda/agda/pull/6354}.}.
-
-Lean introduced elaborator extensions
-\citep{leonardodemouraLeanMetaprogramming2021, ullrichNotationsHygienicMacro2020}.
-They allow the user to overload the commands, but if one defines a
-particular elaborator it becomes hard to interleave with others. In a
-way, this is an imperative view on extensibility.
-
-Idris
-\citep{bradyIdrisGeneralpurposeDependently2013, christiansenElaboratorReflectionExtending2016}
-appeared as a programming language first and proof-assistant second and
-does not provide either a plugin or hook system at all, except for
-reflection. Idris also focuses on tactics as the main mechanism for
-elaboration.
-
-Turstile+ by \citet{changDependentTypeSystems2019} uses macros to
-elaborate surface syntax to a smaller core. Macros allow them to
-modularly implement individual features, however, exactly as for Lean,
-combining different features which happens a lot in practise requires
-the user to re-define all macros from scratch.
-
-TypOS \citep{allaisTypOSOperatingSystem2022a, guillaumeallaisTypOS2022}
-is perhaps the closest to our work, but there are two important
-differences. First, they are building a domain-specific language for
-building type-checkers, while our design is language-agnostic, as long
-as it can model extensible datatypes in some capacity. Second, their
-approach settles features of the language as they are decided by the
-main developer and doesn't concern future changes and evolution.
-Finally, we try to stay close to the designs of existing dependently
-typed languages, while TypOS doesn't showcase support for them.
-
-\hypertarget{future-work}{%
-\section{Future work}\label{future-work}}
-
-We see three main prospects for future work:
-
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Exploration of different kinds of metavariables.} Currently we
-  implement metavariables only for terms, while for some applications
-  such as erasure \citep{tejiscakDependentlyTypedCalculus2020} or
-  irrelevance inference, it would be beneficial to have metavariables
-  representing erasure and relevance annotations. Alternatively, we can
-  introduce metavariables for names and implement data constructor
-  disambiguation in a simpler manner, since as of now we simply block on
-  expected type to disambiguate.
-\item
-  \textbf{Rendering more elements of the elaborator as constraints} --
-  currently components such as occurs-checker and reduction are simple
-  functions. Including them into the constraint machinery would make the
-  implementation more uniform and allow users to extend them.
-\item
-  \textbf{Potential optimisations}. Currently our system has a lot of
-  room for potential optimisations. The first step would be to allow
-  handlers to pass some information to their respective solvers, which
-  require introduction an existential type in the plugin. Additionally
-  some expandable per-plugin store for the solvers would be handy, for
-  example to not recompute type class instances on every invocation.
-  Somewhat more challenging, one can imagine a caching system for
-  constraints, allowing to avoid solving the same constraint twice. This
-  might be beneficial for reduction, since as of now we do a lot of
-  redundant computations. However, the memory usage might be
-  prohibitive. Finally, we would also like to explore possibilities for
-  concurrent solving, similar to the future plans of
-  \citet{allaisTypOSOperatingSystem2022a} with LVars for metavariables
-  \citep{kuperLatticebasedDataStructures2015}.
-\end{itemize}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\end{CSLReferences}
+\input{body.tex}
 

### 2024-03-04 paper/main.tex
intro and abstract attempt #1

@@
 \begin{abstract}
-Dependently typed languages such as Agda, Coq, Lean, and Idris are used for statically enforcing properties of programs and for enabling type-driven development.
-While there has been a lot of work to find the right theoretical foundations for the core type theory underlying these languages, their implementations have been studied to a smaller extent.
-In particular, theoretical works rarely consider the plethora of features that exist in big languages like Agda.
-As a consequence the developers have to make a lot of adaptations in the implementations, which leads to codebases that are hard to maintain.
-In this work we focus on a part of the type-checker called the elaborator that transforms user-friendly surface syntax into a small well-behaved core theory.
+  There has been a lot of work to find the right theoretical foundations for the core type theory underlying proof assistants and dependently-typed programming languages like Coq, Agda, Lean or Idris.
+  However, their implementations have been studied to a smaller extent.
+  Together with the additions of features that aren't reflected in theoretical studies, this means that developers have to make a lot of adaptations in the implementations, making the codebases that are hard to maintain.
+  In this work we consider some common problems in the implementations of the type-checkers and propose a new architecture that allows for a more modular development and containment of features to one module each.
+  We focus focus on a part of the type-checker called the elaborator that transforms user-friendly surface syntax with a lot of features into a small well-behaved core theory.
   Our core idea is to use an open datatype for constraints and a plugin system for solvers.
-This allows for a more compact base elaborator implementation while enabling extensions to the type system.
+  We showcase the design on a proof-of-concept implementation.
 \end{abstract}
@@
   pdftitle={ExEl: Building an Elaborator Using Extensible Constraints},
+  pdfauthor={Bohdan Liesnikov and Jesper Cockx},
+  pdfkeywords={dependent types, elaborators, elaboration, open datatype, extensible datatype, type-checking},
+  pdfcreator={LaTeX via pandoc},
+  pdfproducer={pdftex},
   hidelinks,
-  pdfcreator={LaTeX via pandoc}}
+  }
 

### 2024-03-04 paper/main.tex
Picking some nits in the abstract

@@
 \begin{abstract}
-  There has been a lot of work to find the right theoretical foundations for the core type theory underlying proof assistants and dependently-typed programming languages like Coq, Agda, Lean or Idris.
-  However, their implementations have been studied to a smaller extent.
-  Together with the additions of features that aren't reflected in theoretical studies, this means that developers have to make a lot of adaptations in the implementations, making the codebases that are hard to maintain.
-  In this work we consider some common problems in the implementations of the type-checkers and propose a new architecture that allows for a more modular development and containment of features to one module each.
-  We focus focus on a part of the type-checker called the elaborator that transforms user-friendly surface syntax with a lot of features into a small well-behaved core theory.
-  Our core idea is to use an open datatype for constraints and a plugin system for solvers.
-  We showcase the design on a proof-of-concept implementation.
+  Proof assistants and dependently typed languages such as Coq, Agda, Lean, and Idris can be used to ascertain the correctness of software with mathematical precision.
+  While much research has been done on their theoretical foundations, their actual implementations have been studied to a much lesser extent.
+  As a result, features that are not considered part of the theoretical foundations - such as implicit arguments and type classes - have their own bespoke implementation for each language, making for code bases that are hard to understand and maintain.
+  To address some of the common problems in the implementations of dependently typed languages, we present a modular architecture for implementing the transformation from user-friendly surface syntax into a small and well-behaved core language, also known as elaboration.
+  Our architecture is made modular through the use of an open type of constraints and a plugin system for solvers that work on these constraints, which means that each new feature is contained in its own module.
+  We showcase our design with a proof-of-concept elaborator for a language with dependent types, implicit arguments, higher-order unification, implicit coercions, and instance arguments.
 \end{abstract}

### 2024-03-15 paper/main.tex
last minute wording changes

@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\usepackage[]{natbib}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+
 \widowpenalty=10000
@@
 \displaywidowpenalty=10000
+
+\title{ExEl: Building an Elaborator Using Extensible Constraints}
+
 \author{Bohdan Liesnikov}
@@
 \email{J.G.H.Cockx@tudelft.nl}
-\begin{abstract}
-  Proof assistants and dependently typed languages such as Coq, Agda, Lean, and Idris can be used to ascertain the correctness of software with mathematical precision.
-  While much research has been done on their theoretical foundations, their actual implementations have been studied to a much lesser extent.
-  As a result, features that are not considered part of the theoretical foundations - such as implicit arguments and type classes - have their own bespoke implementation for each language, making for code bases that are hard to understand and maintain.
-  To address some of the common problems in the implementations of dependently typed languages, we present a modular architecture for implementing the transformation from user-friendly surface syntax into a small and well-behaved core language, also known as elaboration.
-  Our architecture is made modular through the use of an open type of constraints and a plugin system for solvers that work on these constraints, which means that each new feature is contained in its own module.
-  We showcase our design with a proof-of-concept elaborator for a language with dependent types, implicit arguments, higher-order unification, implicit coercions, and instance arguments.
-\end{abstract}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\usepackage[]{natbib}
-%\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+
 \hypersetup{
@@
 
-\title{ExEl: Building an Elaborator Using Extensible Constraints}
-\author{}
-\date{\today}
+\begin{abstract}
+  Proof assistants and dependently typed languages such as Coq, Agda, Lean, and Idris can be used to ascertain the correctness of software with mathematical precision.
+  While much research has been done on their theoretical foundations, their actual implementations have been studied to a much lesser extent.
+  As a result, features that are not considered part of the theoretical foundations - such as implicit arguments and type classes - have their own bespoke implementation for each language, making for code bases that are hard to understand and maintain.
+  To address some of the common problems in the implementations of dependently typed languages, we present a modular architecture for implementing the transformation from user-friendly surface syntax into a small and well-behaved core language, also known as elaboration.
+  Our architecture is made modular through the use of an open datatype of constraints and a plugin system for solvers that work on these constraints, which means that each new feature is contained in its own module.
+  We showcase our design with a proof-of-concept elaborator for a language with dependent types, implicit arguments, higher-order unification, and instance arguments.
+\end{abstract}
 
@@
 
+\begin{acks}
+  Jesper Cockx holds an \grantsponsor{nwoviveni202216}{NWO Veni}{https://www.nwo.nl/en/projects/viveni202216} grant on `A trustworthy and extensible core language for Agda' \grantnum[https://www.nwo.nl/en/projects/viveni202216]{nwoviveni202216}{(VI.Veni.202.216)}.
+\end{acks}
+
 \bibliographystyle{ACM-Reference-Format}

### 2024-03-22 camera-ready-submission/source/body.tex
add all the files

\hypertarget{sec:introduction}{%
\section{Introduction}\label{sec:introduction}}

Statically typed languages allow us to catch large classes of bugs at
compile-time by checking the implementation against its type signature.
When the types are provided by the user they can be viewed as a form of
specification, constraining the behaviour of the programs. This comes
with the benefit of more static guarantees but with an increased toll on
the user to supply more precise information about the program. Many
languages choose to infer types as a result, but another option is to
use the information in the types to infer parts of the program. This
idea was aptly worded by Conor McBride as ``Write more types and fewer
programs.''
\citetext{\citealp{ptoopTypeInferenceThought2022}; \citealp[chap.~2.1]{mcbrideEpigramPracticalProgramming2005}}
Some examples of this include overloaded functions in Java, implicits in
Scala, and type classes in Haskell.

In dependently typed languages like Agda
\citep{norellPracticalProgrammingLanguage2007, theagdateamAgdaUserManual2023a},
Coq \citep{thecoqdevelopmentteamCoqProofAssistant2022} or Idris
\citep{bradyIdrisGeneralpurposeDependently2013} the types can be much
more precise. This allows us to infer even larger parts of the program
from the type. Examples include implicit arguments in Agda, implicit
coercions in Coq, and tactic arguments in Idris. The inference of parts
of the program must not be fully automatic but can also be interactive
or partially automatic. Examples of interactive inference are holes in
Agda and proof obligations in Coq, while canonical structures
\citep{mahboubiCanonicalStructuresWorking2013} in Coq and
program-synthesis for holes in Haskell
\citep{koppelSearchingEntangledProgram2022} are partially automatic.

Each of these different inference features has its own algoritms and
extension points, which often evolved organically over time together
with the language, and are often not well isolated from each other. For
example, implicit arguments and instance search in Agda can interact in
unexpected ways \citep{agdausersPerformanceRegressionIssue2018}. Sized
types in Agda \citep{abelExtensionMartinLofType2016} also come with
their own solver that often interacts poorly with the regular solver for
implicit arguments. In Coq, canonical structures are notorious for
producing unpredictable results yet they were not properly documented
for 15 years \citep{mahboubiCanonicalStructuresWorking2013}. Lean 4 aims
to allow the users to develop new surface-level features
\citep{leonardodemouraLeanMetaprogramming2021} using elaboration monads
\citep{mouraLeanTheoremProver2021}, somewhat akin to elaborator
reflection in Idris
\citep{christiansenElaboratorReflectionExtending2016}, but Lean 3 was
built in a more conventional way \citep{demouraLeanTheoremProver2015}.
All these bespoke algorithms and their interactions put a toll on the
language developer to specify and implement them and on the user to
understand them.

The part of the implementation of a dependently typed language that is
responsible for type checking user-facing surface syntax and inferring
the parts that have been left implicit is known as the
\emph{elaborator}. One common piece of infrastructure used by
elaborators are metavariables, also known as ``existential variables''
\citep[chap.~2.2.1]{thecoqdevelopmentteamCoqProofAssistant2022}, which
represent as-of-yet unknown parts of the program. Together with
metavariables also comes unification, i.e.~the ability to constrain two
terms to be equal. Metavariables and unification are heavily used
throughout many elaborators for infering implicit arguments and for
general type-checking, making them sensitive towards changes in
unification algorithms. Because of the complexity unification, breaking
changes are often discovered only when run against a large existing
project on CI, such as the Standard and Cubical libraries for Agda or
the \texttt{unimath} library for Coq.

To move towards a cleaner and more maintainable model for implementing
elaborators, we propose a new architecture for an extensible elaborator
for dependently typed languages. Practically, our architecture allows
each feature to be contained within one module, as opposed to being
spread around the codebase. The implementations of the individual
features interact with the rest of the elaborator through an API for
providing new kinds of constraints and new solvers for these new
constraints as well as existing ones. This allows developers to put all
solvers related to one feature in one place, thus fulfilling our goal of
modularity. The design also separates the `what' the solvers are doing
from the `when', making the separation between different phases of
elaboration explicit. As a result, this allows the developer to reason
more easily about exceptions and asynchronicity during elaboration and
add new features in a more isolated fashion.

\textbf{Contributions}.

\begin{itemize}
\tightlist
\item
  We propose a new design blueprint for implementing an elaborator for a
  dependently typed language that is extensible with new constraints and
  new solvers. It supports type classes (Section
  \ref{sec:case-typeclasses}), implicit arguments (Section
  \ref{sec:case-implicits}), implicit coercions, and tactic arguments
  (Section \ref{sec:coercion-tactics}).
\item
  We propose a new view on metavariables as communication channels for
  the solvers, drawing parallels with asynchronous programming
  primitives (Section \ref{sec:solvers-implementation}).
\item
  We decompose the usual components of an elaborator, like the unifier
  in Agda, into a suite of solvers which can be extended and interleaved
  by user-provided plugins (Section
  \ref{sec:constraints_and_unification}).
\item
  Following the blueprint, we present a prototype implementation of a
  dependently typed language available at
  \href{https://github.com/liesnikov/extensible-elaborator}{github.com/liesnikov/extensible-elaborator}\footnote{Also archived at \href{https://doi.org/10.4121/e74fd14a-da79-4686-97e8-143ac5e0858d.v1}{doi.org/10.4121/e74fd14a-da79-4686-97e8-143ac5e0858d.v1}.}.
\end{itemize}

\hypertarget{sec:unification_constraint_based_elaboration_and_design_challanges}{%
\section{Unification, constraint-based elaboration and design
challenges}\label{sec:unification_constraint_based_elaboration_and_design_challanges}}

Constraints have been an integral part of compilers for strongly typed
languages for a long time \citep{oderskyTypeInferenceConstrained1999}.
For example, the implementations of both Haskell
\citep{vytiniotisOutsideInModularType2011} and Agda \citep[
chap.~3]{norellPracticalProgrammingLanguage2007} use constraints
extensively. In the former case, they are even reflected and can be
manipulated by the user
\citetext{\citealp[chap.~6.10.3]{orchardHaskellTypeConstraints2010a}; \citealp{ghcdevelopmentteamGHCUserGuide2022}}.
This has proven to be a good design decision for GHC, as is reflected
for example in a talk by \citet{peytonjonesTypeInferenceConstraint2019},
as well as in a few published sources
\citep{vytiniotisOutsideInModularType2011, peytonjonesPracticalTypeInference2007}.

In the land of dependently typed languages constraints are often used in
a much less principled manner. Agda has a family of constraints that
grew organically, currently counting 19 constructors.\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Monad/Base.hs\#L1157-L1191}{./src/full/Agda/TypeChecking/Monad/Base.hs\#L1157-L1191}.
  Here and henceforth we shorten the links in footnotes to paths in the
  repository, the source code can be found at
  \href{https://github.com/agda/agda/blob/v2.6.4/}{github.com/agda/agda/blob/v2.6.4/}.}
Idris technically has constraints, with the only two constructors being
equality constraints for two terms and for two sequences of
terms.\footnote{\href{https://github.com/idris-lang/Idris2/blob/e673d05a67b82591131e35ccd50fc234fb9aed85/src/Core/UnifyState.idr}{./src/Core/UnifyState.idr}
  at\\
  \href{https://github.com/idris-lang/Idris2/blob/e673d05a67b82591131e35ccd50fc234fb9aed85}{github.com/idris-lang/Idris2/blob/e673d0}}
The same holds for Lean.\footnote{\href{https://github.com/leanprover/lean4/blob/0a031fc9bbb43c274bb400f121b13711e803f56c/src/Lean/Meta/Match/Basic.lean\#L161}{./src/Lean/Meta/Match/Basic.lean\#L161}
  at\\
  \href{https://github.com/leanprover/lean4/blob/0a031fc9bbb43c274bb400f121b13711e803f56c/}{github.com/leanprover/lean4/blob/0a031f}}
These languages either use constraints in a restricted, single-use-case
manner -- namely, for unification -- or in an ad-hoc manner.

In this section, we demonstrate why a more methodical approach to
constraints results in more robust elaborators across the board. In
particular, we go over three typical challenges that come up when
building a compiler for a dependently typed language and the way they
are usually solved: the complexity of managing a global state of
metavariables for unification (Section \ref{sec:conversion_checking}),
dealing with different kinds of implicit arguments and their solvers
(\ref{sec:implicit-arguments}), and user-facing extension points to
unification (Section \ref{sec:extending-unification}).

\hypertarget{sec:conversion_checking}{%
\subsection{Unification in the presence of
meta-variables}\label{sec:conversion_checking}}

As mentioned in the introduction, in the process of type-checking we use
unification to compare terms. The unification algorithms are often one
of the most complex parts of a type checker, which stems from the desire
of compiler writers to implement the most powerful unifier while being
limited by the fact that higher-order unification is undecidable in
general. Some of this complexity is unavoidable, but we can manage it
better by splitting up the unifier into smaller modular parts. In
practice, this means that one does not have to fit together an
always-growing unifier but can instead write different unification rules
separately.

As a real-world example, Agda's unification algorithm for solving
implicit arguments is spread between about a hundred functions and 2200
lines of code\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs}{./src/full/Agda/TypeChecking/Conversion.hs}}.
Each of the functions implements part of the ``business logic'' of the
unifier, but a large part of their implementation is just there to deal
with bookkeeping related to metavariables and constraints:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  They throw and catch exceptions, driving the control flow of the
  unification.
\item
  They compute blockers that determine when a postponed constraint
  should be re-tried.
\item
  They have special cases for when either or both of the sides equation
  or its type are metavariables, or for when they are terms whose
  evaluation is blocked on some metavariables.
\end{enumerate}

Concretely, this code uses functions like \texttt{noConstraints},
\texttt{dontAssignMetas}, \texttt{catchConstraint}, and
\texttt{patternViolation} which rely on specific behaviour of the
constraint solver. Other functions like \texttt{reduceB} and
\texttt{abortIfBlocked} force the programmer to choose between letting
the constraint system handle blockers or doing it manually. These things
are known to be brittle and pose an increased mental overhead when
making changes, with a corresponding risk of introducing new bugs.

The unifier is called from many places throughout the type-checker: when
checking applications of variables and defined functions, type
annotations on lambda expressions, forced `dot' patterns, macro calls,
and various primitive operations of Cubical Agda. At the same time, it
is unintuitive and full of intricacies as indicated by multiple
comments\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs\#L484-L485}{./src/full/Agda/TypeChecking/Conversion.hs\#\#L484-L485},
  \href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs\#L541-L549}{./src/full/Agda/TypeChecking/Conversion.hs\#L541-L549}}.

We would like the compiler-writer to be able to separate managing
constraints and blockers from the actual logic of the comparison
function. If we zoom in on the \texttt{compareAtom} function, the core
logic can be expressed in about 20 lines\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Conversion.hs\#L550-L599}{./src/full/Agda/TypeChecking/Conversion.hs\#L550-L599}}
of simplified code (ignoring advanced features like sized types,
cumulativity, polarity, and forcing). This is precisely what we would
like the developer to write.

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{case}\NormalTok{ (m, n) }\KeywordTok{of}
\NormalTok{  (}\DataTypeTok{Lit}\NormalTok{ l1, }\DataTypeTok{Lit}\NormalTok{ l2) }\OperatorTok{|}\NormalTok{ l1 }\OperatorTok{==}\NormalTok{ l2 }\OtherTok{{-}\textgreater{}} \FunctionTok{return}\NormalTok{ ()}
\NormalTok{  (}\DataTypeTok{Var}\NormalTok{ i es, }\DataTypeTok{Var}\NormalTok{ i\textquotesingle{} es\textquotesingle{}) }\OperatorTok{|}\NormalTok{ i }\OperatorTok{==}\NormalTok{ i\textquotesingle{} }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
\NormalTok{      a }\OtherTok{\textless{}{-}}\NormalTok{ typeOfBV i}
\NormalTok{      compareElims [] [] a (var i) es es\textquotesingle{}}
\NormalTok{  (}\DataTypeTok{Con}\NormalTok{ x ci xArgs, }\DataTypeTok{Con}\NormalTok{ y \_ yArgs) }\OperatorTok{|}\NormalTok{ x }\OperatorTok{==}\NormalTok{ y }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
\NormalTok{      t\textquotesingle{} }\OtherTok{\textless{}{-}}\NormalTok{ conType x t}
\NormalTok{      compareElims t\textquotesingle{} (}\DataTypeTok{Con}\NormalTok{ x ci []) xArgs yArgs}
  \OperatorTok{...}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

The functions described above are specific to Agda but in other major
languages we can find similar problems with unifiers being large pieces
of code that are hard to understand. The sizes of modules with unifiers
are as follows: Idris has 1.5kloc\footnote{\href{https://github.com/idris-lang/Idris2/blob/102d7ebc18a9e881021ed4b05186cccda5274cbe/src/Core/Unify.idr}{./src/Core/Unify.idr}}
of unification code, Lean 1.8kloc\footnote{\href{https://github.com/leanprover/lean4/blob/75252d2b85df8cb9231020a556a70f6d736e7ee5/src/Lean/Meta/ExprDefEq.lean}{./src/Lean/Meta/ExprDefEq.lean}},
and Coq 1.8kloc\footnote{\href{https://github.com/coq/coq/blob/b35c06c3ab3ed4911311b4a9428a749658d3eff1/pretyping/evarconv.mli}{./pretyping/evarconv.mli}
  at
  \href{github.com/coq/coq/blob/b35c06c3ab3ed4911311b4a9428a749658d3eff1/}{github.com/coq/coq/blob/b35c06}}.
For Haskell, which is not a dependently typed language yet, but does
have a constraints system
\citep{peytonjonesTypeInferenceConstraint2019}, this number is at
2kloc\footnote{\href{https://gitlab.haskell.org/ghc/ghc/-/blob/b81cd709df8054b8b98ac05d3b9affcee9a8b840/compiler/GHC/Core/Unify.hs}{./compiler/GHC/Core/Unify.hs}
  at\\
  \href{https://gitlab.haskell.org/ghc/ghc/-/blob/b81cd709df8054b8b98ac05d3b9affcee9a8b840}{gitlab.haskell.org/ghc/ghc/-/blob/b81cd709d}}.

\hypertarget{sec:implicit-arguments}{%
\subsection{Type-checking function application in the presence of
implicit arguments}\label{sec:implicit-arguments}}

During the type-checking of function application, there may be different
kinds of implicit arguments to infer, for example, instance arguments,
or tactic arguments. If we start from a simple case of type-checking an
application of a function symbol to regular arguments, every next
extension needs to be handled as a special case.

Take Agda -- when checking an application of a function with implicit
arguments\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Implicit.hs\#L96-L128}{./src/full/Agda/TypeChecking/Implicit.hs\#L96-L128}}
we already have to carry the information on how the arguments will be
resolved (e.g.~through instance search) and then create a specific kind
of metavariable\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Implicit.hs\#L130-L149}{./src/full/Agda/TypeChecking/Implicit.hs\#L130-L149}}
\citep[chap.~3]{norellPracticalProgrammingLanguage2007} for each of
those cases.

For handling \texttt{auto} variables, Idris 2
\citep[chap.~13.1]{theidristeamIdrisTutorial2021} has to essentially
inline the search procedure through a chain of elaboration function
calls (\texttt{checkApp} to \texttt{checkAppWith} to
\texttt{checkAppWith\textquotesingle{}}) to
\texttt{makeAutoImplicit}\footnote{\href{https://github.com/idris-lang/Idris2/blob/870bc824371d504a03af937f326216302210a875/src/TTImp/Elab/App.idr\#L224-L241}{./src/TTImp/Elab/App.idr\#L224-L241}}.
This can accommodate interfaces (or type classes), but one can imagine
that if a different kind of implicit was added, like tactic arguments or
canonical structures, we would have to inline the search again,
requiring a non-trivial modification to the elaboration mechanism.

While the codebases above show that it is certainly possible to extend
languages with new features if the language was not written with
extensibility in mind, this can lead to rather ad hoc solutions.

\hypertarget{sec:extending-unification}{%
\subsection{Extending unification}\label{sec:extending-unification}}

While writing a unifier is hard enough as it is, at times the developers
might want to give their users the ability to extend the unification
procedure.

Canonical structures
\citep{saibiOutilsGeneriquesModelisation1999, mahboubiCanonicalStructuresWorking2013}
are one example as they are in the overlap between type classes and
unification hints. Adding them to a language that does not support them
requires an extension of the unification algorithm with a rule that says
that projection from a canonical structure is injective \citep[eq.
1]{mahboubiCanonicalStructuresWorking2013}.

One could also provide means to do so manually in a more general case,
by allowing users to declare certain symbols as injective. This is one
of the features requested by the Agda users
\citep{agdausersInjectiveUnificationPragma2023}.

Another example of user-extensible unification can be found in
extensions with rules for associativity and commutativity, as described
in the thesis by \citet{holtenDependentTypeCheckingModulo2023}. It
required changes to 2000 lines of code\footnote{\href{https://github.com/LHolten/agda-commassoc/tree/defenitional-commutativity}{github.com/LHolten/agda-commassoc/tree/defenitional-commutativity}
  (sic)}. We would like to make changes such as this more feasible.

\hypertarget{sec:summary-of-the-issues}{%
\subsection{Summary of the issues}\label{sec:summary-of-the-issues}}

The examples above show that when building a dependently typed language
the core might be perfectly elegant and simple, but the features that
appear on top of it complicate the design. While the code in existing
implementations often relies on constraints, the design at large does
not put them at the centre of the picture and instead views them
primarily as a gadget. For example, Agda's constraint solver\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Constraints.hs\#L247-L298}{src/full/Agda/TypeChecking/Constraints.hs\#L247-L298}}
relies on the type-checker to call it at the point where it is needed
and has to be carefully engineered to work with the rest of the
codebase.

\hypertarget{sec:what-is-our-design-bringing-into-the-picture}{%
\section{A blueprint for extensible
elaborators}\label{sec:what-is-our-design-bringing-into-the-picture}}

Our idea for a new design is to shift focus more towards the constraints
themselves:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We give an API for raising constraints that can be called by the
  type-checker, essentially creating an ``ask'' to be fulfilled by the
  solvers. This is similar to the idea of mapping object-language
  unification variables to host-language ones as done by
  \citet{guidiImplementingTypeTheory2017}, the view of the ``asks'' as a
  general effect \citep[chap.~4.4]{bauerEqualityCheckingGeneral2020}, or
  the communication between actors
  \citep{allaisTypOSOperatingSystem2022a}.
\item
  To make the language more modular, we make constraints an extensible
  data type and give an API to define new solvers with the ability to
  specify what kinds of constraints they can solve. Since we're working
  in Haskell we encode constraints in the style of Data types à la carte
  \citep{swierstraDataTypesCarte2008}.
\end{enumerate}

In the examples in this paper, we follow the bidirectional style of
type-checking. In practice, however, the design decisions are agnostic
of the underlying system, as long as it adheres to the principle of
stating the requirements on terms in terms of raising a constraint and
not by, say, pattern-matching on a concrete term representation.

\begin{figure*}
  \center
  \includegraphics[width=\textwidth]{architecture-diagram}
  \caption{Architecture diagram}
  \label{architecture-figure}
\end{figure*}

From a birds-eye view, the architecture looks as depicted in Figure
\ref{architecture-figure}. Its main characteristics are the separation
of plugins/unification into independent pieces and a clear boundary
between solver dispatcher and the rest of the system.

Let us walk through each part of the architecture in more detail. The
type-checking begins by initializing the state and traversing the
syntax. The traversal raises the constraints, and for the moment, the
constraints are simply stored. As soon as we finish the traversal of a
block (one declaration in our case), the solver dispatcher is called. It
traverses the set of constraints, and tries to solve each active
constraint by making calls to different plugins. Each plugin, whether
user-supplied (\texttt{Plugin\ A}) or provided by us
(\texttt{unification}) consists of a handler and a solver. The handler
determines if the plugin can potentially solve a constraint, if so, the
dispatcher runs the corresponding solver.

All components have some read access to the state, including handlers
which might for example verify that there are no extra constraints on
the metavariable. For the write access: syntax traversal writes new
metavariables to the state and elaborated definitions; the solver
dispatcher writes updated meta-information; solvers write solutions to
the metavariables and can raise new constraints.

For the moment we need to recompile the project to include new plugins.
This is not necessity and a system that dynamically loads plugins is
possible to implement in a way that is similar to GHC Plugins\footnote{\href{https://mpickering.github.io/plugins.html}{mpickering.github.io/plugins.html}}
or Accelerate\footnote{\href{https://github.com/tmcdonell/accelerate-llvm/blob/master/accelerate-llvm-native/src/Data/Array/Accelerate/LLVM/Native/Link/Runtime.hs\#L40}{github.com/tmcdonell/accelerate-llvm/blob/master/accelerate-llvm-native/src/Data/Array/Accelerate/LLVM/Native/Link/Runtime.hs}}
\citep{mcdonellTypesafeRuntimeCode2015}.

\hypertarget{sec:bidirectional}{%
\section{Dependently typed calculus and bidirectional
typing}\label{sec:bidirectional}}

Now that we have described our architecture, we will demonstrate its
design through a practical implementation. First, in this section, we
describe the core of the type system we implement. The following
sections will then describe the constraints and their solvers. We take
pi-forall \citep{weirichImplementingDependentTypes2022} as a basis for
the system and extend it with metavariables in the core syntax and
implicit arguments in the surface syntax. However, for all other
purposes, we leave the core rules intact and, therefore, the core
calculus too.

\hypertarget{basic-language-and-rules}{%
\subsection{Basic language and rules}\label{basic-language-and-rules}}

This is a dependently typed calculus that includes Pi, Sigma and indexed
inductive types.

Here is the internal syntax term data type, apart from service
constructor omissions, like a placeholder \texttt{TRUSTME}.

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{Term} \OtherTok{=}
  \CommentTok{{-}{-} type of types Type}
    \DataTypeTok{Type}
  \CommentTok{{-}{-} variables x}
  \OperatorTok{|} \DataTypeTok{Var} \DataTypeTok{TName}
  \CommentTok{{-}{-} abstraction \textbackslash{}x. a}
  \OperatorTok{|} \DataTypeTok{Lam}\NormalTok{ (}\DataTypeTok{Bind} \DataTypeTok{TName} \DataTypeTok{Term}\NormalTok{)}
  \CommentTok{{-}{-} application a b}
  \OperatorTok{|} \DataTypeTok{App} \DataTypeTok{Term} \DataTypeTok{Arg}
  \CommentTok{{-}{-} function type (x : A) {-}\textgreater{} B}
  \OperatorTok{|} \DataTypeTok{Pi} \DataTypeTok{Type}\NormalTok{ (}\DataTypeTok{Bind} \DataTypeTok{TName} \DataTypeTok{Type}\NormalTok{)}
  \CommentTok{{-}{-} Sigma{-}type \{ x : A | B \}}
  \OperatorTok{|} \DataTypeTok{Sigma} \DataTypeTok{Term}\NormalTok{ (}\DataTypeTok{Bind} \DataTypeTok{TName} \DataTypeTok{Term}\NormalTok{)}
  \OperatorTok{|} \DataTypeTok{Prod} \DataTypeTok{Term} \DataTypeTok{Term}
  \OperatorTok{|} \DataTypeTok{LetPair} \DataTypeTok{Term}\NormalTok{ (}\DataTypeTok{Bind}\NormalTok{ (}\DataTypeTok{TName}\NormalTok{, }\DataTypeTok{TName}\NormalTok{) }\DataTypeTok{Term}\NormalTok{)}
  \CommentTok{{-}{-} Equality type  a = b}
  \OperatorTok{|} \DataTypeTok{TyEq} \DataTypeTok{Term} \DataTypeTok{Term}
  \OperatorTok{|} \DataTypeTok{Refl}
  \OperatorTok{|} \DataTypeTok{Subst} \DataTypeTok{Term} \DataTypeTok{Term}
  \OperatorTok{|} \DataTypeTok{Contra} \DataTypeTok{Term}
  \CommentTok{{-}{-} inductive datatypes}
  \OperatorTok{|} \DataTypeTok{TCon} \DataTypeTok{TCName}\NormalTok{ [}\DataTypeTok{Arg}\NormalTok{] }\CommentTok{{-}{-} types (fully applied)}
  \OperatorTok{|} \DataTypeTok{DCon} \DataTypeTok{DCName}\NormalTok{ [}\DataTypeTok{Arg}\NormalTok{] }\CommentTok{{-}{-} terms (fully applied)}
  \OperatorTok{|} \DataTypeTok{Case} \DataTypeTok{Term}\NormalTok{ [}\DataTypeTok{Match}\NormalTok{]}
    \CommentTok{{-}{-} metavariables}
  \OperatorTok{|} \DataTypeTok{MetaVar} \DataTypeTok{MetaClosure}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

Equality is not defined as a regular inductive type but is instead
built-in. The user has access to the type and term constructor, but not
the ability to pattern-match on it. Instead, the language provides a
\texttt{subst} primitive of type
\texttt{(A\ x)\ -\textgreater{}\ (x=y)\ -\textgreater{}\ A\ y} and
\texttt{contra} that takes an equality of two different constructors of
an inductive type and produces an element of any type.

On top of the above, the language includes indexed inductive datatypes
and case-constructs for their elimination. Indexed inductive datatypes
are encoded as parameterised datatypes with an equality argument
constraining the index, also known as ``Henry Ford'' equality
\citep{chapmanGentleArtLevitation2010a}.

As for metavariables \texttt{MetaVar}: as mentioned in the introduction,
they are placeholders in the syntax tree (AST) that are produced in the
process of elaboration. Metavariables do not appear in the surface
syntax as they are not created by the user. In this paper we implement
metavariables in the contextual style, as described by
Abel and Pientka \cite{abelHigherOrderDynamicPattern2011}, therefore they are paired
with a closure of type \texttt{MetaClosure}.

\hypertarget{syntax-traversal}{%
\subsection{Syntax traversal}\label{syntax-traversal}}

We implement the core of the elaborator as a bidirectional syntax
traversal, raising a constraint every time we need to assert something
about the type.

This includes the expected use of unification constraints, for example
the case when we enter checking mode with a term whose type can be
inferred:

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checkType tm ty }\OtherTok{=} \KeywordTok{do}
\NormalTok{  (etm, ty\textquotesingle{}) }\OtherTok{\textless{}{-}}\NormalTok{ inferType tm}
\NormalTok{  constrainEquality ty\textquotesingle{} ty }\DataTypeTok{I.Type}
  \FunctionTok{return}\NormalTok{ etm}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

It also includes any time we want to decompose the type provided in
checking mode we pose a constraint that guarantees the type being in a
specific shape:

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checkType (}\DataTypeTok{S.Lam}\NormalTok{ lam) ty }\OtherTok{=} \KeywordTok{do}
\NormalTok{  mtyA }\OtherTok{\textless{}{-}}\NormalTok{ createMetaTerm}
\NormalTok{  mtx }\OtherTok{\textless{}{-}}\NormalTok{ createUnknownVar}
  \CommentTok{{-}{-} extend the context with a new variable signature}
\NormalTok{  mtyB }\OtherTok{\textless{}{-}}\NormalTok{ extendCtx (}\DataTypeTok{I.TypeSig}\NormalTok{ (}\DataTypeTok{I.Sig}\NormalTok{ mtx mtyA))}
\NormalTok{                    (createMetaTerm)}
  \CommentTok{{-}{-} metaPi is the shape we want ty to be in}
  \KeywordTok{let}\NormalTok{ metaPi }\OtherTok{=} \DataTypeTok{I.Pi}\NormalTok{ mtyA (bind mtx mtyB)}

\NormalTok{  constrainEquality ty metaPi }\DataTypeTok{I.Type}
  \CommentTok{{-}{-} rest of the traversal can now use mtyA and mbnd}
  \OperatorTok{...}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

At certain points, we have to raise a constraint which has an associated
continuation. For example, when checking the type of a data constructor,
the part of the program that comes as an argument to
\texttt{constrainTConAndFreeze} will be suspended (or ``blocked'') until
the meta is solved with something of the shape \texttt{TCon\ \_\ \_}.

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checkType t}\OperatorTok{@}\NormalTok{(}\DataTypeTok{S.DCon}\NormalTok{ c args) ty }\OtherTok{=} \KeywordTok{do}
\NormalTok{  elabpromise }\OtherTok{\textless{}{-}}\NormalTok{ createMetaTerm}
\NormalTok{  constrainTConAndFreeze ty }\OperatorTok{$} \KeywordTok{do}
\NormalTok{    mty }\OtherTok{\textless{}{-}}\NormalTok{ substMetas ty}
    \KeywordTok{case}\NormalTok{ mty }\KeywordTok{of}
\NormalTok{      (}\DataTypeTok{I.TCon}\NormalTok{ tname params) }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
        \OperatorTok{...}
\NormalTok{      \_ }\OtherTok{{-}\textgreater{}} \FunctionTok{error} \StringTok{"impossible"}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

We add one more case to the elaborator for implicit arguments of any
kind, as will be described in more detail in Section
\ref{sec:case-implicits}.

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checkType (}\DataTypeTok{Implicit}\NormalTok{) ty }\OtherTok{=} \KeywordTok{do}
\NormalTok{  m }\OtherTok{\textless{}{-}}\NormalTok{ createMetaTerm}
\NormalTok{  raiseConstraint }\OperatorTok{$} \DataTypeTok{FillInTheMeta}\NormalTok{ m ty}
  \FunctionTok{return}\NormalTok{ m}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

As we see, the syntax traversal does not need to know anything about how
implicit arguments are resolved. The only thing we require is that the
elaboration of the argument is called with the type information
available. This corresponds to how, in bidirectional typing, function
application is done in the inference mode but the arguments are
processed in checking mode.

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{inferType (}\DataTypeTok{App}\NormalTok{ t1 t2) }\OtherTok{=} \KeywordTok{do}
\NormalTok{  (et1, }\DataTypeTok{Pi}\NormalTok{ tyA tyB) }\OtherTok{\textless{}{-}}\NormalTok{ inferType t1}
\NormalTok{  et2 }\OtherTok{\textless{}{-}}\NormalTok{ checkType t2 tyA}
  \FunctionTok{return}\NormalTok{ (}\DataTypeTok{App}\NormalTok{ et1 et2, subst tyB et2)}
\end{Highlighting}
\end{Shaded}
\end{imageonly}
\hypertarget{sec:constraints_and_unification}{%
\section{Constraints and
unification}\label{sec:constraints_and_unification}}

In Section \ref{sec:bidirectional} we described the syntax traversal
part of the elaborator, which generates the constraints. In this
section, we will go over the constraint datatype needed for the base
language, how unification is implemented and how we can extend the
unification procedure.

\hypertarget{base-constraints}{%
\subsection{Base constraints}\label{base-constraints}}

The datatype of constraints is open, which means the user can write a
plugin to extend it. However, we offer a few constraints out of the box
to be able to type-check the base language. For the base language, it
suffices to have the following:

\begin{itemize}
\item
  A constraint that enforces equality of two terms of a given type

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} two terms given should be equal}
\KeywordTok{data} \DataTypeTok{EqualityConstraint}\NormalTok{ e }\OtherTok{=}
     \DataTypeTok{EqualityConstraint} \DataTypeTok{Term} \DataTypeTok{Term} \DataTypeTok{Type} \DataTypeTok{MetaVarId}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

  Here, \texttt{MetaVarId} refers to the anti-unification variable (see
  Section \ref{sec:solvers-implementation}).
\item
  A constraint that ensures that a metavariable is resolved eventually:

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} this terms has to be filled in}
\KeywordTok{data} \DataTypeTok{FillInTheTerm}\NormalTok{ e }\OtherTok{=}
     \DataTypeTok{FillInTheTerm} \DataTypeTok{Term}\NormalTok{ (}\DataTypeTok{Maybe} \DataTypeTok{Type}\NormalTok{)}
\end{Highlighting}
\end{Shaded}
\end{imageonly}
\item
  Lastly, a constraint which ensures that a term is a type constructor:

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} the term passed to the constraint}
\CommentTok{{-}{-} should be a type constructor}
\KeywordTok{data} \DataTypeTok{TypeConstructorConstraint}\NormalTok{ e }\OtherTok{=}
     \DataTypeTok{TypeConstructorConstraint} \DataTypeTok{Type}
\end{Highlighting}
\end{Shaded}
\end{imageonly}
\end{itemize}

The type-checker raises constraints and supplies the necessary
information in the process, but it is agnostic of how they will be
solved. In all of the examples above, the parameter \texttt{e} is
spurious. The need for this parameter comes from the technique we use to
encode open datatypes, and other constraint types can use it to encode
recursive occurrences of the constraints.

\hypertarget{sec:solvers-interface}{%
\subsection{Interface of the solvers}\label{sec:solvers-interface}}

On the solver side, we provide a suite\footnote{In the prototype we
  implement a subset of all unification rules, specifically:
  \texttt{identityPlugin}, \texttt{propagateMetasEqPlugin},
  \texttt{reduceLeftPlugin}, \texttt{reduceRightPlugin},
  \texttt{leftMetaPlugin}, \texttt{rightMetaPlugin},
  \texttt{typeConstructorPlugin},
  \texttt{typeConstructorWithMetasPlugin},
  \texttt{piEqInjectivityPlugin}, \texttt{tyEqInjectivityPlugin},
  \texttt{consInjectivityPlugin}, \texttt{typeInjectivityPlugin},
  \texttt{unificationStartMarker}, and \texttt{unificationEndMarker}.}
of solvers for unification that handle different cases of the problem.
The most basic plugin is the \texttt{syntactic} plugin which resolves
equality constraints where both sides are syntactically equal.

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{{-}{-} solves syntactically equal terms}
\OtherTok{syntacticHandler ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ c)}
                 \OtherTok{=\textgreater{}} \DataTypeTok{HandlerType}\NormalTok{ c}
\NormalTok{syntacticHandler constr }\OtherTok{=} \KeywordTok{do}
  \KeywordTok{let}\NormalTok{ eqcm }\OtherTok{=}\NormalTok{ match }\OperatorTok{@}\DataTypeTok{EqualityConstraint}\NormalTok{ constr}
  \KeywordTok{case}\NormalTok{ eqcm }\KeywordTok{of}
    \CommentTok{{-}{-} check for alpha{-}equality}
    \DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{EqualityConstraint}\NormalTok{ t1 t2 ty \_) }\OtherTok{{-}\textgreater{}}
      \FunctionTok{return} \OperatorTok{$}\NormalTok{ aeq t1 t2}
    \DataTypeTok{Nothing} \OtherTok{{-}\textgreater{}}
      \FunctionTok{return} \DataTypeTok{False}
\OtherTok{syntacticSolver ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ c)}
                \OtherTok{=\textgreater{}} \DataTypeTok{SolverType} \DataTypeTok{Bool}
\OtherTok{syntactic ::} \DataTypeTok{Plugin}
\NormalTok{syntactic  }\OtherTok{=} \DataTypeTok{Plugin}\NormalTok{ \{ solver  }\OtherTok{=}\NormalTok{ syntacticSolver}
\NormalTok{                    , handler }\OtherTok{=}\NormalTok{ syntacticHandler}
                    \OperatorTok{...}
\NormalTok{                    \}}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

We first define the class of constraints that will be handled by the
solver via providing a ``handler'' -- a function that decides whether a
given solver has to fire.\footnote{The code shown above and in the rest
  of the paper is close to the actual implementation, but has been
  simplified for presentation purposes. \texttt{HandlerType\ a} and
  \texttt{SolverType\ a} both morally correspond to
  \texttt{(ConstraintF\ cs)\ -\textgreater{}\ SolverMonad\ Bool}.} In
this case, this amounts to checking that the constraint given is indeed
an \texttt{EqualityConstraint} and that the two terms given to it are
syntactically equal. Then we define the solver itself. In this case does
not have to do anything except return \texttt{True} to indicate that the
constraint is solved. This is because it shall only fire once it has
been cleared to do so by the handler and the equality has already been
checked. Finally, we register the solver by declaring it using a plugin
interface.

The reason for the separation between a decision procedure and the
execution of the solver is to ensure separation between a potentially
slow, effectful solving and fast read-only decision-making in the
handler. We opt for this division since handlers will be run on many
constraints that do not fit them, therefore any write effects would have
to be rolled back. Solvers, on the other hand, should only fire in cases
when we can reasonably hope that the constraint will be solved and the
effects will not have to be rolled back.

In a similar fashion, we define \texttt{leftMetaSolver} and
\texttt{rightMetaSolver}, which only work on problems where one of the
sides is a metavariable. Here the job of the solver is not as trivial --
it has to check that the type of the other side indeed matches the
needed one and then register the instantiation of the metavariable in
the state.

Since a single constraint can often be handled by multiple solvers, we
can provide priority preferences, using the \texttt{pre} and
\texttt{suc} fields of the plugin interface. They are used to indicate
whether the currently defined plugin should run before or after,
respectively, which other plugins. At the time of running the compiler,
these preferences are loaded into a big pre-order relation for all the
plugins, which is then linearised and used to guide the solving
procedure.

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{rightMetaPlugin ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
                \OtherTok{=\textgreater{}} \DataTypeTok{Plugin}\NormalTok{ cs}
\NormalTok{rightMetaPlugin }\OtherTok{=}
  \DataTypeTok{Plugin}\NormalTok{ \{ handler }\OtherTok{=}\NormalTok{ rightMetaHandler}
\NormalTok{         , solver  }\OtherTok{=}\NormalTok{ rightMetaSolver}
\NormalTok{         , symbol  }\OtherTok{=}\NormalTok{ rightMetaSymbol}
\NormalTok{         , pre }\OtherTok{=}\NormalTok{ []}
\NormalTok{         , suc }\OtherTok{=}\NormalTok{ [leftMetaSymbol]}
\NormalTok{         \}}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

\hypertarget{sec:solvers-implementation}{%
\subsection{Implementation of the solvers and unification
details}\label{sec:solvers-implementation}}

We implement a system close to the one described by
Abel and Pientka \cite{abelHigherOrderDynamicPattern2011}. We modularise the
implementation by mapping every function call in the simplification
procedure to a raised constraint and every simplification rule to a
separate solver. For example, the ``decomposition of functions''
\citep[fig.~2]{abelHigherOrderDynamicPattern2011} rule is translated to
the following implementation.

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{piEqInjectivityHandler ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
                       \OtherTok{=\textgreater{}} \DataTypeTok{HandlerType}\NormalTok{ cs}
\NormalTok{piEqInjectivityHandler constr }\OtherTok{=} \KeywordTok{do}
  \KeywordTok{let}\NormalTok{ eqcm }\OtherTok{=}\NormalTok{ match }\OperatorTok{@}\DataTypeTok{EqualityConstraint}\NormalTok{ constr}
  \KeywordTok{case}\NormalTok{ eqcm }\KeywordTok{of}
    \DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{EqualityConstraint}\NormalTok{ pi1 pi2 \_ \_) }\OtherTok{{-}\textgreater{}}
      \KeywordTok{case}\NormalTok{ (pi1, pi2) }\KeywordTok{of}
\NormalTok{        (}\DataTypeTok{I.Pi}\NormalTok{ \_ \_ \_, }\DataTypeTok{I.Pi}\NormalTok{ \_ \_ \_) }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{True}
\NormalTok{        \_ }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
\NormalTok{    \_ }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

The handler is simply checking that both sides of the equality are
indeed Pi-types, and in case either of the matches fails, it will be
reported and the solver will not be fired. The \texttt{match} function
above comes from the open datatype of constraints
\citep[sec.~5]{swierstraDataTypesCarte2008}, checking if \texttt{constr}
can be projected from \texttt{cs} to \texttt{EqualityConstraint}.

Now let us take a look at the solver. The rule we are implementing here
states that two \(\Pi\)-types can only be equal if both the types of the
domains (\texttt{a1}, \texttt{a2}) and the co-domains (\texttt{b1},
\texttt{b2}) are equal.

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{piEqInjectivitySolver ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
                      \OtherTok{=\textgreater{}} \DataTypeTok{SolverType}\NormalTok{ cs}
\NormalTok{piEqInjectivitySolver constr }\OtherTok{=} \KeywordTok{do}
  \KeywordTok{let}\NormalTok{ (}\DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{EqualityConstraint}\NormalTok{ (}\DataTypeTok{I.Pi}\NormalTok{ a1 b1)}
\NormalTok{                                (}\DataTypeTok{I.Pi}\NormalTok{ a2 b2) \_ m)) }\OtherTok{=}
\NormalTok{        match }\OperatorTok{@}\DataTypeTok{EqualityConstraint}\NormalTok{ constr}
\NormalTok{  ma }\OtherTok{\textless{}{-}}\NormalTok{ constrainEquality a1 a2 }\DataTypeTok{I.Type}
\NormalTok{  (x, tyB1) }\OtherTok{\textless{}{-}}\NormalTok{ unbind b1}
\NormalTok{  (y, tyB2\textquotesingle{}) }\OtherTok{\textless{}{-}}\NormalTok{ unbind b2}
  \KeywordTok{let}\NormalTok{ tyB2 }\OtherTok{=}\NormalTok{ subst y (}\DataTypeTok{I.Var}\NormalTok{ x) tyB2\textquotesingle{}}
\NormalTok{      mat  }\OtherTok{=}\NormalTok{ I.identityClosure ma}
\NormalTok{  mb }\OtherTok{\textless{}{-}}\NormalTok{ extendCtx (}\DataTypeTok{I.TypeSig}\NormalTok{ (}\DataTypeTok{I.Sig}\NormalTok{ x e1 mat)) }\OperatorTok{$}
\NormalTok{                  constrainEquality tyB1 tyB2 }\DataTypeTok{I.Type}
  \KeywordTok{let}\NormalTok{ mbt }\OtherTok{=}\NormalTok{ bind x }\OperatorTok{$}\NormalTok{ I.identityClosure mb}
\NormalTok{  solveMeta m (}\DataTypeTok{I.Pi}\NormalTok{ mat mbt)}
  \FunctionTok{return} \DataTypeTok{True}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

Following pi-forall, we use the unbound-generics\footnote{\href{https://hackage.haskell.org/package/unbound-generics-0.4.3}{hackage.haskell.org/package/unbound-generics-0.4.3}}
library to deal with the names and binders (\texttt{unbind},
\texttt{subst} functions). As it will fire after a handler returns
\texttt{True}, we can assume the pattern-matches will not fail.

First, we constrain the equality of the domain of the Pi-type:
\texttt{a1} and \texttt{a2}. The seemingly spurious metavariable
\texttt{ma} returned from this call serves as an anti-unification
\citep{pfenningUnificationAntiunificationCalculus1991} communication
channel. Every time an equality constraint is created we return a
metavariable that stands for the unified term. This metavariable is used
for unification problems that are created in the extended contexts -- in
this case second argument of the Pi-type, but also when solving
equalities concerning two data constructors. We do this to tackle the
``spine problem''
\citep[sec.~1.4]{victorlopezjuanPracticalHeterogeneousUnification2021}
-- as we operate according to the ``well-typed modulo constraints''
principle, essentially providing a placeholder that is guaranteed to
preserve well-typedness in the extended context.\footnote{\(\text{Tog}^{+}\)
  \citep{victorlopezjuanTog2020, victorlopezjuanPracticalHeterogeneousUnification2021}
  focuses on extending the unification algorithm for the case where two
  sides of equality might not be of the same type, which is also a
  problem relevant for us. Their main argument against the usage of
  anti-unification in Agda provided there is that it is bug-prone. We
  think that in Agda the problems were stemming from the fact that
  anti-unification was implemented separately from unification, in which
  case it is indeed hard to keep the two in sync. There is no such
  duplication in our case since unification and anti-unification are
  one.} Finally, \texttt{ma} has to be applied to a closure, which will
keep track of the delayed substitution.

Then we can constrain the co-domains of Pi-types in an extended context.
In case one of the solvers the constraints created in the extended
context might need to know the exact shape of \texttt{ma}, we can block
on the metavariable later, freezing the rest of the problem until it is
instantiated.

As for the actual unification steps, we implement them in a similar
fashion to the simplification procedure. For example,
\texttt{leftMetaSolver} below handles the case where the left-hand side
of an equality constraint is an unsolved meta:

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{leftMetaSolver ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
               \OtherTok{=\textgreater{}} \DataTypeTok{SolverType}\NormalTok{ cs}
\NormalTok{leftMetaSolver constr }\OtherTok{=} \KeywordTok{do}
  \KeywordTok{let}\NormalTok{ (}\DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{EqualityConstraint}\NormalTok{ t1 t2 \_ m)) }\OtherTok{=}
\NormalTok{        match }\OperatorTok{@}\DataTypeTok{EqualityConstraint}\NormalTok{ constr}
\NormalTok{      (}\DataTypeTok{MetaVar}\NormalTok{ (}\DataTypeTok{MetaVarClosure}\NormalTok{ m1 c1)) }\OtherTok{=}\NormalTok{ t1}
\NormalTok{  mt2 }\OtherTok{\textless{}{-}}\NormalTok{ occursCheck m1 t2}
  \KeywordTok{case}\NormalTok{ mt2 }\KeywordTok{of}
    \CommentTok{{-}{-} indicates a failure in occurs{-}check}
    \DataTypeTok{Left}\NormalTok{ e }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
    \CommentTok{{-}{-} indicates a passed occurs{-}check}
    \DataTypeTok{Right}\NormalTok{ t2 }\OtherTok{{-}\textgreater{}}
      \KeywordTok{case}\NormalTok{ (invertClosureOn c1 (freeVarList t2)) }\KeywordTok{of}
        \DataTypeTok{Just}\NormalTok{ s }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
          \KeywordTok{let}\NormalTok{ st2 }\OtherTok{=}\NormalTok{ substs s t2}
          \CommentTok{{-}{-} apply the subsitution}
\NormalTok{          solveMeta m1 st2}
          \CommentTok{{-}{-} instantiate the anti{-}unification variable}
\NormalTok{          solveMeta m st2}
          \FunctionTok{return} \DataTypeTok{True}
        \DataTypeTok{Nothing} \OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

Once the occurs-check returns and if it was successful, we apply the
inverted closure to the right-hand side of the equality.

By splitting up the rules into individual, simple solvers we can
compartmentalise the complexity of the unifier, making sure that each
rule is as decoupled from the others as possible. This does not
deteriorate the properties of the system but does not help to enforce
them either. We talk more about the challenge of proving correctness in
Section \ref{sec:limitations}.

\hypertarget{extending-unification}{%
\subsection{Extending unification}\label{extending-unification}}

Now that unification is implemented let us create a simple plugin that
makes certain symbols declared by the user injective
\citep{agdausersInjectiveUnificationPragma2023}. The actual
implementation is relatively simple and is not dissimilar to the
Pi-injectivity solver we showed above.

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{userInjectivitySolver ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
                      \OtherTok{=\textgreater{}} \DataTypeTok{SolverType}\NormalTok{ cs}
\NormalTok{userInjectivitySolver constr }\OtherTok{=} \KeywordTok{do}
  \KeywordTok{let}\NormalTok{ (}\DataTypeTok{Just} \DataTypeTok{EqualityConstraint}
\NormalTok{              (}\DataTypeTok{I.App}\NormalTok{ (}\DataTypeTok{I.Var}\NormalTok{ f) a)}
\NormalTok{              (}\DataTypeTok{I.App}\NormalTok{ (}\DataTypeTok{I.Var}\NormalTok{ g) b) \_ m)) }\OtherTok{=}
\NormalTok{        match }\OperatorTok{@}\DataTypeTok{EqualityConstraint}\NormalTok{ constr}
  \KeywordTok{if}\NormalTok{ f }\OperatorTok{==}\NormalTok{ g}
  \KeywordTok{then} \KeywordTok{do}
\NormalTok{    ifM (queryInjectiveDeclarations f)}
\NormalTok{        (}\KeywordTok{do}
\NormalTok{           ms }\OtherTok{\textless{}{-}}\NormalTok{ constrainEquality a b }\DataTypeTok{I.Type}
\NormalTok{           solveMeta m ms}
           \FunctionTok{return} \DataTypeTok{True}\NormalTok{)}
\NormalTok{        (}\FunctionTok{return} \DataTypeTok{False}\NormalTok{)}
  \KeywordTok{else} \FunctionTok{return} \DataTypeTok{False}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

where \texttt{queryInjectiveDeclarations} simply scans the available
declarations for a marker that \texttt{f} has been declared injective.

The only big thing left is to make sure that this solver fires at the
right time. This can only conflict with the ``decomposition of
neutrals'' rule, so we indicate to the solver dispatcher that our plugin
should run before it:

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{userInjectivityPlugin ::}\NormalTok{ (}\DataTypeTok{EqualityConstraint} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
                      \OtherTok{=\textgreater{}} \DataTypeTok{Plugin}\NormalTok{ cs}
\NormalTok{userInjectivityPlugin }\OtherTok{=}
  \DataTypeTok{Plugin}\NormalTok{ \{ }\OperatorTok{...}
\NormalTok{         , solver }\OtherTok{=}\NormalTok{ userInjectivitySolver}
\NormalTok{         , symbol }\OtherTok{=} \DataTypeTok{PluginId} \OperatorTok{$} \StringTok{"userInjectivity"}
\NormalTok{         , pre }\OtherTok{=}\NormalTok{ [ unifyNeutralsDecomposition}
\NormalTok{                 , unificationEndMarkerSymbol]}
\NormalTok{         , suc }\OtherTok{=}\NormalTok{ [unificationStartMarkerSymbol]}
\NormalTok{         \}}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

This modification does not alter the core of the language.

\hypertarget{sec:casestudies}{%
\section{Case studies}\label{sec:casestudies}}

Once we implement basic elaboration and unification we can extend the
language. This is where we make use of the fact that the constraints
datatype is open.

We saw before in Section \ref{sec:implicit-arguments} that conventional
designs require separate handling of different kinds of implicit
variables. To simplify the design we would like to uniformly dispatch a
search for the solution, which would be handled by a fitting solver. We
can achieve this by communicating the kind of the meta through its type
in the second argument of \texttt{FillInTheMeta\ m\ ty}. The solvers
then match on the shape of the type of the metavariable and handle it in
a case-specific manner: instance-search for type classes, tactic
execution for a tactic argument, or waiting for regular unification to
solve the metavariable for regular implicit arguments.

In this section, we will describe the implementation details of regular
implicit arguments (Section \ref{sec:case-implicits}) and the
implementation of type classes added on top of the implicit arguments
(Section \ref{sec:case-typeclasses}). And finally (Section
\ref{sec:coercion-tactics}) we sketch the implementation of coercive
subtyping and tactic arguments.

\hypertarget{sec:case-implicits}{%
\subsection{Implicit arguments}\label{sec:case-implicits}}

As we mentioned, the rule for \texttt{Implicit} had to be added to the
syntax traversal part of the elaborator. In fact, we require not one but
two modifications that lie outside of the solvers-constraints part of
the system. The first one is, indeed, the addition of a separate case in
the syntax traversal, however contained. The second one lies in the
purely syntactical part of the compiler: we need the pre-processor to
insert the placeholder terms in the surface syntax.

In particular, we need to desugar declarations of functions in the
following way. For any declaration of a function with some implicit
arguments, we wrap the type of each implicit argument with an
\texttt{Implicit}:

\begin{verbatim}
def f : {A : Type} -> {a : A} -> B a -> C
\end{verbatim}

becomes

\begin{verbatim}
def f : (A : Implicit Type)
     -> (a : Implicit (deImp A))
     -> (b : B (deImp a)) -> C
\end{verbatim}

Then, for each function call we insert the corresponding number of
implicit arguments, for example transforming \texttt{f\ b1} to
\texttt{f\ \_\ \_\ b1}.

For basic usage of implicit arguments this suffices -- as soon as we
have the placeholders in the surface syntax we create the constraints in
the \texttt{Implicit} case of the syntax traversal. The only solver that
is needed in this case is a trivial one that checks that the
metavariable has been instantiated in the end. The actual solving in
this case is done by the unifier itself, and the constraint simply
serves as a way to guarantee that all implicits are eventually
instantiated.

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{fillInImplicitSymbol ::} \DataTypeTok{PluginId}

\OtherTok{fillInImplicitHandler ::}\NormalTok{ (}\DataTypeTok{FillInImplicit} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
                      \OtherTok{=\textgreater{}} \DataTypeTok{HandlerType}\NormalTok{ cs}
\NormalTok{fillInImplicitHandler constr }\OtherTok{=} \KeywordTok{do}
  \KeywordTok{let}\NormalTok{ ficm }\OtherTok{=}\NormalTok{ match }\OperatorTok{@}\DataTypeTok{FillInImplicit}\NormalTok{ constr}
  \KeywordTok{case}\NormalTok{ ficm }\KeywordTok{of}
    \DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{FillInImplicit}\NormalTok{ term ty) }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
      \KeywordTok{case}\NormalTok{ term }\KeywordTok{of}
        \DataTypeTok{MetaVar}\NormalTok{ (}\DataTypeTok{MetaVarClosure}\NormalTok{ mid \_) }\OtherTok{{-}\textgreater{}}
\NormalTok{          isMetaSolved mid}
\NormalTok{        \_ }\OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
    \DataTypeTok{Nothing} \OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}

\OtherTok{fillInImplicitPlugin ::}\NormalTok{ (}\DataTypeTok{FillInImplicit} \OperatorTok{:\textless{}:}\NormalTok{ cs)}
                     \OtherTok{=\textgreater{}} \DataTypeTok{Plugin}\NormalTok{ cs}
\NormalTok{fillInImplicitPlugin }\OtherTok{=} \DataTypeTok{Plugin}\NormalTok{ \{}
\NormalTok{    solver }\OtherTok{=}\NormalTok{ fillInImplicitSolver,}
\NormalTok{  , handler }\OtherTok{=}\NormalTok{ fillInImplicitHandler}
\NormalTok{  , symbol }\OtherTok{=}\NormalTok{ fillInImplicitSymbol}
\NormalTok{  , suc }\OtherTok{=}\NormalTok{ []}
\NormalTok{  , pre }\OtherTok{=}\NormalTok{ []}
\NormalTok{  \}}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

There is a design choice to be made in the implementation of
\texttt{Implicit\ A} and \texttt{deImp}. One option is to turn them into
a constructor and a projection of a record type, and the other is to
make them computationally equivalent to \texttt{id}. In the former case,
we need to manually unwrap them both in the pre-processor and during the
constraint-solving, but since the head symbol is distinct we can
guarantee that other solvers will not match on it, unless explicitly
instructed to. In the latter case, one has to be cautious of the order
in which the solvers are activated, particularly in the case of
different search procedures, should they be implemented. However, in the
example above it does not make a difference.

A limitation of this scheme is that we can only support functions with
``obvious'' implicit arguments -- i.e.~those that appear syntactically
in the declaration of the function. This is due to the fact that
insertion of metavariables happens before any type information is
available. For this reason Haskell-like impredicativity
\citep{serranoQuickLookImpredicativity2020a} in type inference cannot be
supported. If a more comprehensive support for implicit arguments is
desired, our system could be extended with support for first-class
implicits \citep{kovacsElaborationFirstclassImplicit2020}.

\hypertarget{sec:case-typeclasses}{%
\subsection{Type classes}\label{sec:case-typeclasses}}

Next, let us implement a plugin that adds support for type classes by
means of instance arguments \citep{DevrieseP11-1}. As in the case for
implicit arguments in general, we rely again on a pre-processor to
insert placeholder arguments of type \texttt{Instance\ a} for each
instance argument of type \texttt{a}.

Let us start by going through an example of the elaboration process for
a simple term. For type classes we need a few declarations, listed below
in Agda-like syntax:

\begin{verbatim}
plus  :  {A : Type} -> {{PlusOperation A}}
     -> (a : A) -> (b : A) -> A

instance BoolPlus : PlusOperation Bool where
  plus = orb
\end{verbatim}

And the exemplary term itself is:

\begin{verbatim}
m = plus True False
\end{verbatim}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  First, the pre-processor eliminates the implicits and type class
  arguments. We end up with the following declarations:

\begin{verbatim}
plus : (impA : Implicit Type)
    -> Instance PlusOperation (deImp impA)
    -> (a : deImp impA) -> (b :  deImp impA)
    -> deImp impA

instanceBoolPlus : InstanceT PlusOperation Bool
instanceBoolPlus =
  InstanceC (TypeClassC (Plus orb))

m = plus _ _ True False
\end{verbatim}

  Turning the declaration \texttt{instance\ BoolPlus} into a usage of
  the constructor \texttt{InstanceC} is precisely the part we need the
  pre-processor to do.
\item
  Next, we go into the elaboration of \texttt{m}. The elaborator applies
  \texttt{inferType\ (App\ t1\ t2)} rule four times and
  \texttt{checkType\ (Implicit)\ ty} twice on the two placeholders. The
  output of the elaborator is

\begin{verbatim}
m = plus ?_1 ?_2 True False
\end{verbatim}

  The state of the elaborator now contains four constraints:

\begin{verbatim}
C1: FillInTheTerm ?_1 (Implicit Type)
C2: FillInTheTerm ?_2 (InstanceT PlusOperation
                                 (deImp ?_1))
C3: EqualityConstraint (deImp ?_1) Bool Type
C4: EqualityConstraint (deImp ?_1) Bool Type
\end{verbatim}

  The first two correspond to implicit arguments, while the latter two
  are unification problems rendered into constraints.
\item
  Now we step into the constraint-solving world. First, the unifier
  solves the latter two constraints, instantiating \texttt{?\_1} to
  \texttt{Implicit\ Bool}. \texttt{C1} is then discarded as solved since
  \texttt{?\_1} is already instantiated to \texttt{Implicit\ Bool}.
  Next, the type class resolution launches a search for the instance of
  type \texttt{Instance\ PlusOperation\ Bool}.
\item
  This is where the type class plugin can take over. It transforms
  \texttt{C2:\ FillInTheTerm\ ?\_2\ (InstanceT\ PlusOperation\ Bool)} to
  \texttt{C5:\ InstanceSearch\ PlusOperation\ Bool\ ?\_2}. \texttt{C5}
  then gets matched by the search plugin for concrete instances, simply
  weeding through available declarations, looking for something of the
  shape \texttt{InstanceT\ PlusOperation\ Bool}. Such a declaration
  exists indeed and we can instantiate \texttt{?\_2} to
  \texttt{instanceBoolPlus}.
\end{enumerate}

Now let us take a look at the plugin for the constraint system. It is
contained in a single file
\href{https://github.com/liesnikov/extensible-elaborator/blob/elaborator-experiments/exel/src/Plugins/Typeclasses.hs}{\texttt{./exel/src/Plugins/Typeclasses.hs}}.
In it, we define a new constraint type \texttt{InstanceSearch}, a solver
that transforms constraints of the shape
\texttt{FillInTheType\ ?\ (InstanceT\ \_)} to the new constraint, and
finally, a solver for instance search problems. The reason to transform
the original constraint to the new type is to ensure that no other
solver will make an attempt at this problem, therefore modulating
interactions with other plugins.

Finally, the implementation of search for concrete instances is quite
simple:

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{instanceConcreteSolver constr }\OtherTok{=} \KeywordTok{do}
  \KeywordTok{let}\NormalTok{ (}\DataTypeTok{Just}\NormalTok{ (}\DataTypeTok{InstanceSearch}\NormalTok{ tcn ty m)) }\OtherTok{=}
\NormalTok{        match }\OperatorTok{@}\DataTypeTok{InstanceSearch}\NormalTok{ constr}
\NormalTok{  alldecls }\OtherTok{\textless{}{-}}\NormalTok{ collectInstancesOf tcn }\OperatorTok{\textless{}$\textgreater{}}\NormalTok{ SA.getDecls}
\NormalTok{  sty }\OtherTok{\textless{}{-}}\NormalTok{ SA.substAllMetas ty}
  \KeywordTok{case}\NormalTok{ Map.lookup sty alldecls }\KeywordTok{of}
    \DataTypeTok{Just}\NormalTok{ i }\OtherTok{{-}\textgreater{}} \KeywordTok{do}
\NormalTok{      SA.solveMeta m (}\DataTypeTok{I.Var}\NormalTok{ i)}
      \FunctionTok{return} \DataTypeTok{True}
    \DataTypeTok{Nothing} \OtherTok{{-}\textgreater{}} \FunctionTok{return} \DataTypeTok{False}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

We conjecture that implementation of canonical structures
\citep{mahboubiCanonicalStructuresWorking2013} would be relatively
simple in such a system due to the openness of both unification
procedure and instance search.

\hypertarget{sec:coercion-tactics}{%
\subsection{Tactic arguments and coercions}\label{sec:coercion-tactics}}

In a similar fashion to the transformation of the
\texttt{FillInImplicit} constraints to \texttt{InstanceSearch}, we can
implement plugins for tactic arguments and coercive subtyping.

The former would be quite similar to what we saw in the previous section
\ref{sec:case-typeclasses}, except we would have to resolve
\texttt{FillInImplicit} to a \texttt{RunTactic} constraint (or run the
tactic directly). The question of how to actually run the tactics is
independent of the constraint machinery, noting that we allow for
running arbitrary type-checking actions during solving of constraints.

Coercive subtyping is of a slightly different nature. First, it would be
the most reliant on a pre-processor out of all of the examples described
above, since coercions could potentially be inserted around each
function argument, and potentially also around heads of applications and
types of abstractions \citep{tassiBiDirectionalRefinementAlgorithm2012}.
However, blindly inserting placeholders for coercions in every position
leads to constraints that cannot be resolved locally. Consider the
following example:

\begin{verbatim}
f : (A : Type) -> A -> A
arg : ArgTy

t : ExpTy
t = f _ arg
\end{verbatim}

This declaration would desugar to

\begin{verbatim}
t = coerce _ (f _ (coerce _ arg))
\end{verbatim}

which gives rise to two constraints, where \texttt{Coercion\ A\ B}
stands for the evidence of \texttt{A} being a subtype of \texttt{B}:

\begin{verbatim}
C1 : Coercion ArgTy ?1
C3 : Coercion ?1 ExpTy
\end{verbatim}

Clearly, we can not solve the first constraint in isolation and need to
consider all coercion constraints related to a particular type at the
same time. Hence this feature would be anti-modular in the sense that in
order to side-step the conflict between inference and coercions the
solver would need to look at the whole graph of subtyping constraints at
the same time. While we can accommodate such solvers, due to solvers
having write access to the state, it becomes much harder to modulate
interactions between different plugins. It also comes with a potential
performance penalty, which we describe further in Section
\ref{sec:limitations}.

\hypertarget{sec:limitations}{%
\section{Limitations}\label{sec:limitations}}

While our design offers a lot of flexibility, it does not solve every
problem. In this section, we describe a few examples of extensions that
do not quite fit in this framework as well as more general limitations.

\hypertarget{handling-of-meta-variables-outside-of-definition-sites}{%
\subsection{Handling of meta-variables outside of definition
sites}\label{handling-of-meta-variables-outside-of-definition-sites}}

After we elaborate a definition, there can still be unsolved
metavariables in it. This presents us with a design choice. The first
option is to generalize the definition over the metavariables and
instantiate them per-use site, essentially considering the metavariables
as additional (implicit) arguments to the definition. The second option
is to leave them up to be solved later, which might make elaboration
less predictable since now the use sites can influence whether a
particular definition type-checks. The third option is to freeze the
metavariables by considering them to be postulates, and report them as
an error at the end of type checking.

In particular, the second option allows us to incorporate more involved
inference algorithms into the system. For example, if we were to
implement an erasure inference algorithm as described by
Tejiščák \cite{tejiscakDependentlyTypedCalculus2020}, we would have to create
metavariable annotations (described as ``evars'' in the paper) that can
be instantiated beyond the definition site. The downside is that the
meaning of a definition can then change depending on where they are
used, making the outcome of elaboration depend on the file as a whole
rather than the definition itself.

\hypertarget{the-language-is-only-as-extensible-as-the-syntax-traversal-is}{%
\subsection{The language is only as extensible, as the syntax traversal
is}\label{the-language-is-only-as-extensible-as-the-syntax-traversal-is}}

Extensibility via constraints allows for a flexible user-specified
control flow as soon as we step into the constraints world. However, not
all features can be supported purely by adding constraints and solvers.
For example, consider the following simplified lambda-function
type-checking function\footnote{\href{https://github.com/agda/agda/blob/v2.6.4/src/full/Agda/TypeChecking/Rules/Term.hs\#L430-L518}{./src/full/Agda/TypeChecking/Rules/Term.hs\#L430-L518}}
from Agda:

\begin{imageonly}
\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{checkLambda\textquotesingle{} cmp b xps typ body target }\OtherTok{=} \KeywordTok{do}
  \DataTypeTok{TelV}\NormalTok{ tel btyp }\OtherTok{\textless{}{-}}\NormalTok{ telViewUpTo numbinds target}
  \KeywordTok{if}\NormalTok{ size tel }\OperatorTok{\textless{}}\NormalTok{ numbinds }\OperatorTok{||}\NormalTok{ numbinds }\OperatorTok{/=} \DecValTok{1}
    \KeywordTok{then}\NormalTok{ dontUseTargetType}
    \KeywordTok{else}\NormalTok{ useTargetType tel btyp}
\end{Highlighting}
\end{Shaded}
\end{imageonly}

Here Agda steps away from the bidirectional discipline and infers a
(lambda) function if the target type is not fully known. In order to add
such a rule to our implementation, we would have to update syntax
traversal to be more flexible in where it allows lambda expressions to
appear. The solution, in this case, is effectively to replicate what
Agda is doing by implementing each type-checking rule in inference mode,
essentially factoring out \texttt{dontUseTargetType} in Agda's code
snippet above.

Likewise, implementing the commutativity and associativity unifier
plugins from Holten \cite{holtenDependentTypeCheckingModulo2023} requires
modifications in the core language, since we two terms that are equal
during elaboration have to equal during core type-checking too.

\hypertarget{lack-of-backtracking}{%
\subsection{Lack of backtracking}\label{lack-of-backtracking}}

We do not implement any backtracking in the solver dispatcher as it is
now. This means that every step taken is committing to a specific choice
to how to solve the constraint, which can be a limitation in cases where
one would like to have backtracking -- for example, Agda's instance
arguments search (with \texttt{-\/-overlapping-instances} flag)
\citep[chap.~3.18]{theagdateamAgdaUserManual2023a}, as well as type
classes in Lean \citep{selsamTabledTypeclassResolution2020}.

Backtracking in principle could be achieved by tracking changes to the
state of the elaborator and the production graph for constraints, but
such a system would be rather awkward. Alternatively, controlled
backtracking can be implemented within one solver, removing the
extension points, but allowing arbitrary control flow within the
algorithm.

\hypertarget{reliance-on-a-pre-processor}{%
\subsection{Reliance on a
pre-processor}\label{reliance-on-a-pre-processor}}

This work crucially relies on a pre-processor of some kind, be it macro
expansion or some other way to extend the parser with custom desugaring
rules. In particular, in order to implement n-ary implicit arguments
correctly and easily we need the pre-processor to expand them to the
right arity, similar to Matita
\citep[chap.~5]{tassiBiDirectionalRefinementAlgorithm2012} and others
\citep{serranoQuickLookImpredicativity2020a, kovacsElaborationFirstclassImplicit2020}.

\hypertarget{eager-reduction-and-performance}{%
\subsection{Eager reduction and
performance}\label{eager-reduction-and-performance}}

As the pre-processing step inserts a fair number of wrappers and
un-wrappers for all implicits and even more for coercions, we can expect
a performance penalty due to larger terms and spurious computation
steps. In particular, we expect this to be a major concern for coercive
subtyping.

As one possible mitigation, we suggest a discipline where constraint
solvers latch on to non-reduced types and terms in constraints. With
this discipline, we can borrow a trick from the implementation of Coq,
where the wrappers and unwrappers are identity functions and
\texttt{coerce\ f\ t} computes to \texttt{f\ t}. This also means that
constraints can/have to match on unreduced types in the
e.g.~\texttt{FillInTheTherm}.

In fact, we already use this trick to an extent for a different reason
-- since the calculus allows only fully applied type constructors, we
have to wrap each type class constructor in a lambda-abstraction for it
to appear as an argument to \texttt{TypeClassT\ typeClassName\ argType}.
Or, concretely, we have to define and use
\texttt{PlusOperation\textquotesingle{}\ =\ \textbackslash{}A\ .\ PlusOperation\ A}
in place of \texttt{PlusOperation} in the elaboration example in Section
\ref{sec:case-typeclasses}.

\hypertarget{proving-correctness}{%
\subsection{Proving correctness}\label{proving-correctness}}

As soon as we allow the users to implement their own solvers, there is
little we can say about the correctness of the system as a whole without
imposing proof obligations on the plugin writers. This is simply a
consequence of the fact that we do not invent a new unification
algorithm, but rather provide means of easier implementation for it. We
conjecture that our system satisfies the solution and type preservation
properties -- Theorems 2 and 3 from
Abel and Pientka \cite{abelHigherOrderDynamicPattern2011}, respectively -- assuming that
each solver on its own satisfies these properties. For termination the
situation is similar -- we conjecture that if every solver in the system
``reduces the weight'' of the unification problem, in terms of Theorem 1
by \citet{abelHigherOrderDynamicPattern2011}, we can guarantee
termination of the solving process as a whole.

\hypertarget{sec:related_work}{%
\section{Related work}\label{sec:related_work}}

We are certainly not the first ones to try to tackle the extensibility
of a language implementation. Dependently typed language implementations
usually consist of at least four parts: parser, elaborator, core
type-checker, and code generation backend. The code generation part is
currently irrelevant to our interests, since for a language to be
specified usually means for specification of the core, anything that
happens after the core does not extend the language, but rather tries to
preserve its semantics in some form. Therefore we're left with three
parts: parser, elaborator, and core type-checker.

We see parser or syntax extensibility as a necessary part of an
extensible language. This problem has been studied extensively in the
past and has a multitude of existing solutions. Macros are one of them
and are used heavily in various forms in many established languages
\citep{thecoqdevelopmentteamCoqProofAssistant2022, theagdateamAgdaUserManual2023a, ullrichNotationsHygienicMacro2020}
and can be powerful enough to build a whole language around
\citep{changDependentTypeSystems2019}.

Core extensibility, on the other hand, appears to be a problem with too
many degrees of freedom. Andromeda
\citep{bauerDesignImplementationAndromeda2018, bauerEqualityCheckingGeneral2020}
made an attempt at extensible definitional equality but is quite far
from a usable dependently typed language. Agda's philosophy allows
developers to experiment with the core but also results in a larger
amount of unexpected behaviours. In general, modifications of the core
rules will result in fundamental changes in the type theory, which can
break plenty of important properties like soundness or subject
reduction.

This leaves us with the question of the extensibility of an elaborator.
We will make a division here between syntax traversals, constraint
solving and all other features. The syntax traversal part of the
elaborator is relatively stable and commonly implemented following a
(roughly) bidirectional
discipline\citep{norellPracticalProgrammingLanguage2007, tassiBiDirectionalRefinementAlgorithm2012, ferreiraBidirectionalElaborationDependently2014},
so there seems little reason to make it extensible.

GHC has a plugin system that allows users to dynamically add custom
constraint solvers, but the type of constraints itself is not
extensible\footnote{\href{https://gitlab.haskell.org/ghc/ghc/-/wikis/plugins/type-checker/notes}{gitlab.haskell.org/ghc/ghc/-/wikis/plugins/type-checker/notes}}
\citep{peytonjonesTypeInferenceConstraint2019, vytiniotisOutsideInModularType2011, peytonjonesPracticalTypeInference2007}.

Coq \citep{thecoqdevelopmentteamCoqProofAssistant2022}, being one of the
most popular proof assistants, invested a lot effort into user-facing
features: work on tactics like a new tactic engine
\citep{spiwackVerifiedComputingHomological2011} and tactic languages
(Ltac2 \citep{pedrotLtac2TacticalWarfare2019}, SSReflect
\citep{gonthierSmallScaleReflection2008}, etc.), the introduction of a
virtual machine for performance
\citep{gregoireCompiledImplementationStrong2002} and others. However,
the implementation is quite hard to extend. One either has to modify the
source code, which is mostly limited to the core development team, as
seen from the
\href{https://github.com/coq/coq/graphs/contributors}{contributors
graph}, or one has to use Coq plugin system, which is rather
challenging, and in the end, the complexity of it gave rise to 
TemplateCoq/MetaCoq  \aptLtoX[graphic=no,type=html]{\cite{malechaExtensibleProofEngineering2014, sozeauMetaCoqProject2020}}{\citetext{\citealp[
]{malechaExtensibleProofEngineering2014}; \citealp{sozeauMetaCoqProject2020}}}.
While MetaCoq did open the possibility for some plugins
\citep{nielsenFormalisingDecentralisedExchanges2023, liesnikovGeneratingInductionPrinciples2020, forsterCertifyingExtractionTime2019}
to be written in a simpler way, their capabilities are still limited.

Agda has historically experimented a lot with different extensions to
both the type system and the elaborator, even though the design does not
accommodate these changes naturally. Instead, each of these extensions
is spread throughout many different parts of the code base\footnote{some
  recent examples:
  \href{https://github.com/agda/agda/pull/6385}{github.com/agda/agda/pull/6385},
  \href{https://github.com/agda/agda/pull/6354}{github.com/agda/agda/pull/6354}.}.

Lean introduced elaborator extensions
\citep{leonardodemouraLeanMetaprogramming2021, ullrichNotationsHygienicMacro2020}.
They allow the user to overload the commands, but if one defines a
particular elaborator it becomes hard to interleave with others. In a
way, this is an imperative view on extensibility.

Idris
\citep{bradyIdrisGeneralpurposeDependently2013, christiansenElaboratorReflectionExtending2016}
appeared as a programming language first and proof-assistant second and
does not provide either a plugin or hook system at all, except for
reflection. Idris also focuses on tactics as the main mechanism for
elaboration.

Turnstile+ by \citet{changDependentTypeSystems2019} uses macros to
elaborate surface syntax to a smaller core. Macros allow them to
modularly implement individual features, however combining different
features requires the user to re-define all macros from scratch. This is
the same problem as the one we mentioned for Lean.

TypOS \citep{allaisTypOSOperatingSystem2022a, guillaumeallaisTypOS2022}
is perhaps the closest to our work, but there are two important
differences. First, it is a domain-specific language for building
type-checkers, while our design is language-agnostic, as long as the
host language can model extensible datatypes in some capacity. Second,
their approach settles features of the language as they are decided by
the main developer and does not concern future changes and evolution.
Finally, we try to stay close to the designs of existing dependently
typed languages and offer flexibility in terms of choices, while TypOS
requires the developer to start from scratch and restricts certain
capabilities like overlapping rules for unification.

\hypertarget{future-work}{%
\section{Future work}\label{future-work}}

We see three main prospects for future work:

\begin{itemize}
\tightlist
\item
  \textbf{Exploration of different kinds of metavariables.} Currently,
  we implement metavariables only for terms, while for some applications
  such as erasure \citep{tejiscakDependentlyTypedCalculus2020} or
  irrelevance inference, it would be beneficial to have metavariables
  representing erasure and relevance annotations. Additionally, we can
  introduce metavariables for names and implement data constructor
  disambiguation more simply, which would remove the current need to
  block on the expected type of overloaded constructors.
\item
  \textbf{Rendering more elements of the elaborator as constraints.}
  Currently, components such as the occurs checking and reduction are
  simple function calls. Including them in the constraint machinery
  would make the implementation more uniform and allow users to extend
  them.
\item
  \textbf{Error messages} are not the focus of present work, but it
  would be interesting to see if we can incorporate ideas by
  \citet{heerenScriptingTypeInference2003} into our system.
\item
  \textbf{Potential optimisations}. Currently, our system has a lot of
  room for potential optimisations. The first step would be to allow
  handlers to pass some information to their respective solvers, which
  is conceptually an easy change but technically requires introduction
  of an existential type in the plugin. Additionally, some expandable
  per-plugin store for the solvers would be useful, for example, to
  avoid recompututation of type class instances on every invocation.
  Somewhat more ambitiously, one can imagine a caching system for
  constraints, avoiding the need for solving the same constraint more
  than once. In particular, caching of reduction seems like it would be
  beneficial since we currently do a lot of redundant computations.
  However, the memory usage of such a caching system might be
  prohibitive. Finally, we would also like to explore possibilities for
  concurrent solving, similar to the plans of
  \citet{allaisTypOSOperatingSystem2022a} to use LVars for representing
  metavariables \citep{kuperLatticebasedDataStructures2015}.
\end{itemize}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\end{CSLReferences}

### 2024-03-22 camera-ready-submission/source/main.tex
add all the files

--- 2024-03-15 paper/main.tex
@@
+\pdfminorversion=7
 % Options for packages loaded elsewhere
@@
 \documentclass[sigconf, screen]{acmart}
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
+
 \usepackage{xcolor}
+%\usepackage{cleveref}
 \usepackage{color}
 \usepackage{fancyvrb}
+\usepackage{acmart-taps}
 \newcommand{\VerbBar}{|}
@@
 
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\usepackage[]{natbib}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+%\ifLuaTeX
+%  \usepackage{selnolig}  % disable illegal ligatures
+%\fi
+%\usepackage[]{natbib}
+%\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+%\urlstyle{same} % disable monospaced font for URLs
 
@@
 \displaywidowpenalty=10000
+
+\copyrightyear{2023}
+\acmYear{2023}
+\setcopyright{rightsretained}
+\acmConference[IFL 2023]{The 35th Symposium on Implementation and Application of Functional Languages}{August 29--31, 2023}{Braga, Portugal}
+\acmBooktitle{The 35th Symposium on Implementation and Application of Functional Languages (IFL 2023), August 29--31, 2023, Braga, Portugal}
+\acmDOI{10.1145/3652561.3652565}
+\acmISBN{979-8-4007-1631-7/23/08}
+\keywords{dependent types, elaborators, elaboration, open datatype, extensible datatype, type-checking}
+
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
+       <concept_desc>Software and its engineering~Functional languages</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011074.10011075.10011077</concept_id>
+       <concept_desc>Software and its engineering~Software design engineering</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10003752.10003790.10011740</concept_id>
+       <concept_desc>Theory of computation~Type theory</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10002944.10011123.10011673</concept_id>
+       <concept_desc>General and reference~Design</concept_desc>
+       <concept_significance>100</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Functional languages}
+\ccsdesc[300]{Software and its engineering~Software design engineering}
+\ccsdesc[300]{Theory of computation~Type theory}
+\ccsdesc[100]{General and reference~Design}
+
+\begin{document}
 
@@
 
-\begin{document}
 \maketitle
@@
 \begin{acks}
-  Jesper Cockx holds an \grantsponsor{nwoviveni202216}{NWO Veni}{https://www.nwo.nl/en/projects/viveni202216} grant on `A trustworthy and extensible core language for Agda' \grantnum[https://www.nwo.nl/en/projects/viveni202216]{nwoviveni202216}{(VI.Veni.202.216)}.
+  Jesper Cockx holds an \grantsponsor{nwoviveni202216}{NWO Veni}{https://www.nwo.nl/en/projects/viveni202216} grant on `A trustworthy and extensible core language for Agda' (\href{https://www.nwo.nl/en/projects/viveni202216}{VI.Veni.202.216}).
 \end{acks}
@@
 \bibliographystyle{ACM-Reference-Format}
+
 \bibliography{bib.bib}

### 2024-03-22 paper/main.tex
add acm-provided info
sync paper with camera-ready

@@
 
+\copyrightyear{2023}
+\acmYear{2023}
+\setcopyright{rightsretained}
+\acmConference[IFL 2023]{The 35th Symposium on Implementation and Application of Functional Languages}{August 29--31, 2023}{Braga, Portugal}
+\acmBooktitle{The 35th Symposium on Implementation and Application of Functional Languages (IFL 2023), August 29--31, 2023, Braga, Portugal}
+\acmDOI{10.1145/3652561.3652565}
+\acmISBN{979-8-4007-1631-7/23/08}
+
 \title{ExEl: Building an Elaborator Using Extensible Constraints}
